### 2015-02-16 main_text.tex
First commit, starting introduction

\documentclass[11pt, letterpaper]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

% Math, graphics, and bibliography
\usepackage{amsmath,amssymb,amsfonts,mathrsfs,mathtools}
\usepackage{graphicx}
\usepackage[round,numbers,sort&compress]{natbib}
\renewcommand{\bibnumfmt}[1]{#1.}

% My preferred fonts
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[sc]{mathpazo}

% For fancier fractions and figure captions
\usepackage[labelfont={bf}, margin=1cm]{caption}
\usepackage{units}
\usepackage{booktabs}

% For supplemental figures
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

% Puts the affiliations on the cover page
\usepackage{authblk}

\begin{document}
\title{Quantifying stochastic noise in cultured\\ circadian reporter cells}
\author[1]{Peter C. St. John}
\author[1,*]{Francis J. Doyle III}
\affil[1]{Department of Chemical Engineering, University of California Santa
Barbara, Santa Barbara, California 93106-5080}
\affil[*]{Email: \texttt{doyle@engineering.ucsb.edu}}
\date{\today}
\maketitle

\begin{center}
Running head:\\ {Quantifying circadian stochastic noise} \\[1ex]
Keywords:\\ Systems Biology | Circadian Rhythms \\ Gene
Regulatory Network | Stochastic | Synchronization
\end{center}

\pagebreak
\begin{abstract}

\end{abstract}

\section*{Introduction}

Circadian rhythms in mammals are daily changes in physiology and gene expression that persist even in the absence of external environmental cues.

\section*{Materials and Methods}

\section*{Results and Discussion}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/noise_ts_and_boot.pdf}
  \end{center}
  \caption{{\bfseries Figure Title.}
({\bfseries A}) Part A.
({\bfseries B}) Part B.
({\bfseries C}) Part C.}
\label{fig:fibroblast_noise}
\end{figure}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/mainfig_dose.pdf}
  \end{center}
  \caption{{\bfseries Figure Title.}
({\bfseries A}) Part A.
({\bfseries B}) Part B.}
\label{fig:dose_dependence}
\end{figure}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/main_fig_simulations.pdf}
  \end{center}
  \caption{{\bfseries Figure Title.}
({\bfseries A}) Part A.
({\bfseries B}) Part B.}
\label{fig:simulation}
\end{figure}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/r2_vs_decay.pdf}
  \end{center}
  \caption{{\bfseries Figure Title.}
({\bfseries A}) Part A.
({\bfseries B}) Part B.}
\label{fig:fit_quality}
\end{figure}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/fitted_parameters.pdf}
  \end{center}
  \caption{{\bfseries Figure Title.}
({\bfseries A}) Part A.
({\bfseries B}) Part B.}
\label{fig:fit_distributions}
\end{figure}

\begin{table}
  \begin{center}
    \begin{tabular}{lrrrrrr}\toprule
      {} & \multicolumn{2}{c}{$T$} & \multicolumn{2}{c}{$\ln A$} & \multicolumn{2}{c}{$d$} \\
      {}         & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}               & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           \\\midrule
      $\mu$      & $-0.234$  & $0.187$  &  $0.443$  & $-0.343$  &  $0.043$  & $0.090$    \\
      $\sigma$   & $ 0.774$  & $1.820$  &  $0.778$  & $ 1.753$  &  $0.878$  & $1.688$    \\
      Skew & $ 0.153$  & $0.367$  & -$1.823$  & $-0.580$  & -$0.107$  & $0.371$    \\
      Kurt & $ 3.772$  & $0.591$  &  $8.329$  & $ 0.476$  &  $2.423$  & $0.373$    \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{{\bfseries Fitted Parameters}}
  \label{tab:fit_distributions}
\end{table}

\begin{table}
  \begin{center}
    \begin{tabular}{rrrrr}
      \toprule
      {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\\
      $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
      $T$      & $-0.142$ & $-0.022$  & $1.000 $ & $-0.113$\\
      $\theta$ & $-0.269$ & $-0.112$  & $-0.113$ & $1.000 $\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{{\bfseries Correlation matrix}}
  \label{tab:corr}
\end{table}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/outliers.pdf}
  \end{center}
  \caption{{\bfseries Figure Title.}
({\bfseries A}) Part A.
({\bfseries B}) Part B.}
\label{fig:outlier_dist}
\end{figure}

\section*{Conclusion}

\section*{Acknowledgments}
This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.

% \bibliographystyle{biophysj.bst}
% \bibliography{condensed_library.bib}

\beginsupplement

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/volume_calibration.pdf}
  \end{center}
  \caption{{\bfseries Figure Title.}
({\bfseries A}) Part A.
({\bfseries B}) Part B.}
\label{fig:vol_calibration}
\end{figure}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/zhang_plates.pdf}
  \end{center}
  \caption{{\bfseries figure title.}
({\bfseries a}) part a.
({\bfseries b}) part b.}
\label{fig:plate_variation}
\end{figure}

\begin{figure}[tbp]
  \begin{center}
    \includegraphics[]{figures/pdfs/zhang_wells.pdf}
  \end{center}
  \caption{{\bfseries figure title.}
({\bfseries a}) part a.
({\bfseries b}) part b.}
\label{fig:well_variation}
\end{figure}

\begin{table}
\begin{center}
\begin{tabular}{lclc}
\toprule
\textbf{Dep. Variable:}    &      decay       & \textbf{  R-squared:         } &$     0.169  $\\
\textbf{Model:}            &       OLS        & \textbf{  Adj. R-squared:    } &$     0.169  $\\
\textbf{Method:}           &  Least Squares   & \textbf{  F-statistic:       } &$     4782.  $\\
\textbf{Date:}             & Wed, 11 Feb 2015 & \textbf{  Prob (F-statistic):} &$     0.00   $\\
\textbf{Time:}             &     16:26:22     & \textbf{  Log-Likelihood:    } &$-1.7248e+05 $\\
\textbf{No. Observations:} &       94053      & \textbf{  AIC:               } &$ 3.450e+05  $\\
\textbf{Df Residuals:}     &       94048      & \textbf{  BIC:               } &$ 3.450e+05  $\\
\bottomrule
\end{tabular}
\begin{tabular}{lccccc}
                   & \textbf{coef} & \textbf{std err} & \textbf{t} & \textbf{P$>$$|$t$|$} & \textbf{[95.0\% Conf. Int.]}  \\
\midrule
\textbf{const}     &     $-0.0370$ &       $0.014$    &   $-2.572$ &        $0.010$       &       $-0.065,  -0.009$      \\
\textbf{amplitude} &     $ 0.2375$ &       $0.003$    &   $86.282$ &        $0.000$       &       $ 0.232,   0.243$      \\
\textbf{period}    &     $-0.1521$ &       $0.003$    &   $56.798$ &        $0.000$       &       $-0.157,  -0.147$      \\
\textbf{phase}     &     $-0.2354$ &       $0.003$    &   $85.598$ &        $0.000$       &       $-0.241,  -0.230$      \\
\textbf{type}      &     $ 0.3197$ &       $0.015$    &   $20.664$ &        $0.000$       &       $ 0.289,   0.350$      \\
\bottomrule
\end{tabular}
\begin{tabular}{lclc}
\textbf{Omnibus:}       &$9769.391$& \textbf{  Durbin-Watson:     } &$    1.876$ \\
\textbf{Prob(Omnibus):} &$  0.000 $& \textbf{  Jarque-Bera (JB):  } &$18459.719$ \\
\textbf{Skew:}          &$  0.697 $& \textbf{  Prob(JB):          } &$     0.00$ \\
\textbf{Kurtosis:}      &$  4.664 $& \textbf{  Cond. No.          } &$     8.34$ \\
\bottomrule
\end{tabular}
\end{center} 
\caption{OLS Regression Results}
\label{tab:ols_reg}
\end{table}

\end{document}

### 2015-02-18 main_text.tex
finished methods, starting results

@@
 \usepackage[sc]{mathpazo}
+
+% For hyperref links
+\usepackage[linktoc=all]{hyperref}
+\hypersetup{
+    colorlinks,
+    citecolor=black,
+    filecolor=black,
+    linkcolor=black,
+    urlcolor=black
+}
 
@@
 
-Circadian rhythms in mammals are daily changes in physiology and gene expression that persist even in the absence of external environmental cues.
+Circadian rhythms in mammals are daily changes in gene expression and physiology that persist even in the absence of external environmental cues \cite{Herzog2007}.
+Such rhythms are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations \cite{Ueda2005}.
+The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
+Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998}, and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
+Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
+This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014}, and has lead to the search for small-molecule therapeutics which boost clock amplitude \cite{Chen2013}.
+
+Transcription at the single-cell level is strongly affected by the low molecular counts of the mRNA and protein species involved. 
+As a result, bioluminescence traces of individual cells are stochastic, with significant period-to-period variability \cite{Welsh2004}. 
+The collective behavior of thousands of cells results in more reliable oscillations at the tissue-level, especially in the suprachiasmatic nucleus (SCN) where cell-to-cell coupling keeps individual oscillators at a consistent phase \cite{Herzog2004}.
+While communication between cells is common in many tissues, it is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues or cultured reporter cells, is very weak, if present \cite{Guenthner2014, Noguchi2013}.
+In populations which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level \cite{Welsh2004}. 
+Despite the averaging which occurs at the population-level, stochastic noise at the single-cell level plays an important role in determining the function of the circadian oscillator.
+A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues \cite{An2013}.
+Additional studies have even argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators - when simulated stochastically - better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
+
+Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful plating and recording of fibroblast cells and subsequent image processing \cite{Leise2012}.
+As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.
+In this study, we demonstrate single-cell stochasticity can be reliably inferred from the damping rate of population-level bioluminescence recordings.
+Additionally, we show that a small-molecule modulator is able to change circadian noise in a dose-dependent fashion. 
+Finally, we calculate the genome-wide effects of siRNA knockdown on single-cell noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
+Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on single-cell noise to be quantified in a high-throughput manner.
 
 \section*{Materials and Methods}
+
+\subsection*{Fitting a damped sinusoid to experimental data}
+A damped sinusoid, specified by:
+\[
+  \hat{y}(t) = A e^{-d t} \sin\left(\frac{2\pi t}{T} + \theta\right)
+\]
+is fit to experimental data $x_i(t_i), \  i \in \{0, \dots, N-1\}$.
+The $x_i$ data points are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{s}\right)^4$, in which $s$ is the sampling rate (in hours) \cite{Ravn2002}.
+The detrended data is then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1)$.
+We denote the detrended and filtered experimental data by $y_i(t_i)$.
+For numerical efficiency, the period, $T$, and damping rate, $d$, parameters are fit first using a matrix pencil method \cite{Hua1990}, reviewed in \cite{Zielinski2011}.
+Amplitude, $A$, and phase, $\theta$, parameters are subsequently fit using a linear least-squares regression.
+Overall $R^2$ values for the regression were calculated from the residual error between the detrended data and fitted sinusoid:
+\[
+  R^2 = 1 - \frac{\sum_{i = 0}^{N-1} (y_i(t_i) - \hat{y}(t_i))^2}{\sum_{i = 0}^{N-1} (y_i(t_i) - \bar{y}(t_i))^2}
+\]
+
+\subsection*{Processing single-cell bioluminescence data}
+
+Single-cell bioluminescence data for 79 cells was obtained from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
+As was done in the original study, a discrete wavelet transform (using PyWavelets, \url{http://www.pybytes.com/pywavelets}) was performed to detrend and remove noise.
+
+\subsubsection*{Sorting cells by noise level}
+As in the original study, various parameters describing the average noise level of each cell were collected.
+Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components.
+From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
+An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
+From these three noise metric, a combined noise variable was constructed via a principle component analysis (using scikit-learn, \url{http://scikit-learn.org/}).
+Cells were ranked according to this combined noise metric, and a high-noise group and low-noise group were constructed by taking the 39 highest-noise and lowest-noise cells, respectively.
+The raw bioluminescence profiles were not initially synchronized, so in order to simulate the gradual desynchronization of a group of oscillators the traces were offset to have the same starting phase.
+This was accomplished by starting each trace at the first phase zero-crossing, which were found using a Hilbert transform.
+
+\subsubsection*{Bootstrap estimations of the damping rate difference}
+Averaged traces for low and high-noise group displayed a damped sinusoidal rhythm.
+The first 4 days of rhythms ($s = 0.5,\ N = 192$) were fit using a damped sinusoid.
+To ensure the difference in damping rate between groups was statistically significant, a bootstrap analysis was performed.
+In each of $10,000$ bootstrap trials, the cells were randomly assigned evenly to either the low-noise or the high-noise group.
+The absolute difference in damping rate between the two populations was recorded to yield a two-tailed test.
+The observed test statistic, $|d_h - d_l| = 6.65\times10^{-3}$, was found to be significant at the $\alpha = 0.05$ confidence threshold ($p = 0.0264$).
+
+\subsection*{Quantifying dose-dependent effects of small molecule modulators}
+Bioluminescence traces ($s = 1.67,\ N=71$) from the application of increasing concentrations of small molecule were fit using the method described in a previous section.
+Because the small molecules were toxic to the cells at very high concentrations, experiments were removed from further analysis when the $R^2$ of the sinusoidal regression fell below $0.80$.
+
+\subsection*{{In silico} prediction of small molecule experiments}
+A previously published mathematical model of circadian rhythms was used to predict the effects on population damping rate from the dose-dependent small molecule experiments \cite{Hirota2012}.
+The parameters used to capture the effects of each small molecule were the same as described previously \cite{St.John2014}.
+The model was converted to a stochastic biochemical system and subsequently simulated using StochKit2 \cite{Sanft2011a}.
+Population-level rhythms were found by taking the average of $1,000$ noninteracting oscillators, starting at identical initial conditions.
+The only parameter left unspecified by the deterministic model was the cell volume $\Omega$, which controlled the amount of noise in the system.
+For a given $\Omega$, the population-level damping rate was found by finding the phase diffusivity parameter which gave the closest fit to the simulated population-level rhythms, as described in \cite{St.John2014a}.
+An $R^2$ value was calculated for each fit, taking into account all eight state variables.
+
+\subsubsection*{Fitting the volume parameter}
+We calculated an average experimental damping rate of $d = 0.0151$ from the $0 \mu M$ bioluminescence trajectories for both KL001 and longdaysin.
+{\itshape In silico} damping rates were calculated for logarithmically space values of $\Omega \in (100, 500)$.
+Ten independent groups of $1,000$ oscillators were simulated for each $\Omega$, from which the means and standard errors were found.
+Simulations in which $R^2 < 0.90$ were removed from further analysis.
+A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight.
+The best fit was found to be $\Omega = 226.3 \pm 9.0$.
+
+\subsubsection*{Parameter knockdown experiments}
+We mathematically replicated the effects of the small molecules KL001 and longdaysin through the reductions of the $vdcn$ and $vac1p$ parameters, respectively.
+Simulations were performed with $20$ values of each parameter, linearly spaced between $100\%$ and $15\%$ of their nominal value.
+Similar to the volume calibration simulations, $10$ independent groups of $1,000$ oscillators were simulated from an initially entrained state.
+Means and standard errors in period and damping rate were calculated from each trajectory.
+Simulations in which $R^2 < 0.90$ were removed from further analysis.
+
+\subsection*{Fitting the genome-wide siRNA screen}
+We analyzed the supplemental annotations and data for the $111,743$ wells ($s=2,\ N=72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
+Fits for which the $R^2 < 0.80$ were discarded.
+We rescaled amplitude values on a logarithmic basis, as these were more normally distributed.
+Parameters were normalized on a plate-by-plate basis using a robust z-score, median $\pm$ median absolute deviation \cite{Birmingham2009}.
+We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
+
+\subsubsection*{Detecting outlier perturbations}
+To find which target genes confidently shifted the fitted parameters away from the control population, we performed a test similar to a two-sample t-test with equal variance.
+Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the outlier test is dependent almost entirely on the control distribution and the perturbed mean.
+We found the average perturbed response by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
+To extend the t-test to two dimensions, we used a one-class support vector machine (SVM) with a radial basis function kernel ($\nu = 0.1,\ \gamma = 0.2$).
+The one-class SVM, after training with the control dataset, draws a decision boundary around $90\%$ of the control points, flagging the remaining $10\% \ (n = 1124)$ as outliers.
+The decision boundary was then used to classify the averaged perturbed points, from which $43\% \ (n = 3139)$ were flagged as outliers.
 
@@
 
-% \bibliographystyle{biophysj.bst}
-% \bibliography{condensed_library.bib}
+\bibliographystyle{msb}
+\bibliography{library.bib}
 
 \beginsupplement
+
+\begin{figure}[tbp]
+  \begin{center}
+    \includegraphics[]{figures/pdfs/small_molecule_r2.pdf}
+  \end{center}
+  \caption{{\bfseries Figure Title.}
+({\bfseries A}) Part A.
+({\bfseries B}) Part B.}
+\label{fig:small_molecule_r2}
+\end{figure}
 
@@
 \toprule
-\textbf{Dep. Variable:}    &      decay       & \textbf{  R-squared:         } &$     0.169  $\\
+\textbf{Dep. Variable:}    &      Damping Rate       & \textbf{  R-squared:         } &$     0.169  $\\
 \textbf{Model:}            &       OLS        & \textbf{  Adj. R-squared:    } &$     0.169  $\\

### 2015-02-19 main_text.tex
Adding some supplementary figures, further additions to text

@@
 The $x_i$ data points are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{s}\right)^4$, in which $s$ is the sampling rate (in hours) \cite{Ravn2002}.
-The detrended data is then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1)$.
+The detrended data is then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1$).
 We denote the detrended and filtered experimental data by $y_i(t_i)$.
@@
 An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
-From these three noise metric, a combined noise variable was constructed via a principle component analysis (using scikit-learn, \url{http://scikit-learn.org/}).
+From these three noise variables, a combined noise metric was constructed by projecting the variables along their first principle component (using scikit-learn, \url{http://scikit-learn.org/}).
 Cells were ranked according to this combined noise metric, and a high-noise group and low-noise group were constructed by taking the 39 highest-noise and lowest-noise cells, respectively.
@@
 Bioluminescence traces ($s = 1.67,\ N=71$) from the application of increasing concentrations of small molecule were fit using the method described in a previous section.
-Because the small molecules were toxic to the cells at very high concentrations, experiments were removed from further analysis when the $R^2$ of the sinusoidal regression fell below $0.80$.
+Because the small molecules were toxic to the cells at very high concentrations, experiments were removed from further analysis when the $R^2$ of the sinusoidal regression fell below $0.80$ (Figure~\ref{fig:small_molecule_r2}).
 
@@
 Simulations in which $R^2 < 0.90$ were removed from further analysis.
-A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight.
+A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight (Figure~\ref{fig:vol_calibration}).
 The best fit was found to be $\Omega = 226.3 \pm 9.0$.
@@
 We rescaled amplitude values on a logarithmic basis, as these were more normally distributed.
-Parameters were normalized on a plate-by-plate basis using a robust z-score, median $\pm$ median absolute deviation \cite{Birmingham2009}.
+Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
+Parameters were therefore normalized on a plate-by-plate basis using a robust z-score, median $\pm$ median absolute deviation \cite{Birmingham2009}. 
 We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
@@
 
+\subsection*{Single-cell stochastic noise is manifested in damped population-level rhythms}
+
+We first sought to determine whether single-cell stochastic noise affects the population-level damping rate in cultured bioluminescence cells.
+To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
+We sorted the cells into two groups, a low-noise group and a high-noise group, based on the relative high-frequency noise, period variability, and amplitude variability present in each trace.
+Example rhythms from cells in both groups are shown in Figure~\ref{fig:fibroblast_noise}A.
+Because the cells were not synchronized at the start of the recording, we replicate this effect {\itshape in silico} by shifting each series in time to align their start phases.
+We next calculated expected population-level bioluminescence traces by averaging the cellular PER2::LUC signal in each group.
+Both populations displayed averaged rhythms which resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures .
+Fitting the averaged expression of each group with a damped sinusoid revealed that the low-noise group also had a lower damping rate (Figure~\ref{fig:fibroblast_noise}B).
+The significance of this difference was confirmed via a bootstrap analysis (Figure~\ref{fig:fibroblast_noise}C), where in each bootstrap trial cells were randomly assigned to either the low-noise or the high-noise group.
+
 \begin{figure}[tbp]
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.
-({\bfseries C}) Part C.}
+  \caption{{\bfseries Single-cell bioluminescence recordings show that higher stochastic noise results in faster damping at the population-level.}
+  Data on the bioluminescence of single-cell fibroblasts was taken from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
+({\bfseries A}) Cells were sorted into two groups depending on their degree of stochastic noise.
+An example trace from each of the two groups is shown, demonstrating different levels of noise present in the dataset.
+({\bfseries B}) After artificially synchronizing each cell, averaged bioluminescence rhythms of each group (solid lines) resemble damped recordings taken at cell culture level.
+A damped sinusoid fit to both both groups reveals a difference in damping rate, demonstrated by fitted envelope functions ($\pm A\exp{-dt}$, dashed lines).
+({\bfseries C}) The observed absolute difference in damping rate was shown to be significant ($p = 0.0264$) using $10,000$ bootstrap trials.}
 \label{fig:fibroblast_noise}
 \end{figure}
+
+\subsection*{Clock perturbations can change single-cell stochastic noise}
+
+% \subsubsection*{A small molecule causes dose-dependent changes in the population-level damping rate of cultured cells}
+
+We next demonstrate that perturbations to the transcriptional oscillator are capable of altering single-cell level stochastic noise.
+The actions small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
+By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence}).
+This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc}), indicating it is a fundamental property of the overall gene regulatory network.
 
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.}
+  \caption{{\bfseries Small molecule modulator KL001 increases damping rate in a dose-dependent fashion.}
+  Experimental data on the dose-dependent effect of small molecules KL001 and Longdaysin on cultured circadian reporter cells was taken from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.
+({\bfseries A}) Detrended bioluminescence signals from the two cells and two reporter systems, normalized by the fitted amplitude, period, and phase.
+The normalized bioluminescence highlights the dose-dependent change in damping rate seen with the KL001 application, but not with longdaysin.
+({\bfseries B}) Quantification of the dose-dependent change in damping rate caused by small molecule modulators.
+While both molecules lead to a dose-dependent increase in period, only KL001 shows a reliable change in damping rate.}
 \label{fig:dose_dependence}
 \end{figure}
+
+% \subsection*{Mathematical model predicts KL001 damping rate increase comes from increased single-cell stochastic noise}
+
+To lend further support to the conclusion that the dose-dependent change in damping rate from KL001 is due to changes in single-cell noise characteristics, we employed a model of circadian rhythms previously used to explain the effects of both small molecule perturbations \cite{St.John2014}.
+In order to capture changes to noise characteristics, the model had to first be converted to a stochastic biochemical system.
+Population-level rhythms were therefore generated by averaging the rhythms of $1,000$ individual, noninteracting oscillators.
+The only free parameter in converting the existing deterministic model to a stochastic one is the cell volume, which was determined by fitting the observed population-level damping rate to the experimental unperturbed controls (Figure~\ref{fig:vol_calibration}).
+
+The model was then used to predict the effects of KL001 and longdaysin administration on single-cell noise and population damping rates (Figure~\ref{fig:simulation}A).
+Reductions in parameters previously attributed to the activities of each small molecule caused dose dependent changes in period and damping rate at the population-level which closely matched experimental results (Figure~\ref{fig:simulation}B).
+As the model includes no cell to cell communication, this difference is manifested solely by changing the noise characteristics of individual cells.
 
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.}
+  \caption{{\bfseries Mathematical model accurately predicts dose-dependent changes in damping rate.}
+({\bfseries A}) Example single-cell trajectories (top) and population-averaged trajectories (mean of $1,000$ cells, bottom) of cells under various treatments.
+Cells with the nominal parameter set (left, black) closely match the experimental damping rate for unperturbed cells.
+Cells with simulated KL001 action (red, center) are noisier at the single-cell level, and show faster damping at the population-level.
+Cells with simulated longdaysin action show slightly more accurate single-cell rhythms, with a corresponding dip in the population-level damping rate.
+({\bfseries B}) The model accurately predicts the general trend of period vs.\ damping rate for both KL001 and longdaysin perturbations.
+Experimental data points represent the mean of two replications at each concentration.
+Computational data points represent the mean of ten independent population simulations.}
 \label{fig:simulation}
@@
 
+\subsection*{Genome-wide effects of siRNA knockdown on single-cell stochastic noise}
+
+Unlike with single-cell imaging of fibroblast rhythms, inferring stochastic noise from the desynchronization rate of population-level recordings can be applied to existing and future high-throughput circadian screens.
+We demonstrate the insights that can be gained from such an approach by analyzing the publicly available genome-wide siRNA screen from Zhang {\itshape et al.,} 2009 \cite{Zhang2009}.
+The results of fitting a damped sinusoid to each of the $111,743$ bioluminescence trajectories is shown in Figure~\ref{fig:fit_quality}, in which $86\%$ of fits had an $R^2 > 0.80$ and were kept for further analysis.
+Additionally, of the fits with a high $R^2$ value, only a small minority had a negative damping rate ($0.1\%$): lending further support to the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
+
 \begin{figure}[tbp]
   \begin{center}
-    \includegraphics[]{figures/pdfs/r2_vs_decay.pdf}
+    \includegraphics[]{figures/pdfs/r2_vs_damping.pdf}
   \end{center}

### 2015-02-20 main_text.tex
main drafting done, starting self-edits

@@
 \newcommand{\beginsupplement}{%
+        \clearpage
         \setcounter{table}{0}
@@
 Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
-Parameters were therefore normalized on a plate-by-plate basis using a robust z-score, median $\pm$ median absolute deviation \cite{Birmingham2009}. 
+Parameters were therefore normalized on a plate-by-plate basis using a robust z-score \cite{Birmingham2009}:
+\[
+  z_{R,i} = \frac{x_i - M(x_i)}{M\left(\left|x_i - M(x_i)\right|\right)}, \qquad i \in \{0, \dots, {\cal P} - 1 \}
+\]
+where $M(\cdot)$ denotes the median of a vector, and $x_i$ contains all the points in one plate, and $\cal P$ is the number of plates.
+We removed outlier points prior to calculating the moments of the distributions, Pearson correlation coefficients, and performing the multivariable linear regression.
+Outliers were defined as points which contained any z-metric with an absolute value greater than eight.
 We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
@@
 \subsubsection*{Detecting outlier perturbations}
-To find which target genes confidently shifted the fitted parameters away from the control population, we performed a test similar to a two-sample t-test with equal variance.
-Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the outlier test is dependent almost entirely on the control distribution and the perturbed mean.
 We found the average perturbed response by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
-To extend the t-test to two dimensions, we used a one-class support vector machine (SVM) with a radial basis function kernel ($\nu = 0.1,\ \gamma = 0.2$).
-The one-class SVM, after training with the control dataset, draws a decision boundary around $90\%$ of the control points, flagging the remaining $10\% \ (n = 1124)$ as outliers.
-The decision boundary was then used to classify the averaged perturbed points, from which $43\% \ (n = 3139)$ were flagged as outliers.
+A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
+A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
+Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), we assume the combined covariance is equal to the covariance of the control sample alone.
 
@@
 
-Unlike with single-cell imaging of fibroblast rhythms, inferring stochastic noise from the desynchronization rate of population-level recordings can be applied to existing and future high-throughput circadian screens.
+Unlike from using single-cell imaging, inferring stochastic noise from the desynchronization rate of population-level recordings can be applied to existing and future high-throughput circadian screens.
 We demonstrate the insights that can be gained from such an approach by analyzing the publicly available genome-wide siRNA screen from Zhang {\itshape et al.,} 2009 \cite{Zhang2009}.
-The results of fitting a damped sinusoid to each of the $111,743$ bioluminescence trajectories is shown in Figure~\ref{fig:fit_quality}, in which $86\%$ of fits had an $R^2 > 0.80$ and were kept for further analysis.
+The results of fitting a damped sinusoid to each of the $111,743$ bioluminescence trajectories is shown in Figure~\ref{fig:fit_quality}, in which $86\%$ of fits had an $R^2 > 0.8$.
+Since sinusoidal parameters can only be confidently inferred for fits in which the $R^2$ was sufficiently high, wells were removed from further analysis if $R^2 < 0.8$.
 Additionally, of the fits with a high $R^2$ value, only a small minority had a negative damping rate ($0.1\%$): lending further support to the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.}
+  \caption{{\bfseries Fit quality vs.\ damping rate for the genome-wide siRNA screen.}
+({\bfseries A}) A plot of the $111,743$ individual fits shows that the majority of fits have a high $R^2$ value and positive damping rate. Only fits with $R^2 > 0.8$ were kept for further analysis.
+({\bfseries B}) Examples chosen randomly from each of the four quadrant regions in (A). Sinusoidal parameters for fits 1-2 can be more confidently inferred than those for fits 3-4.}
 \label{fig:fit_quality}
 \end{figure}
+
+We next checked variation of the parameters on a plate-to-plate and well-to-well basis.
+Well-to-well variation was relatively absent, aside from expected variation due to long- and short-period controls (Figure~\ref{fig:well_variation}).
+Fits were normalized to remove plate-to-plate variation (Figure~\ref{fig:plate_variation}) using a robust z-score \cite{Birmingham2009}.
+Additionally, we separated wells into a ``perturbed'' category and ``control'' category, depending on whether or not the well contained an siRNA perturbation.
+As we show in Figure~\ref{fig:fit_distributions}, all fitted parameters displayed normal-like distributions, in which the control distributions showed much tighter clustering around the most likely values.
+Quantifications of the mean, variance, skew, and kurtosis for each distribution are shown in Table~\ref{tab:fit_distributions}.
 
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.}
+  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} Distributions in normalized parameter values closely resemble Gaussian distributions.
+For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.
+}
 \label{fig:fit_distributions}
@@
   \end{center}
-  \caption{{\bfseries Fitted Parameters}}
+  \caption{{\bfseries Moments of the fitted parameter distributions (Figure~\ref{fig:fit_distributions}) after normalization and outlier removal.} Parameters were normalized by subtracting the median and dividing by the median absolute deviation on a plate-by-plate basis.
+Control distributions were less skewed and more peaked then their perturbed counterparts.}
   \label{tab:fit_distributions}
 \end{table}
+
+\subsubsection*{Damping rate is independent of other sinusoidal parameters}
+
+Since an siRNA's effect on period length does not effectively predict its effect on amplitude or phase (and vice-versa), we hypothesized that stochastic noise would similarly be independently affected by siRNA perturbations.
+Low Pearson correlation coefficients between normalized parameter distributions defend this hypothesis, with the highest correlations among variables seen between amplitude and damping rate ($\rho = 0.285$, Table~\ref{tab:corr}).
+Additionally, a multivariate linear regression of damping rate as a function of period, amplitude, phase, and perturbation type (control or perturbed, a categorical variable) produced an $R^2$ value of only $0.169$ (Table~\ref{tab:ols_reg}).
 
@@
   \end{center}
-  \caption{{\bfseries Correlation matrix}}
+  \caption{{\bfseries Correlation among normalized parameters of the high-throughput siRNA screen.} Pearson correlation coefficients are relatively low between fitted parameters, indicating that changes to damping rate (and thereby stochastic noise) are not explained by changes to period, amplitude, or phase.}
   \label{tab:corr}
 \end{table}
+
+\subsubsection*{The unperturbed clock may lie at the Pareto frontier of high amplitude and low noise}
+
+The ability of a population of oscillators to maintain robust oscillations is a function both of its initial amplitude as well as its damping rate.
+A robust oscillator is therefore one with high amplitude and low stochastic noise.
+Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate perturbations tend to shift the clock towards less robust regions of oscillations.
+In order to find perturbations which confidently change robustness, we grouped the siRNA perturbations by target gene and performed a two-sample Hotelling's $T^2$ test against the control population.
+The resulting significant gene perturbations ($75\%$ of all genes) are shown in Figure~\ref{fig:outlier_dist}C.
+Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations, $3.3\%$, simultaneously increase amplitude and decrease stochastic noise.
+It therefore appears that the unperturbed clock optimizes some trade-off between oscillator amplitude and stochastic noise, such that it is not feasible to make oscillations more robust by knocking down one gene at a time.
 
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.}
+  \caption{{\bfseries Effects of siRNA knockdowns on amplitude and damping rate.}
+  Clock robustness is a function of both amplitude and damping rate.
+  Distributions in amplitude and damping rate for control wells ({\bfseries A}) or perturbed wells ({\bfseries B}) indicate that perturbations tend to shift the clock towards regions of higher damping rate or lower amplitude.
+Green dots in each figure indicates the mean of the control population.
+({\bfseries C}) Averaged effect of siRNA knockdown after grouping the perturbed population by Gene ID.
+Only those genes which were significantly different from the control distribution are shown (Hotelling's $T^2$ test, $\alpha = 0.01$).
+({\bfseries D}) Radial histogram of the significant gene perturbations shown in (C).
+The area of each slice is proportional to the frequency of perturbations away from the mean in that direction.
+Very few gene knockdowns resulting in both higher amplitude and lower damping rate (red slices, lower right quadrant).
+}
 \label{fig:outlier_dist}
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.}
+  \caption{{\bfseries Fit quality for the dose-dependent small molecule screens.} The bioluminescence rhythms in both reporter systems were well-described by a damped sinusoid.
+As the molecules were toxic to the reporter cells at high concentrations, fit quality declined with increasing dosage.
+Only fits with $R^2 > 0.8$ were kept for further analysis.}
 \label{fig:small_molecule_r2}
@@
   \end{center}
-  \caption{{\bfseries Figure Title.}
-({\bfseries A}) Part A.
-({\bfseries B}) Part B.}
+  \caption{{\bfseries Calibration curve for fitting the volume parameter to experimental data.} The model's volume parameter was linearly related to the population-averaged damping rate on a log-log scale.
+Error bars represent the standard error of the mean, calculated by $10$ independent replicates for each volume.
+Points shown in gray had an average $R^2 < 0.9$ and were excluded from the linear regression.
+Solid and dashed grey lines indicate the mean and $95\%$ confidence intervals of the linear regression, respectively.}
 \label{fig:vol_calibration}
@@
   \end{center}
-  \caption{{\bfseries figure title.}
-({\bfseries a}) part a.
-({\bfseries b}) part b.}
+  \caption{{\bfseries Plate-to-plate variation of fitted parameters in the Zhang {\itshape et al.,} 2009 genome-wide siRNA screen.} Dots indicate the median of each plate, with lines extending from the $5^\text{th}$ to $95^\text{th}$ percentile.
+While parameter fits were largely consistent between plates, some plates had definitively different value ranges.
+In order to accurately compare perturbations and controls between plate experiments, we normalized fitted parameters on a plate-by-plate basis.}
 \label{fig:plate_variation}
@@
   \end{center}
-  \caption{{\bfseries figure title.}
-({\bfseries a}) part a.
-({\bfseries b}) part b.}
+  \caption{{\bfseries Well-to-well variation in the Zhang {\itshape et al.,} 2009 genome-wide siRNA screen.} Similar to Figure~\ref{fig:plate_variation}, dots indicate the median value of each parameter in each well, with lines extending from the $5^\text{th}$ to $95^\text{th}$ percentile.
+Well position did not seem to affect the fitted values, particularly in the middle regions which contained the siRNA knockdown library.
+Wells on either end showed significant variation, but these are likely due to the spotting of long- and short-period controls in the same position on each plate.}
 \label{fig:well_variation}
@@
 \bottomrule
-\end{tabular}
-\begin{tabular}{lccccc}
+\end{tabular}\\[5ex]
+\begin{tabular}{lccccc}\toprule
                    & \textbf{coef} & \textbf{std err} & \textbf{t} & \textbf{P$>$$|$t$|$} & \textbf{[95.0\% Conf. Int.]}  \\
 \midrule
-\textbf{const}     &     $-0.0370$ &       $0.014$    &   $-2.572$ &        $0.010$       &       $-0.065,  -0.009$      \\
+\textbf{intercept}     &     $-0.0370$ &       $0.014$    &   $-2.572$ &        $0.010$       &       $-0.065,  -0.009$      \\
 \textbf{amplitude} &     $ 0.2375$ &       $0.003$    &   $86.282$ &        $0.000$       &       $ 0.232,   0.243$      \\
@@
 \textbf{phase}     &     $-0.2354$ &       $0.003$    &   $85.598$ &        $0.000$       &       $-0.241,  -0.230$      \\
-\textbf{type}      &     $ 0.3197$ &       $0.015$    &   $20.664$ &        $0.000$       &       $ 0.289,   0.350$      \\
+\textbf{perturbation type}      &     $ 0.3197$ &       $0.015$    &   $20.664$ &        $0.000$       &       $ 0.289,   0.350$      \\
 \bottomrule
-\end{tabular}
-\begin{tabular}{lclc}
+\end{tabular}\\[5ex]
+\begin{tabular}{lclc}\toprule
 \textbf{Omnibus:}       &$9769.391$& \textbf{  Durbin-Watson:     } &$    1.876$ \\
@@
 \end{center} 
-\caption{OLS Regression Results}
+\caption{{\bfseries Multivariable linear regression results.} Fit statistics which demonstrate the effect of each other fitted parameter on damping rate.
+Of particular note is the perturbation type categorical variable, which demonstrates that the presence of siRNA perturbation increases damping rate on average.
+Higher amplitude is also correlated with higher damping rate.
+However, in total damping rate is poorly predicted by the other fitted variables $R^2 = 0.169$, indicating it describes an independent oscillatory feature.}
 \label{tab:ols_reg}

### 2015-02-20 main_text.tex
adding abstract and conclusion

@@
 \begin{abstract}
-
+Stochastic noise at the cellular level has been shown to play a fundamental role in circadian oscillations, influencing how groups of cells entrain to external cues and perhaps serving as the mechanism by which cell-autonomous rhythms are generated.
+Despite this importance, few studies have investigated how clock perturbations affect stochastic noise - even as increasing numbers of high-throughput screens categorize how gene knockdowns or small molecules can change clock period and amplitude.
+This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which requires the careful collection and processing of single-cell data.
+In this study, we show that since cultured circadian reporters lack significant intercellular coupling, the damping rate of population-level bioluminescence recordings can serve as an accurate measure of cell-autonomous stochastic noise which can be applied to future and existing high-throughput circadian screens.
+Using cell-autonomous fibroblast data, we first show directly that higher noise at the single-cell results in faster damping at the population level.
+Next, we show that the damping rate of cultured cells can be changed in a dose-dependent fashion by small molecule modulators, and confirm that such a change can be explained by single-cell noise using a mathematical model.
+We further demonstrate the insights that can be gained by applying such a method to a genome-wide siRNA screen, revealing that stochastic noise is altered independently from period, amplitude, and phase.
+Finally, we hypothesize that the unperturbed clock may be highly optimized for robust rhythms, as very few gene perturbations are capable of simultaneously increasing amplitude and lowering stochastic noise.
+Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on cell-autonomous noise, especially with regard to the development of small-molecule circadian therapeutics.
 \end{abstract}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/mainfig_dose.pdf}
+    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/mainfig_dose.pdf}}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/main_fig_simulations.pdf}
+    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/main_fig_simulations.pdf}}
   \end{center}
@@
 
+In this study, we have described a method by which stochastic noise can be estimated from population-level bioluminescence recordings.
+As high amplitude circadian oscillations are important for maintaining metabolic health, many studies have sought to find small molecule candidates which increase oscillatory amplitude \cite{Chen2013}.
+In the search for circadian clock therapeutics, high-throughput methods are frequently used to screen for such drugs, often neglecting potential effects on cell-autonomous noise.
+While we have demonstrated a method by which the effects of small molecules on noise can be inferred from high-throughput methods, we have also shown that the potential improvement of clock robustness may be limited.
+Amplitudes of circadian rhythms may be best bolstered by small molecule therapies by finding drugs which act transiently to synchronize peripheral oscillators.
+While such a method would require accurate alignment of drug administration to the correct circadian phase, a recent {\itshape in silico} study has demonstrated the potential effectiveness of such an approach in improving amplitudes in peripheral tissues \cite{St.John2014a}.
+
 \section*{Acknowledgments}

### 2015-02-21 main_text.tex
some updates from the starbucks

@@
 This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which requires the careful collection and processing of single-cell data.
-In this study, we show that since cultured circadian reporters lack significant intercellular coupling, the damping rate of population-level bioluminescence recordings can serve as an accurate measure of cell-autonomous stochastic noise which can be applied to future and existing high-throughput circadian screens.
+% since cultured circadian reporters lack significant intercellular coupling,
+In this study, we show that the damping rate of population-level bioluminescence recordings can serve as an accurate measure of cell-autonomous stochastic noise, and one which can be applied to future and existing high-throughput circadian screens.
 Using cell-autonomous fibroblast data, we first show directly that higher noise at the single-cell results in faster damping at the population level.
@@
 The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
-Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998}, and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
+Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
 Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
@@
 As a result, bioluminescence traces of individual cells are stochastic, with significant period-to-period variability \cite{Welsh2004}. 
-The collective behavior of thousands of cells results in more reliable oscillations at the tissue-level, especially in the suprachiasmatic nucleus (SCN) where cell-to-cell coupling keeps individual oscillators at a consistent phase \cite{Herzog2004}.
-While communication between cells is common in many tissues, it is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues or cultured reporter cells, is very weak, if present \cite{Guenthner2014, Noguchi2013}.
+The collective behavior of thousands of cells results in more reliable oscillations at the tissue-level, especially in the suprachiasmatic nucleus (SCN) where cell-to-cell coupling keeps individual oscillators in tight synchrony \cite{Herzog2004}.
+While communication between cells is common in many tissues, it is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues or cultured reporter cells, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
 In populations which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level \cite{Welsh2004}. 
-Despite the averaging which occurs at the population-level, stochastic noise at the single-cell level plays an important role in determining the function of the circadian oscillator.
+Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
 A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues \cite{An2013}.
-Additional studies have even argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators - when simulated stochastically - better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
+Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
 
@@
 While such a method would require accurate alignment of drug administration to the correct circadian phase, a recent {\itshape in silico} study has demonstrated the potential effectiveness of such an approach in improving amplitudes in peripheral tissues \cite{St.John2014a}.
+Finally, the ability to extract one more biologically relevant parameter from existing datasets will prove useful for many studies looking to differentiate between otherwise similar perturbation effects.
 

### 2015-02-23 main_text.tex
more edits

@@
 
-Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful plating and recording of fibroblast cells and subsequent image processing \cite{Leise2012}.
+Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of fibroblast cells and subsequent image processing \cite{Leise2012}.
 As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.
 In this study, we demonstrate single-cell stochasticity can be reliably inferred from the damping rate of population-level bioluminescence recordings.
-Additionally, we show that a small-molecule modulator is able to change circadian noise in a dose-dependent fashion. 
+First, we show using cell-autonomous fibroblast data that stochastic noise is directly related to population-level damping.
+Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to single-cell stochasticity is a likely mechanism.
 Finally, we calculate the genome-wide effects of siRNA knockdown on single-cell noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
+Using this additional information, we show that circadian rhythms have likely evolved to an optimal point of high amplitude and low stochastic noise.
 Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on single-cell noise to be quantified in a high-throughput manner.
@@
 \subsection*{Fitting a damped sinusoid to experimental data}
+Raw experimental data $x_i(t_i), \  i \in \{0, \dots, N-1\}$ are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{s}\right)^4$, in which $s$ is the sampling rate (in hours) \cite{Ravn2002}.
+The detrended data are then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1$).
+We denote the detrended and filtered experimental data by $y_i(t_i)$.
 A damped sinusoid, specified by:
@@
 \]
-is fit to experimental data $x_i(t_i), \  i \in \{0, \dots, N-1\}$.
-The $x_i$ data points are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{s}\right)^4$, in which $s$ is the sampling rate (in hours) \cite{Ravn2002}.
-The detrended data is then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1$).
-We denote the detrended and filtered experimental data by $y_i(t_i)$.
+is then fit to the filtered data.
 For numerical efficiency, the period, $T$, and damping rate, $d$, parameters are fit first using a matrix pencil method \cite{Hua1990}, reviewed in \cite{Zielinski2011}.
@@
 To ensure the difference in damping rate between groups was statistically significant, a bootstrap analysis was performed.
-In each of $10,000$ bootstrap trials, the cells were randomly assigned evenly to either the low-noise or the high-noise group.
+In each of $10,000$ bootstrap trials, the cells were randomly assigned evenly to either the low-noise or the high-noise group (one cell was omitted randomly in each trial to ensure even group sizes).
 The absolute difference in damping rate between the two populations was recorded to yield a two-tailed test.
@@
 
+\begin{table}[p]
+  \centering
+  \titlecaption{A model of the mammalian core circadian feedback loop, from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}}{Lower case letters (p: {\it Per}, c1: {\it Cry1}, c2: {\it Cry2}) are mRNA state variables.
+Uppercase letters (P: PER, C1: CRY1, C2: CRY2) are the free (cytosolic) proteins.
+C1N: CRY1 and C2N: CRY2 are the nuclear proteins.}
+
+  \begin{align}
+    \label{eq:p}
+    \frac{d\mathrm{\bf p}}{dt}    & = \frac{v_{\text{txn,{\bf p}}}}{k_{\text{txn,{\bf p}}} + \left(\text{\bf C1N} + \text{\bf C2N} \right)^3} - \frac{v_{\text{deg,{\bf p}}} \;\text{\bf p} }{k_{\text{deg,{\bf p}}} +\text{\bf p} } \\
+        %
+    \label{eq:c1}
+    \frac{d\text{\bf c1}}{dt}   & = \frac{v_{\text{txn,{\bf c1}}}}{k_{\text{txn,{\bf c}}} + \left(\text{\bf C1N} + \text{\bf C2N} \right)^3} - \frac{v_{\text{deg,{\bf c1}}} \;\text{\bf c1} }{k_{\text{deg,{\bf c}}} +\text{\bf c1} } \\
+        %
+    \label{eq:c2}
+    \frac{d\text{\bf c2}}{dt}   & = \frac{v_{\text{txn,{\bf c2}}}}{k_{\text{txn,{\bf c}}} + \left(\text{\bf C1N} + \text{\bf C2N} \right)^3} - \frac{v_{\text{deg,{\bf c2}}} \;\text{\bf c2} }{k_{\text{deg,{\bf c}}} +\text{\bf c2} } \\
+        %
+    \begin{split}
+      \frac{d\text{\bf P}}{dt}  & = k_\text{tln,{\bf p}} \;\text{\bf p}  - \frac{v_{\text{deg,{\bf P}}} \;\text{\bf P} }{k_{\text{deg,{\bf P}}} +\text{\bf P} } - v_{\text{a,{\bf CP}}} \;\text{\bf P}  \;\text{\bf C1}  + v_{\text{d,{\bf CP}}} \;\text{\bf C1N} \\
+      & \quad - v_{\text{a,{\bf CP}}} \;\text{\bf P}  \;\text{\bf C2}  + v_{\text{d,{\bf CP}}} \; \text{\bf C2N}
+    \end{split} \\
+        %
+    \frac{d\text{\bf C1}}{dt}   & =\text{\bf c1}  - \frac{v_{\text{deg,{\bf C1}}} \;\text{\bf C1} }{k_{\text{deg,{\bf C}}} +\text{\bf C1} } - v_{\text{a,{\bf CP}}} \;\text{\bf P}  \;\text{\bf C1}  + v_{\text{d,{\bf CP}}} \;\text{\bf C1N} \\
+        %
+    \frac{d\text{\bf C2}}{dt}   & =\text{\bf c2} - \frac{ v_{\text{deg,{\bf C2}}} \;\text{\bf C2} }{k_{\text{deg,{\bf C}}} +\text{\bf C2} } - v_{\text{a,{\bf CP}}} \;\text{\bf P}  \;\text{\bf C2}  + v_{\text{d,{\bf CP}}} \; \text{\bf C2N} \\
+        %
+    \frac{d\text{\bf C1N}}{dt}  & = - \frac{ v_{\text{deg,{\bf CP}}} \;\text{\bf C1N} }{k_{\text{deg,{\bf CP}}} +\text{\bf C1N}  + \text{\bf C2N} } + v_{\text{a,{\bf CP}}} \;\text{\bf P}  \;\text{\bf C1} - v_{\text{d,{\bf CP}}} \;\text{\bf C1N} \\
+        %
+    \label{eq:C2N}
+    \frac{d\text{\bf C2N }}{dt} & = - \frac{ \left(v_{\text{deg,{\bf CP}}} \; m_{\text{\bf C2N}}\right) \; \text{\bf C2N} }{k_{\text{deg,{\bf CP}}} + \text{\bf C2N}  +\text{\bf C1N} } + v_{\text{a,{\bf CP}}} \;\text{\bf P}  \;\text{\bf C2} - v_{\text{d,{\bf CP}}} \; \text{\bf C2N}
+  \end{align}
+  \label{mod:hirota}
+\end{table}
+
+\begin{table}[p]
+  \titlecaption{Parameter values for \fref{mod:hirota}}{Parameters for model described in \fref{mod:hirota}.}
+  \label{tab:parset}
+  \vspace{2mm}
+  \centering
+  \begin{tabular}{cllrr} \toprule
+    & Parameter                 & Description                       & Degradation & Activation \\ \midrule
+    1  & $v_{\text{txn,{\bf p}}}$  & {\it Per} Transcription rate      & 0.195       & 0.276 \\
+    2  & $v_{\text{txn,{\bf c1}}}$ & {\it Cry1} Transcription rate     & 0.131       & 0.062 \\
+    3  & $v_{\text{txn,{\bf c2}}}$ & {\it Cry1} Transcription rate     & 0.114       & 0.053 \\
+    4  & $k_{\text{txn,{\bf p}}}$  & {\it Per} Repression constant     & 0.425       & 0.425 \\
+    5  & $k_{\text{txn,{\bf c}}}$  & {\it Cry1/2} Repression constant  & 0.259       & 0.262 \\
+    6  & $v_{\text{deg,{\bf p}}}$  & {\it Per} Max degradation rate    & 0.326       & 0.472 \\
+    7  & $v_{\text{deg,{\bf c1}}}$ & {\it Cry1} Max degradation rate   & 0.676       & 0.322 \\
+    8  & $v_{\text{deg,{\bf c2}}}$ & {\it Cry2} Max degradation rate   & 0.608       & 0.290 \\
+    9  & $k_{\text{deg,{\bf p}}}$  & {\it Per} Degradation constant    & 0.011       & 0.024 \\
+    10 & $k_{\text{deg,{\bf c}}}$  & {\it Cry1/2} Degradation constant & 1.149       & 0.809 \\
+    11 & $v_{\text{deg,{\bf P}}}$  & Max PERc degradation rate         & 2.970       & 2.970 \\
+    12 & $k_{\text{deg,{\bf P}}}$  & PERc degradation constant         & 0.034       & 0.034 \\
+    13 & $v_{\text{deg,{\bf C1}}}$ & Max CRY1c degradation rate        & 1.523       & 1.048 \\
+    14 & $v_{\text{deg,{\bf C2}}}$ & Max CRY2c degradation rate        & 1.686       & 1.134 \\
+    15 & $k_{\text{deg,{\bf C}}}$  & CRYc degradation constant         & 2.017       & 2.028 \\
+    16 & $v_{\text{deg,{\bf CP}}}$ & CRYn degradation rate             & 0.101       & 0.070 \\
+    17 & $m_{\text{\bf C2N}}$      & CRY2n degradation multiplier      & 3.318       & 3.334 \\
+    18 & $k_{\text{deg,{\bf CP}}}$ & CRYn degradation constant         & 0.053       & 0.053 \\
+    19 & $v_{\text{a,{\bf CP}}}$   & CRYn association rate             & 0.041       & 0.028 \\
+    20 & $v_{\text{d,{\bf CP}}}$   & CRYn dissociation rate            & 0.002       & 0.001 \\
+    21 & $k_{\text{tln,{\bf p}}}$  & PER translation rate              & 3.000       & 1.000 \\ \bottomrule
+    \hline
+  \end{tabular}
+\end{table}
+
 \begin{table}

### 2015-02-23 main_text.tex
Further edits, sent to frank

@@
 The parameters used to capture the effects of each small molecule were the same as described previously \cite{St.John2014}.
-The model was converted to a stochastic biochemical system and subsequently simulated using StochKit2 \cite{Sanft2011a}.
+The model was converted to a stochastic biochemical system and subsequently simulated using StochKit2 \cite{Sanft2011a} (via GillesPy, \url{https://github.com/JohnAbel/gillespy}).
 Population-level rhythms were found by taking the average of $1,000$ noninteracting oscillators, starting at identical initial conditions.
@@
 \subsubsection*{Parameter knockdown experiments}
-We mathematically replicated the effects of the small molecules KL001 and longdaysin through the reductions of the $vdcn$ and $vac1p$ parameters, respectively.
-Simulations were performed with $20$ values of each parameter, linearly spaced between $100\%$ and $15\%$ of their nominal value.
-Similar to the volume calibration simulations, $10$ independent groups of $1,000$ oscillators were simulated from an initially entrained state.
-Means and standard errors in period and damping rate were calculated from each trajectory.
+We replicated the effects of the small molecules KL001 and longdaysin mathematically through the reductions of the $vdcn$ and $vac1p$ parameters, respectively (Tables~\ref{tab:model}-\ref{tab:parset}).
+Knockdown simulations were performed with $20$ values of each parameter, linearly spaced between $100\%$ and $15\%$ of their nominal value.
+Similar to the volume calibration simulations, $10$ independent populations of $1,000$ oscillators were simulated from an initially entrained state.
+Means and standard errors in period and damping rate were calculated from each population-averaged trajectory.
 Simulations in which $R^2 < 0.90$ were removed from further analysis.
@@
 \subsection*{Fitting the genome-wide siRNA screen}
-We analyzed the supplemental annotations and data for the $111,743$ wells ($s=2,\ N=72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
+We analyzed the data and annotations for the $111,743$ wells ($s=2,\ N=72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
 Fits for which the $R^2 < 0.80$ were discarded.
-We rescaled amplitude values on a logarithmic basis, as these were more normally distributed.
+We analyze the natural logarithm of the amplitude, as these are more normally distributed and on a similar scale to the damping rate.
 Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
@@
 \[
-  z_{R,i} = \frac{x_i - M(x_i)}{M\left(\left|x_i - M(x_i)\right|\right)}, \qquad i \in \{0, \dots, {\cal P} - 1 \}
+  z_{R,i} = \frac{p_i - M(p_i)}{M\left(\left|p_i - M(p_i)\right|\right)}, \qquad i \in \{0, \dots, {\cal P} - 1 \}
 \]
-where $M(\cdot)$ denotes the median of a vector, and $x_i$ contains all the points in one plate, and $\cal P$ is the number of plates.
+where $M(\cdot)$ denotes the median of a vector, and $p_i$ contains all the points in one plate, and $\cal P$ is the number of plates.
 We removed outlier points prior to calculating the moments of the distributions, Pearson correlation coefficients, and performing the multivariable linear regression.
-Outliers were defined as points which contained any z-metric with an absolute value greater than eight.
+Outliers were defined as points which contained a z-metric (in either period, phase, amplitude, or damping rate) with an absolute value greater than eight.
 We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
@@
 \subsubsection*{Detecting outlier perturbations}
-We found the average perturbed response by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
+We found the average response to each gene perturbation by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
 A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
 A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
-Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), we assume the combined covariance is equal to the covariance of the control sample alone.
+Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), we assume the pooled covariance can be approximated by the covariance of the control sample alone.
 
@@
 
-\subsection*{Single-cell stochastic noise is manifested in damped population-level rhythms}
+\subsection*{Higher noise results in faster damping in population-level rhythms}
 
@@
 We next calculated expected population-level bioluminescence traces by averaging the cellular PER2::LUC signal in each group.
-Both populations displayed averaged rhythms which resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures .
+Both populations displayed averaged rhythms which resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures.
 Fitting the averaged expression of each group with a damped sinusoid revealed that the low-noise group also had a lower damping rate (Figure~\ref{fig:fibroblast_noise}B).
-The significance of this difference was confirmed via a bootstrap analysis (Figure~\ref{fig:fibroblast_noise}C), where in each bootstrap trial cells were randomly assigned to either the low-noise or the high-noise group.
+The significance of this difference was confirmed via a bootstrap analysis (Figure~\ref{fig:fibroblast_noise}C), where cells were randomly assigned in each bootstrap trial to either the low-noise or the high-noise group.
 
@@
 An example trace from each of the two groups is shown, demonstrating different levels of noise present in the dataset.
-({\bfseries B}) After artificially synchronizing each cell, averaged bioluminescence rhythms of each group (solid lines) resemble damped recordings taken at cell culture level.
-A damped sinusoid fit to both both groups reveals a difference in damping rate, demonstrated by fitted envelope functions ($\pm A\exp{-dt}$, dashed lines).
+({\bfseries B}) After artificially synchronizing each cell, we calculate averaged bioluminescence rhythms of each group (solid lines).
+A damped sinusoid fit to both groups reveals a difference in damping rate, demonstrated by fitted envelope functions ($\pm A\exp{-dt}$, dashed lines).
 ({\bfseries C}) The observed absolute difference in damping rate was shown to be significant ($p = 0.0264$) using $10,000$ bootstrap trials.}
@@
 We next demonstrate that perturbations to the transcriptional oscillator are capable of altering single-cell level stochastic noise.
-The actions small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
+The actions of small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
 By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence}).
@@
   \caption{{\bfseries Small molecule modulator KL001 increases damping rate in a dose-dependent fashion.}
-  Experimental data on the dose-dependent effect of small molecules KL001 and Longdaysin on cultured circadian reporter cells was taken from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.
-({\bfseries A}) Detrended bioluminescence signals from the two cells and two reporter systems, normalized by the fitted amplitude, period, and phase.
-The normalized bioluminescence highlights the dose-dependent change in damping rate seen with the KL001 application, but not with longdaysin.
+  Experimental data on the dose-dependent effects of small molecules KL001 and longdaysin on cultured circadian reporter cells was taken from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.
+({\bfseries A}) Detrended bioluminescence signals from the two reporter systems and two small molecules are shown normalized by the fitted amplitude, period, and phase.
+The normalized bioluminescence highlights the dose-dependent change in damping rate seen with the KL001 application (top), but not with longdaysin (bottom).
 ({\bfseries B}) Quantification of the dose-dependent change in damping rate caused by small molecule modulators.
@@
 
-To lend further support to the conclusion that the dose-dependent change in damping rate from KL001 is due to changes in single-cell noise characteristics, we employed a model of circadian rhythms previously used to explain the effects of both small molecule perturbations \cite{St.John2014}.
-In order to capture changes to noise characteristics, the model had to first be converted to a stochastic biochemical system.
-Population-level rhythms were therefore generated by averaging the rhythms of $1,000$ individual, noninteracting oscillators.
-The only free parameter in converting the existing deterministic model to a stochastic one is the cell volume, which was determined by fitting the observed population-level damping rate to the experimental unperturbed controls (Figure~\ref{fig:vol_calibration}).
+To lend further support to the conclusion that the dose-dependent changes in damping rate from KL001 is due to changes in single-cell noise characteristics, we employed a model of circadian rhythms previously used to explain the effects of both small molecule perturbations (Tables~\ref{tab:model}-\ref{tab:parset}) \cite{St.John2014}.
+In order to capture changes to noise characteristics, we first converted the model to a stochastic biochemical system.
+Population-level rhythms were generated by averaging the trajectories of $1,000$ individual, noninteracting oscillators.
+The only free parameter in converting the existing deterministic model to a stochastic one is the cell volume, which was determined by fitting the observed population-level damping rate to that of the experimental control traces (Figure~\ref{fig:vol_calibration}).
 
 The model was then used to predict the effects of KL001 and longdaysin administration on single-cell noise and population damping rates (Figure~\ref{fig:simulation}A).
-Reductions in parameters previously attributed to the activities of each small molecule caused dose dependent changes in period and damping rate at the population-level which closely matched experimental results (Figure~\ref{fig:simulation}B).
-As the model includes no cell to cell communication, this difference is manifested solely by changing the noise characteristics of individual cells.
+Reductions in parameters previously attributed to the activities of each small molecule caused dose-dependent changes in period and damping rate at the population-level that closely matched experimental results (Figure~\ref{fig:simulation}B).
+As the model includes no cell-to-cell communication, this difference is manifested solely by changing the noise characteristics of individual cells.
 
@@
   \caption{{\bfseries Mathematical model accurately predicts dose-dependent changes in damping rate.}
-({\bfseries A}) Example single-cell trajectories (top) and population-averaged trajectories (mean of $1,000$ cells, bottom) of cells under various treatments.
+({\bfseries A}) Example single-cell trajectories (top) and population-averaged trajectories (bottom, mean of $1,000$ cells) of cells under various treatments.
 Cells with the nominal parameter set (left, black) closely match the experimental damping rate for unperturbed cells.
 Cells with simulated KL001 action (red, center) are noisier at the single-cell level, and show faster damping at the population-level.
-Cells with simulated longdaysin action show slightly more accurate single-cell rhythms, with a corresponding dip in the population-level damping rate.
+Cells with simulated longdaysin action show slightly more accurate single-cell rhythms, with a corresponding reduction in the population-level damping rate.
 ({\bfseries B}) The model accurately predicts the general trend of period vs.\ damping rate for both KL001 and longdaysin perturbations.
@@
 Since sinusoidal parameters can only be confidently inferred for fits in which the $R^2$ was sufficiently high, wells were removed from further analysis if $R^2 < 0.8$.
-Additionally, of the fits with a high $R^2$ value, only a small minority had a negative damping rate ($0.1\%$): lending further support to the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
+Additionally, of the fits with a high $R^2$ value, only a small minority ($0.1\%$) had a negative damping rate.
+This trend lends further support to the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
 
@@
 
-We next checked variation of the parameters on a plate-to-plate and well-to-well basis.
+We next checked how parameters varied on a plate-to-plate and well-to-well basis.
 Well-to-well variation was relatively absent, aside from expected variation due to long- and short-period controls (Figure~\ref{fig:well_variation}).
@@
   \end{center}
-  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} Distributions in normalized parameter values closely resemble Gaussian distributions.
+  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} Distributions in robust z-scores closely resemble normal distributions.
 For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.
@@
   \caption{{\bfseries Moments of the fitted parameter distributions (Figure~\ref{fig:fit_distributions}) after normalization and outlier removal.} Parameters were normalized by subtracting the median and dividing by the median absolute deviation on a plate-by-plate basis.
-Control distributions were less skewed and more peaked then their perturbed counterparts.}
+Control distributions had less variance, were less skewed, and were more peaked then their perturbed counterparts.}
   \label{tab:fit_distributions}
@@
 A robust oscillator is therefore one with high amplitude and low stochastic noise.
-Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate perturbations tend to shift the clock towards less robust regions of oscillations.
+Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate that few outlier points reside in the lower damping rate, higher amplitude quadrant.
 In order to find perturbations which confidently change robustness, we grouped the siRNA perturbations by target gene and performed a two-sample Hotelling's $T^2$ test against the control population.
 The resulting significant gene perturbations ($75\%$ of all genes) are shown in Figure~\ref{fig:outlier_dist}C.
-Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations, $3.3\%$, simultaneously increase amplitude and decrease stochastic noise.
+Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations ($3.3\%$) simultaneously increase amplitude and decrease stochastic noise.
 It therefore appears that the unperturbed clock optimizes some trade-off between oscillator amplitude and stochastic noise, such that it is not feasible to make oscillations more robust by knocking down one gene at a time.
@@
 The area of each slice is proportional to the frequency of perturbations away from the mean in that direction.
-Very few gene knockdowns resulting in both higher amplitude and lower damping rate (red slices, lower right quadrant).
+Very few gene knockdowns result in both higher amplitude and lower damping rate (red slices, lower right quadrant).
 }
@@
 While we have demonstrated a method by which the effects of small molecules on noise can be inferred from high-throughput methods, we have also shown that the potential improvement of clock robustness may be limited.
-Amplitudes of circadian rhythms may be best bolstered by small molecule therapies by finding drugs which act transiently to synchronize peripheral oscillators.
+Amplitudes of circadian rhythms may therefore be best increased by small molecule therapies which act transiently to synchronize peripheral oscillators.
 While such a method would require accurate alignment of drug administration to the correct circadian phase, a recent {\itshape in silico} study has demonstrated the potential effectiveness of such an approach in improving amplitudes in peripheral tissues \cite{St.John2014a}.
-Finally, the ability to extract one more biologically relevant parameter from existing datasets will prove useful for many studies looking to differentiate between otherwise similar perturbation effects.
+Finally, the ability to extract an additional biologically relevant parameter from existing datasets will likely prove useful for many studies, as it allows further differentiation between perturbations which might otherwise have identical effects on bioluminescence rhythms.
 
@@
 \bibliographystyle{msb}
-\bibliography{library.bib}
+\bibliography{condensed_library.bib}
 
@@
   \caption{{\bfseries Plate-to-plate variation of fitted parameters in the Zhang {\itshape et al.,} 2009 genome-wide siRNA screen.} Dots indicate the median of each plate, with lines extending from the $5^\text{th}$ to $95^\text{th}$ percentile.
-While parameter fits were largely consistent between plates, some plates had definitively different value ranges.
+While parameter fits were of similar magnitude for all plates, some inconsistencies were present.
 In order to accurately compare perturbations and controls between plate experiments, we normalized fitted parameters on a plate-by-plate basis.}
@@
   \centering
-  \titlecaption{A model of the mammalian core circadian feedback loop, from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}}{Lower case letters (p: {\it Per}, c1: {\it Cry1}, c2: {\it Cry2}) are mRNA state variables.
+  \caption{{\bfseries A model of the mammalian core circadian feedback loop, from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.} Lower case letters (p: {\it Per}, c1: {\it Cry1}, c2: {\it Cry2}) are mRNA state variables.
 Uppercase letters (P: PER, C1: CRY1, C2: CRY2) are the free (cytosolic) proteins.
@@
   \end{align}
-  \label{mod:hirota}
+  \label{tab:model}
 \end{table}
@@
 \begin{table}[p]
-  \titlecaption{Parameter values for \fref{mod:hirota}}{Parameters for model described in \fref{mod:hirota}.}
+  \caption{{\bfseries Parameter values for the model in Table~\ref{tab:model}.} Nominal values for the kinetic parameters are shown below, as published in Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.}
   \label{tab:parset}
@@
   \centering
-  \begin{tabular}{cllrr} \toprule
-    & Parameter                 & Description                       & Degradation & Activation \\ \midrule
-    1  & $v_{\text{txn,{\bf p}}}$  & {\it Per} Transcription rate      & 0.195       & 0.276 \\
-    2  & $v_{\text{txn,{\bf c1}}}$ & {\it Cry1} Transcription rate     & 0.131       & 0.062 \\
-    3  & $v_{\text{txn,{\bf c2}}}$ & {\it Cry1} Transcription rate     & 0.114       & 0.053 \\
-    4  & $k_{\text{txn,{\bf p}}}$  & {\it Per} Repression constant     & 0.425       & 0.425 \\
-    5  & $k_{\text{txn,{\bf c}}}$  & {\it Cry1/2} Repression constant  & 0.259       & 0.262 \\
-    6  & $v_{\text{deg,{\bf p}}}$  & {\it Per} Max degradation rate    & 0.326       & 0.472 \\
-    7  & $v_{\text{deg,{\bf c1}}}$ & {\it Cry1} Max degradation rate   & 0.676       & 0.322 \\
-    8  & $v_{\text{deg,{\bf c2}}}$ & {\it Cry2} Max degradation rate   & 0.608       & 0.290 \\
-    9  & $k_{\text{deg,{\bf p}}}$  & {\it Per} Degradation constant    & 0.011       & 0.024 \\
-    10 & $k_{\text{deg,{\bf c}}}$  & {\it Cry1/2} Degradation constant & 1.149       & 0.809 \\
-    11 & $v_{\text{deg,{\bf P}}}$  & Max PERc degradation rate         & 2.970       & 2.970 \\
-    12 & $k_{\text{deg,{\bf P}}}$  & PERc degradation constant         & 0.034       & 0.034 \\
-    13 & $v_{\text{deg,{\bf C1}}}$ & Max CRY1c degradation rate        & 1.523       & 1.048 \\
-    14 & $v_{\text{deg,{\bf C2}}}$ & Max CRY2c degradation rate        & 1.686       & 1.134 \\
-    15 & $k_{\text{deg,{\bf C}}}$  & CRYc degradation constant         & 2.017       & 2.028 \\
-    16 & $v_{\text{deg,{\bf CP}}}$ & CRYn degradation rate             & 0.101       & 0.070 \\
-    17 & $m_{\text{\bf C2N}}$      & CRY2n degradation multiplier      & 3.318       & 3.334 \\
-    18 & $k_{\text{deg,{\bf CP}}}$ & CRYn degradation constant         & 0.053       & 0.053 \\
-    19 & $v_{\text{a,{\bf CP}}}$   & CRYn association rate             & 0.041       & 0.028 \\
-    20 & $v_{\text{d,{\bf CP}}}$   & CRYn dissociation rate            & 0.002       & 0.001 \\
-    21 & $k_{\text{tln,{\bf p}}}$  & PER translation rate              & 3.000       & 1.000 \\ \bottomrule
+  \begin{tabular}{cllr} \toprule
+       & Parameter                 & Description                       & Value \\ \midrule
+    1  & $v_{\text{txn,{\bf p}}}$  & {\it Per} Transcription rate      & 0.195 \\
+    2  & $v_{\text{txn,{\bf c1}}}$ & {\it Cry1} Transcription rate     & 0.131 \\
+    3  & $v_{\text{txn,{\bf c2}}}$ & {\it Cry1} Transcription rate     & 0.114 \\
+    4  & $k_{\text{txn,{\bf p}}}$  & {\it Per} Repression constant     & 0.425 \\
+    5  & $k_{\text{txn,{\bf c}}}$  & {\it Cry1/2} Repression constant  & 0.259 \\
+    6  & $v_{\text{deg,{\bf p}}}$  & {\it Per} Max degradation rate    & 0.326 \\
+    7  & $v_{\text{deg,{\bf c1}}}$ & {\it Cry1} Max degradation rate   & 0.676 \\
+    8  & $v_{\text{deg,{\bf c2}}}$ & {\it Cry2} Max degradation rate   & 0.608 \\
+    9  & $k_{\text{deg,{\bf p}}}$  & {\it Per} Degradation constant    & 0.011 \\
+    10 & $k_{\text{deg,{\bf c}}}$  & {\it Cry1/2} Degradation constant & 1.149 \\
+    11 & $v_{\text{deg,{\bf P}}}$  & Max PERc degradation rate         & 2.970 \\
+    12 & $k_{\text{deg,{\bf P}}}$  & PERc degradation constant         & 0.034 \\
+    13 & $v_{\text{deg,{\bf C1}}}$ & Max CRY1c degradation rate        & 1.523 \\
+    14 & $v_{\text{deg,{\bf C2}}}$ & Max CRY2c degradation rate        & 1.686 \\
+    15 & $k_{\text{deg,{\bf C}}}$  & CRYc degradation constant         & 2.017 \\
+    16 & $v_{\text{deg,{\bf CP}}}$ & CRYn degradation rate             & 0.101 \\
+    17 & $m_{\text{\bf C2N}}$      & CRY2n degradation multiplier      & 3.318 \\
+    18 & $k_{\text{deg,{\bf CP}}}$ & CRYn degradation constant         & 0.053 \\
+    19 & $v_{\text{a,{\bf CP}}}$   & CRYn association rate             & 0.041 \\
+    20 & $v_{\text{d,{\bf CP}}}$   & CRYn dissociation rate            & 0.002 \\
+    21 & $k_{\text{tln,{\bf p}}}$  & PER translation rate              & 3.000 \\ \bottomrule
     \hline
@@
 \caption{{\bfseries Multivariable linear regression results.} Fit statistics which demonstrate the effect of each other fitted parameter on damping rate.
-Of particular note is the perturbation type categorical variable, which demonstrates that the presence of siRNA perturbation increases damping rate on average.
+Of particular note is the perturbation type categorical variable, which demonstrates that the presence of siRNA perturbation increases damping rate on average, controlling for changes in other variables.
 Higher amplitude is also correlated with higher damping rate.
-However, in total damping rate is poorly predicted by the other fitted variables $R^2 = 0.169$, indicating it describes an independent oscillatory feature.}
+However, in total damping rate is poorly predicted by the other fitted variables ($R^2 = 0.169$), indicating it describes an independent oscillatory feature.}
 \label{tab:ols_reg}

### 2015-03-02 main_text.tex
updates following the conference

@@
 % My preferred fonts
+\usepackage{lmodern}
 \usepackage[T1]{fontenc}
-\usepackage{lmodern}
 \usepackage[sc]{mathpazo}
+\usepackage{textcomp}
 
@@
 
-Circadian rhythms in mammals are daily changes in gene expression and physiology that persist even in the absence of external environmental cues \cite{Herzog2007}.
-Such rhythms are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations \cite{Ueda2005}.
+Circadian rhythms are daily changes in gene expression and physiology that persist even in the absence of external environmental cues \cite{Herzog2007}.
+In mammals, such rhythms are organized in a hierarchical fashion.
+At the tissue-level, the brain's suprachiasmatic nucleus (SCN) serves as the master pacemaker and keeps circadian oscillations in peripheral tissues in phase with the light-dark cycle.
+Cell-to-cell coupling in the SCN keeps individual cells in tight synchrony \cite{Herzog2004}. 
+Within each tissue, cellular-level rhythms are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations in gene transcription \cite{Ueda2005}.
+The robust oscillation of circadian factors has been linked to metabolic health \cite{Bass2012}, as compromised rhythms, whether due to gene knockout \cite{Marcheva2010} or irregular feeding schedules \cite{Hatori2012}, result in an increased risk of metabolic disease.
+Additionally, since the amplitude of circadian transcription can be affected by lifestyle variables such as diet, age, or work schedules, there has been recent interest in developing pharmacological strategies for increasing the amplitude of circadian cycles in metabolic tissues \cite{Chen2013}.
+
+An detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
 The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
@@
 Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
-This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014}, and has lead to the search for small-molecule therapeutics which boost clock amplitude \cite{Chen2013}.
+This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014}, and has aided the search for small-molecule therapeutics to boost clock amplitude \cite{Chen2013}.
 
+Bioluminescence rhythms at the cell-culture or tissue level are the result of the collective behavior of thousands of cells.
 Transcription at the single-cell level is strongly affected by the low molecular counts of the mRNA and protein species involved. 
-As a result, bioluminescence traces of individual cells are stochastic, with significant period-to-period variability \cite{Welsh2004}. 
-The collective behavior of thousands of cells results in more reliable oscillations at the tissue-level, especially in the suprachiasmatic nucleus (SCN) where cell-to-cell coupling keeps individual oscillators in tight synchrony \cite{Herzog2004}.
-While communication between cells is common in many tissues, it is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues or cultured reporter cells, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
-In populations which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level \cite{Welsh2004}. 
+As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length \cite{Welsh2004}. 
 Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
@@
 Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
-
 Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of fibroblast cells and subsequent image processing \cite{Leise2012}.
 As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.
-In this study, we demonstrate single-cell stochasticity can be reliably inferred from the damping rate of population-level bioluminescence recordings.
+
+It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
+Additionally, it has been shown that in cell cultures which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level \cite{Nagoshi2004, Welsh2004}.
+In this study, we demonstrate single-cell stochasticity can be reliably inferred from the damping rate of population-level bioluminescence recordings if the following conditions are met, which have been shown experimentally for many types of cultured cells \cite{Nagoshi2004, Welsh2004, Guenthner2014, Noguchi2013}.
+We assume oscillations in individual cells are
+\begin{itemize}
+  \item independent (no significant cell-to-cell coupling),
+  \item sustained (do not damp on a single-cell basis),
+  \item and homogeneous, with no differences in expected period length.
+\end{itemize}
+We demonstrate the validity and usefulness of such an approach on several types of circadian data.
 First, we show using cell-autonomous fibroblast data that stochastic noise is directly related to population-level damping.
@@
 
-In this study, we have described a method by which stochastic noise can be estimated from population-level bioluminescence recordings.
+In this study, we have described a method by which stochastic noise can be estimated from population-level bioluminescence recordings of cultured cells.
+Our method assumes independent and sustained oscillators at the single-cell level.
+While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured circadian reporters.
+Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
+However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.
+
 As high amplitude circadian oscillations are important for maintaining metabolic health, many studies have sought to find small molecule candidates which increase oscillatory amplitude \cite{Chen2013}.
@@
 \section*{Acknowledgments}
+We thank Tsuyoshi Hirota, Eric Zhang, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
 This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.

### 2015-03-03 main_text.tex
second pass to frank

@@
 
-An detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
+A detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
 The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
@@
 
-Bioluminescence rhythms at the cell-culture or tissue level are the result of the collective behavior of thousands of cells.
+Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells.
 Transcription at the single-cell level is strongly affected by the low molecular counts of the mRNA and protein species involved. 
 As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length \cite{Welsh2004}. 
+It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
+Additionally, it has been shown that in cell cultures which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.
+
 Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
@@
 Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
-Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of fibroblast cells and subsequent image processing \cite{Leise2012}.
+Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of individual cells and subsequent image processing \cite{Leise2012}.
 As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.
 
-It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
-Additionally, it has been shown that in cell cultures which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level \cite{Nagoshi2004, Welsh2004}.
-In this study, we demonstrate single-cell stochasticity can be reliably inferred from the damping rate of population-level bioluminescence recordings if the following conditions are met, which have been shown experimentally for many types of cultured cells \cite{Nagoshi2004, Welsh2004, Guenthner2014, Noguchi2013}.
-We assume oscillations in individual cells are
+In this study, we demonstrate that changes to single-cell stochasticity can be reliably inferred from the changes to the damping rate of population-level bioluminescence recordings of cultured circadian reporters.
+Our method assumes the following conditions are met, which have been shown to hold experimentally for many types of cultured cells \cite{Nagoshi2004, Welsh2004, Guenthner2014, Noguchi2013}:
+oscillations in individual cells are
 \begin{itemize}
@@
   \item sustained (do not damp on a single-cell basis),
-  \item and homogeneous, with no differences in expected period length.
+  \item and homogeneous, with no differences in expected period length between cells.
 \end{itemize}
 We demonstrate the validity and usefulness of such an approach on several types of circadian data.
-First, we show using cell-autonomous fibroblast data that stochastic noise is directly related to population-level damping.
+First, we show using single-cell fibroblast data that stochastic noise is directly related to population-level damping.
 Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to single-cell stochasticity is a likely mechanism.
@@
 
-In this study, we have described a method by which stochastic noise can be estimated from population-level bioluminescence recordings of cultured cells.
+In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
 Our method assumes independent and sustained oscillators at the single-cell level.
-While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured circadian reporters.
+While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured fibroblast cells.
 Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.

### 2015-03-09 main_text.tex
switching to plos citation style
renaming figures, updating manuscript links

@@
 
+\subsection*{Software}
+All computations were performed using Python.
+Codes used to produced the figures in this manuscript can be found at \url{https://github.com/pstjohn/decay-methods}.
+
 \section*{Results and Discussion}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/noise_ts_and_boot.pdf}
+    % \includegraphics[]{figures/pdfs/Fig1.pdf}
   \end{center}
@@
   \begin{center}
-    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/mainfig_dose.pdf}}
+    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig2.pdf}
   \end{center}
@@
   \begin{center}
-    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/main_fig_simulations.pdf}}
+    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig3.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/r2_vs_damping.pdf}
+    % \includegraphics[]{figures/pdfs/Fig4.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/fitted_parameters.pdf}
+    % \includegraphics[]{figures/pdfs/Fig5.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/outliers.pdf}
+    % \includegraphics[]{figures/pdfs/Fig6.pdf}
   \end{center}
@@
 
-\bibliographystyle{msb}
+\bibliographystyle{plos2015}
 \bibliography{condensed_library.bib}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/small_molecule_r2.pdf}
+    % \includegraphics[]{figures/pdfs/FigS1.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/volume_calibration.pdf}
+    % \includegraphics[]{figures/pdfs/FigS2.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/zhang_plates.pdf}
+    % \includegraphics[]{figures/pdfs/FigS3.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/zhang_wells.pdf}
+    % \includegraphics[]{figures/pdfs/FigS4.pdf}
   \end{center}

### 2015-03-11 main_text.tex
incorperating frank's edits

--- 2015-03-03 main_text.tex
@@
 \usepackage{booktabs}
+
+% Make subsections more clear
+\usepackage{titlesec}
+
+\titleformat{\subsubsection}
+  {\normalfont\itshape}{\thesubsubsection}{1em}{}
 
@@
 \begin{abstract}
-Stochastic noise at the cellular level has been shown to play a fundamental role in circadian oscillations, influencing how groups of cells entrain to external cues and perhaps serving as the mechanism by which cell-autonomous rhythms are generated.
+Stochastic noise at the cellular level has been shown to play a fundamental role in circadian oscillations, influencing how groups of cells entrain to external cues and likely serving as the mechanism by which cell-autonomous rhythms are generated.
 Despite this importance, few studies have investigated how clock perturbations affect stochastic noise - even as increasing numbers of high-throughput screens categorize how gene knockdowns or small molecules can change clock period and amplitude.
@@
 Finally, we hypothesize that the unperturbed clock may be highly optimized for robust rhythms, as very few gene perturbations are capable of simultaneously increasing amplitude and lowering stochastic noise.
-Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on cell-autonomous noise, especially with regard to the development of small-molecule circadian therapeutics.
+Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on cell-autonomous noise, particularly with regard to the development of small-molecule circadian therapeutics.
 \end{abstract}
@@
 \subsection*{Fitting a damped sinusoid to experimental data}
-Raw experimental data $x_i(t_i), \  i \in \{0, \dots, N-1\}$ are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{s}\right)^4$, in which $s$ is the sampling rate (in hours) \cite{Ravn2002}.
+Raw experimental data $x(t_i), \  i \in \{0, \dots, N-1\}$ are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{T_s}\right)^4$, in which $T_s$ is the sampling rate (in hours) \cite{Ravn2002}.
 The detrended data are then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1$).
-We denote the detrended and filtered experimental data by $y_i(t_i)$.
+We denote the detrended and filtered experimental data by $y(t_i)$.
 A damped sinusoid, specified by:
@@
 \[
-  R^2 = 1 - \frac{\sum_{i = 0}^{N-1} (y_i(t_i) - \hat{y}(t_i))^2}{\sum_{i = 0}^{N-1} (y_i(t_i) - \bar{y}(t_i))^2}
+  R^2 = 1 - \frac{\sum_{i = 0}^{N-1} (y(t_i) - \hat{y}(t_i))^2}{\sum_{i = 0}^{N-1} (y(t_i) - \bar{y}(t_i))^2}
 \]
+in which $\bar{y}(t_i)$ represents the mean of the detrended data.
 
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/noise_ts_and_boot.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/noise_ts_and_boot.pdf}
   \end{center}
@@
 
-To lend further support to the conclusion that the dose-dependent changes in damping rate from KL001 is due to changes in single-cell noise characteristics, we employed a model of circadian rhythms previously used to explain the effects of both small molecule perturbations (Tables~\ref{tab:model}-\ref{tab:parset}) \cite{St.John2014}.
+To lend further support to the conclusion that the dose-dependent changes in damping rate from KL001 is due to changes in single-cell noise characteristics, we employed a model of circadian rhythms previously used to explain the effects of both small molecule perturbations \cite{St.John2014}, summarized in Tables \ref{tab:model}-\ref{tab:parset}.
 In order to capture changes to noise characteristics, we first converted the model to a stochastic biochemical system.
@@
 As we show in Figure~\ref{fig:fit_distributions}, all fitted parameters displayed normal-like distributions, in which the control distributions showed much tighter clustering around the most likely values.
-Quantifications of the mean, variance, skew, and kurtosis for each distribution are shown in Table~\ref{tab:fit_distributions}.
+Quantifications of the mean, variance, skewness, and kurtosis for each distribution are shown in Table~\ref{tab:fit_distributions}.
 
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/fitted_parameters.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/fitted_parameters.pdf}
   \end{center}
@@
       {}         & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}               & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           \\\midrule
-      $\mu$      & $-0.234$  & $0.187$  &  $0.443$  & $-0.343$  &  $0.043$  & $0.090$    \\
-      $\sigma$   & $ 0.774$  & $1.820$  &  $0.778$  & $ 1.753$  &  $0.878$  & $1.688$    \\
-      Skew & $ 0.153$  & $0.367$  & -$1.823$  & $-0.580$  & -$0.107$  & $0.371$    \\
-      Kurt & $ 3.772$  & $0.591$  &  $8.329$  & $ 0.476$  &  $2.423$  & $0.373$    \\
+      Mean     & $-0.234$ & $0.187$ & $0.443$  & $-0.343$ & $0.043$  & $0.090$    \\
+      Variance & $ 0.599$ & $3.313$ & $0.605$  & $ 3.074$ & $0.771$  & $2.850$    \\
+      Skewness & $ 0.153$ & $0.367$ & $-1.823$ & $-0.580$ & $-0.107$ & $0.371$    \\
+      Kurtosis & $ 3.772$ & $0.591$ & $8.329$  & $ 0.476$ & $2.423$  & $0.373$    \\
       \bottomrule
@@
 
-\subsubsection*{The unperturbed clock may lie at the Pareto frontier of high amplitude and low noise}
+\subsubsection*{The unperturbed clock lies at the Pareto frontier of high amplitude and low noise}
 
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/outliers.pdf}
+    \includegraphics[width=0.65\textwidth]{figures/pdfs/outliers.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/small_molecule_r2.pdf}
+    \includegraphics[width=0.5\textwidth]{figures/pdfs/small_molecule_r2.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/volume_calibration.pdf}
+    \includegraphics[width=0.5\textwidth]{figures/pdfs/volume_calibration.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/zhang_plates.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/zhang_plates.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/zhang_wells.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/zhang_wells.pdf}
   \end{center}

### 2015-03-11 main_text.tex

@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/noise_ts_and_boot.pdf}
+    % \includegraphics[]{figures/pdfs/Fig1.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/fitted_parameters.pdf}
+    % \includegraphics[]{figures/pdfs/Fig5.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.65\textwidth]{figures/pdfs/outliers.pdf}
+    % \includegraphics[]{figures/pdfs/Fig6.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/pdfs/small_molecule_r2.pdf}
+    % \includegraphics[]{figures/pdfs/FigS1.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/pdfs/volume_calibration.pdf}
+    % \includegraphics[]{figures/pdfs/FigS2.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/zhang_plates.pdf}
+    % \includegraphics[]{figures/pdfs/FigS3.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/zhang_wells.pdf}
+    % \includegraphics[]{figures/pdfs/FigS4.pdf}
   \end{center}

### 2015-03-12 main_text.tex
another edit from frank
further updates, adding FigS6

--- 2015-03-11 main_text.tex
@@
 Despite this importance, few studies have investigated how clock perturbations affect stochastic noise - even as increasing numbers of high-throughput screens categorize how gene knockdowns or small molecules can change clock period and amplitude.
-This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which requires the careful collection and processing of single-cell data.
+This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which currently requires the careful collection and processing of single-cell data.
 % since cultured circadian reporters lack significant intercellular coupling,
@@
 Next, we show that the damping rate of cultured cells can be changed in a dose-dependent fashion by small molecule modulators, and confirm that such a change can be explained by single-cell noise using a mathematical model.
-We further demonstrate the insights that can be gained by applying such a method to a genome-wide siRNA screen, revealing that stochastic noise is altered independently from period, amplitude, and phase.
-Finally, we hypothesize that the unperturbed clock may be highly optimized for robust rhythms, as very few gene perturbations are capable of simultaneously increasing amplitude and lowering stochastic noise.
+We further demonstrate the insights that can be gained by applying our method to a genome-wide siRNA screen, revealing that stochastic noise is altered independently from period, amplitude, and phase.
+Finally, we hypothesize that the unperturbed clock is highly optimized for robust rhythms, as very few gene perturbations are capable of simultaneously increasing amplitude and lowering stochastic noise.
 Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on cell-autonomous noise, particularly with regard to the development of small-molecule circadian therapeutics.
@@
 Circadian rhythms are daily changes in gene expression and physiology that persist even in the absence of external environmental cues \cite{Herzog2007}.
-In mammals, such rhythms are organized in a hierarchical fashion.
-At the tissue-level, the brain's suprachiasmatic nucleus (SCN) serves as the master pacemaker and keeps circadian oscillations in peripheral tissues in phase with the light-dark cycle.
+In mammals, such rhythms are organized in a hierarchical fashion: at the tissue-level, the brain's suprachiasmatic nucleus (SCN) serves as the master pacemaker and keeps circadian oscillations in peripheral tissues in phase with the light-dark cycle.
 Cell-to-cell coupling in the SCN keeps individual cells in tight synchrony \cite{Herzog2004}. 
-Within each tissue, cellular-level rhythms are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations in gene transcription \cite{Ueda2005}.
-The robust oscillation of circadian factors has been linked to metabolic health \cite{Bass2012}, as compromised rhythms, whether due to gene knockout \cite{Marcheva2010} or irregular feeding schedules \cite{Hatori2012}, result in an increased risk of metabolic disease.
-Additionally, since the amplitude of circadian transcription can be affected by lifestyle variables such as diet, age, or work schedules, there has been recent interest in developing pharmacological strategies for increasing the amplitude of circadian cycles in metabolic tissues \cite{Chen2013}.
+Within each tissue, cellular-level rhythms in gene transcription are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations \cite{Ueda2005}.
+The robust oscillation of circadian factors has been linked to metabolic health \cite{Bass2012}, since rhythms compromised by gene knockout \cite{Marcheva2010} or irregular feeding schedules \cite{Hatori2012} result in an increased risk of metabolic disease.
+Additionally, as the amplitude of circadian transcription can be affected by lifestyle variables such as diet, age, or work schedules, there has been recent interest in developing pharmacological strategies for increasing the amplitude of circadian cycles in metabolic tissues \cite{Chen2013}.
 
 A detailed understanding of the underlying transcriptional mechanisms is essential for the development of circadian therapeutics to be successful. 
-The functional roles of different species in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
+The functional roles of different genes in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
 Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
 Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
-This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014}, and has aided the search for small-molecule therapeutics to boost clock amplitude \cite{Chen2013}.
+This additional parameter has proven useful in differentiating between perturbations with the same effect on period \cite{St.John2014} and has aided the search for small-molecule therapeutics to boost clock amplitude \cite{Chen2013}.
 
@@
 Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
-Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation and recording of individual cells and subsequent image processing \cite{Leise2012}.
+Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation, recording, and image processing of individual cells \cite{Leise2012}.
 As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.
@@
 In this study, we demonstrate that changes to single-cell stochasticity can be reliably inferred from the changes to the damping rate of population-level bioluminescence recordings of cultured circadian reporters.
-Our method assumes the following conditions are met, which have been shown to hold experimentally for many types of cultured cells \cite{Nagoshi2004, Welsh2004, Guenthner2014, Noguchi2013}:
-oscillations in individual cells are
+Our method assumes that oscillations in individual cells are both
 \begin{itemize}
-  \item independent (no significant cell-to-cell coupling),
-  \item sustained (do not damp on a single-cell basis),
-  \item and homogeneous, with no differences in expected period length between cells.
+  \item independent (no significant cell-to-cell coupling) \cite{Noguchi2013, Guenthner2014},
+  \item and sustained (do not damp on a single-cell basis) \cite{Welsh2004, Nagoshi2004},
 \end{itemize}
+which have been shown to hold experimentally for cultured fibroblast cells.
 We demonstrate the validity and usefulness of such an approach on several types of circadian data.
@@
 Cells were ranked according to this combined noise metric, and a high-noise group and low-noise group were constructed by taking the 39 highest-noise and lowest-noise cells, respectively.
-The raw bioluminescence profiles were not initially synchronized, so in order to simulate the gradual desynchronization of a group of oscillators the traces were offset to have the same starting phase.
-This was accomplished by starting each trace at the first phase zero-crossing, which were found using a Hilbert transform.
+The raw bioluminescence profiles were not initially synchronized, so the traces were offset to have the same starting phase in order to simulate the gradual desynchronization of a group of oscillators.
+This was accomplished by starting each trace at the first phase zero-crossing, found using a Hilbert transform.
 
@@
 Averaged traces for low and high-noise group displayed a damped sinusoidal rhythm.
-The first 4 days of rhythms ($s = 0.5,\ N = 192$) were fit using a damped sinusoid.
+The first 4 days of rhythms ($T_s = 0.5,\ N = 192$) were fit using a damped sinusoid.
 To ensure the difference in damping rate between groups was statistically significant, a bootstrap analysis was performed.
-In each of $10,000$ bootstrap trials, the cells were randomly assigned evenly to either the low-noise or the high-noise group (one cell was omitted randomly in each trial to ensure even group sizes).
+In each of $10,000$ bootstrap trials, cells were randomly assigned evenly to either the low-noise or the high-noise group (one cell was randomly omitted in each trial to ensure even group sizes).
 The absolute difference in damping rate between the two populations was recorded to yield a two-tailed test.
@@
 \subsection*{Quantifying dose-dependent effects of small molecule modulators}
-Bioluminescence traces ($s = 1.67,\ N=71$) from the application of increasing concentrations of small molecule were fit using the method described in a previous section.
-Because the small molecules were toxic to the cells at very high concentrations, experiments were removed from further analysis when the $R^2$ of the sinusoidal regression fell below $0.80$ (Figure~\ref{fig:small_molecule_r2}).
+Bioluminescence traces ($T_s = 1.67,\ N=71$) with increasing small molecule concentrations were fit with a damped sinusoid using the method described in a previous section.
+Because the small molecules were toxic at very high concentrations, experiments were removed from further analysis where $R^2 < 0.80$ (Figure~\ref{fig:small_molecule_r2}).
 
 \subsection*{{In silico} prediction of small molecule experiments}
-A previously published mathematical model of circadian rhythms was used to predict the effects on population damping rate from the dose-dependent small molecule experiments \cite{Hirota2012}.
+A previously published mathematical model of circadian rhythms \cite{Hirota2012} was used to predict the effects on population damping rate from the dose-dependent small molecule experiments.
 The parameters used to capture the effects of each small molecule were the same as described previously \cite{St.John2014}.
@@
 Population-level rhythms were found by taking the average of $1,000$ noninteracting oscillators, starting at identical initial conditions.
-The only parameter left unspecified by the deterministic model was the cell volume $\Omega$, which controlled the amount of noise in the system.
-For a given $\Omega$, the population-level damping rate was found by finding the phase diffusivity parameter which gave the closest fit to the simulated population-level rhythms, as described in \cite{St.John2014a}.
+The only parameter left unspecified by the deterministic model was the cell volume, $\Omega$, which controlled the amount of noise in the system.
+For each $\Omega$, a damped sinusoid was fit to the population-averaged state trajectory. 
 An $R^2$ value was calculated for each fit, taking into account all eight state variables.
@@
 We calculated an average experimental damping rate of $d = 0.0151$ from the $0 \mu M$ bioluminescence trajectories for both KL001 and longdaysin.
-{\itshape In silico} damping rates were calculated for logarithmically space values of $\Omega \in (100, 500)$.
+{\itshape In silico} damping rates were calculated for logarithmically spaced values of $\Omega \in (100, 500)$.
 Ten independent groups of $1,000$ oscillators were simulated for each $\Omega$, from which the means and standard errors were found.
@@
 A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight (Figure~\ref{fig:vol_calibration}).
-The best fit was found to be $\Omega = 226.3 \pm 9.0$.
+The best fit was found to be $\Omega = 226.3 \pm 9.0$, and was used for subsequent model simulations.
 
@@
 \subsection*{Fitting the genome-wide siRNA screen}
-We analyzed the data and annotations for the $111,743$ wells ($s=2,\ N=72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
+We analyzed the data and annotations for the $111,743$ wells ($T_s = 2,\ N = 72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
 Fits for which the $R^2 < 0.80$ were discarded.
-We analyze the natural logarithm of the amplitude, as these are more normally distributed and on a similar scale to the damping rate.
+The natural logarithm of the amplitude was used, since it more closely resembled a normal distribution and was on a similar scale to the damping rate.
 Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
@@
 A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
-Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), we assume the pooled covariance can be approximated by the covariance of the control sample alone.
+Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the effect of the perturbed sample on the pooled covariance was neglected.
 
@@
 All computations were performed using Python.
-Codes used to produced the figures in this manuscript can be found at \url{https://github.com/pstjohn/decay-methods}.
+Code used to perform the analysis and produce the figures in this manuscript can be found online at \url{https://github.com/pstjohn/decay-methods}.
 
@@
 
-We first sought to determine whether single-cell stochastic noise affects the population-level damping rate in cultured bioluminescence cells.
+We first sought to determine whether changes to single-cell stochastic noise affect population-level damping rate.
 To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
-We sorted the cells into two groups, a low-noise group and a high-noise group, based on the relative high-frequency noise, period variability, and amplitude variability present in each trace.
+Cells were sorted into two groups, a low-noise group and a high-noise group, based on the relative high-frequency noise, period variability, and amplitude variability present in each trace.
 Example rhythms from cells in both groups are shown in Figure~\ref{fig:fibroblast_noise}A.
-Because the cells were not synchronized at the start of the recording, we replicate this effect {\itshape in silico} by shifting each series in time to align their start phases.
-We next calculated expected population-level bioluminescence traces by averaging the cellular PER2::LUC signal in each group.
+Because the cells were not synchronized at the start of the recording, this effect is replicated {\itshape in silico} by shifting each series in time to align their start phases.
+Population-level bioluminescence traces were then found by averaging the cellular PER2::LUC signal in each group.
 Both populations displayed averaged rhythms which resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures.
@@
 By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence}).
-This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc}), indicating it is a fundamental property of the overall gene regulatory network.
+This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc} U2OS cells), indicating it is a fundamental property of the overall gene regulatory network.
 
@@
   \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} Distributions in robust z-scores closely resemble normal distributions.
-For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.
-}
+For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.}
 \label{fig:fit_distributions}
@@
 Additionally, a multivariate linear regression of damping rate as a function of period, amplitude, phase, and perturbation type (control or perturbed, a categorical variable) produced an $R^2$ value of only $0.169$ (Table~\ref{tab:ols_reg}).
+These results reinforce the notion that the population-level damping rate describes an independent feature of the underlying oscillator.
 
@@
       {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
-      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\\
+      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\
       $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
@@
 Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations ($3.3\%$) simultaneously increase amplitude and decrease stochastic noise.
-It therefore appears that the unperturbed clock optimizes some trade-off between oscillator amplitude and stochastic noise, such that it is not feasible to make oscillations more robust by knocking down one gene at a time.
+It therefore appears that the unperturbed clock optimizes some trade-off between high amplitude and low stochastic noise, such that it is unlikely the knockdown of a single gene will make oscillations more robust.
 
@@
 
+\subsection*{Effect of cell heterogeneity}
+
+Experimental work has shown that cell-autonomous fibroblast cells have a distribution of free-running periods \cite{Leise2012}.
+Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
+While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, in the preceding analyses we neglect the effect of period heterogeneity.
+To motivate this assumption, we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis (Figure~\ref{fig:cell2cell-variability}).
+Additionally, it is important to note that the dephasing kinetics of these two alternative processes are distinguishable from population-level data.
+In dephasing due solely to stochastic noise, the probability density of finding a cell with any given phase follows a standard convection-diffusion regime, resulting in exponentially damped rhythms \cite{St.John2014a}.
+In dephasing due solely to period heterogeneity, the variance of the phase probability density broadens in time proportionally to $t^2$, rather than $t$ in standard diffusion.
+This type of anomalous diffusion results in a different damping profile, as highlighted in Figure~\ref{fig:dephasing-rates}.
+Since the data used in this study is well-described by an exponentially damped sinusoid, and since a mathematical model assuming homogeneous mean periods accurately predicts the experimental damping rate changes, this assumption is likely justified.
+Additionally, since this study is mainly concerned with changes in the damping rate (rather than the absolute damping rate), it is unlikely that period heterogeneity plays a major role.
+
 \section*{Conclusion}
 
+% Our method assumes independent and sustained oscillators at the single-cell level.
+% While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured fibroblast cells.
+% Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
+% However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.
+
 In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
-Our method assumes independent and sustained oscillators at the single-cell level.
-While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured fibroblast cells.
-Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
-However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.
-
 As high amplitude circadian oscillations are important for maintaining metabolic health, many studies have sought to find small molecule candidates which increase oscillatory amplitude \cite{Chen2013}.
@@
 \label{fig:well_variation}
+\end{figure}
+
+\begin{figure}[tbp]
+  \begin{center}
+    \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS5.pdf}
+  \end{center}
+  \caption{{\bfseries Variability in period length from cycle to cycle is greater than from cell to cell.} Using single-cell fibroblast data from \cite{Leise2012}, a distribution of period lengths were calculated for each cell.
+Average period length for each cell is shown by a blue dot, with cells sorted from lowest to highest average period.
+Error bars extend from the $5^\text{th}$ to $95^\text{th}$ percentile period within each cell.
+The gray shaded region extends from the $5^\text{th}$ to $95^\text{th}$ percentile of the average period lengths for the entire population.
+$87\%$ of cells show greater variability from cycle-to-cycle than the overall period heterogeneity.}
+  \label{fig:cell2cell-variability}
+\end{figure}
+
+\begin{figure}[tbp]
+  \begin{center}
+    \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS6.pdf}
+  \end{center}
+  \caption{{\bfseries Period heterogeneity and phase diffusion display different damping profiles.} ({\itshape Left}) A simulated population of homogeneous, noisy oscillators dephases due to cycle-to-cycle variability.
+The envelope of the averaged expression is proportional to $t$.
+({\itshape Right}) A deterministic, heterogeneous population dephases due to different free-running periods.
+In this case, the population displays ballistic phase diffusion, in which envelope of the population-averaged expression changes proportionally to $t^2$.}
+  \label{fig:dephasing-rates}
 \end{figure}
@@
 \textbf{Prob(Omnibus):} &$  0.000 $& \textbf{  Jarque-Bera (JB):  } &$18459.719$ \\
-\textbf{Skew:}          &$  0.697 $& \textbf{  Prob(JB):          } &$     0.00$ \\
+\textbf{Skewness:}          &$  0.697 $& \textbf{  Prob(JB):          } &$     0.00$ \\
 \textbf{Kurtosis:}      &$  4.664 $& \textbf{  Cond. No.          } &$     8.34$ \\

### 2015-03-13 main_text.tex
removing figs
wrote cover, need to do author summary, move methods after results
adding author summary

--- 2015-03-12 main_text.tex
@@
 This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which currently requires the careful collection and processing of single-cell data.
-% since cultured circadian reporters lack significant intercellular coupling,
 In this study, we show that the damping rate of population-level bioluminescence recordings can serve as an accurate measure of cell-autonomous stochastic noise, and one which can be applied to future and existing high-throughput circadian screens.
@@
 \end{abstract}
+
+\section*{Author Summary}
+
+As most organisms exist in an environment which changes predictably with a 24-hour period, highly optimized genetic circuits turn on and off the production of key regulatory proteins to anticipate the day/night cycle.
+In humans, the demands of a modern society have required that we deviate from this evolutionarily prescribed sleep and feeding schedule, resulting in increased long-term risks of metabolic disease.
+There is therefore a desire to find pharmacological treatments which would restore the normal functioning of our circadian clock despite irregular behavioral schedules.
+One aspect of these treatments which is often overlooked in searching for candidate drugs is how these treatments might affect the accuracy of the circadian timing system.
+% Since each cell in the body keeps its own internal time, increasing the sloppiness of the clock machinery can result in tissues not being able to agree on what time to turn on protein production.
+Recording the time of each cell is possible but difficult; as a result single-cell approaches cannot be scaled up to high-throughput searches.
+In this paper, we show that it is possible to estimate how much the noise of a system has changed by looking only at the averaged protein production of an entire population of cells.
+Such an approach allows us to analyze prior data for high-throughput screens, and show that the unperturbed clock has been highly optimized to be both accurate and high amplitude.
 

### 2015-03-13 main_text.tex

@@
   \begin{center}
-    \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS5.pdf}
+    % \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS5.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS6.pdf}
+    % \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS6.pdf}
   \end{center}

### 2015-03-13 main_text.tex
reformatting for plos standards

@@
 
-\section*{Materials and Methods}
-
-\subsection*{Fitting a damped sinusoid to experimental data}
-Raw experimental data $x(t_i), \  i \in \{0, \dots, N-1\}$ are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{T_s}\right)^4$, in which $T_s$ is the sampling rate (in hours) \cite{Ravn2002}.
-The detrended data are then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1$).
-We denote the detrended and filtered experimental data by $y(t_i)$.
-A damped sinusoid, specified by:
-\[
-  \hat{y}(t) = A e^{-d t} \sin\left(\frac{2\pi t}{T} + \theta\right)
-\]
-is then fit to the filtered data.
-For numerical efficiency, the period, $T$, and damping rate, $d$, parameters are fit first using a matrix pencil method \cite{Hua1990}, reviewed in \cite{Zielinski2011}.
-Amplitude, $A$, and phase, $\theta$, parameters are subsequently fit using a linear least-squares regression.
-Overall $R^2$ values for the regression were calculated from the residual error between the detrended data and fitted sinusoid:
-\[
-  R^2 = 1 - \frac{\sum_{i = 0}^{N-1} (y(t_i) - \hat{y}(t_i))^2}{\sum_{i = 0}^{N-1} (y(t_i) - \bar{y}(t_i))^2}
-\]
-in which $\bar{y}(t_i)$ represents the mean of the detrended data.
-
-\subsection*{Processing single-cell bioluminescence data}
-
-Single-cell bioluminescence data for 79 cells was obtained from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
-As was done in the original study, a discrete wavelet transform (using PyWavelets, \url{http://www.pybytes.com/pywavelets}) was performed to detrend and remove noise.
-
-\subsubsection*{Sorting cells by noise level}
-As in the original study, various parameters describing the average noise level of each cell were collected.
-Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components.
-From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
-An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
-From these three noise variables, a combined noise metric was constructed by projecting the variables along their first principle component (using scikit-learn, \url{http://scikit-learn.org/}).
-Cells were ranked according to this combined noise metric, and a high-noise group and low-noise group were constructed by taking the 39 highest-noise and lowest-noise cells, respectively.
-The raw bioluminescence profiles were not initially synchronized, so the traces were offset to have the same starting phase in order to simulate the gradual desynchronization of a group of oscillators.
-This was accomplished by starting each trace at the first phase zero-crossing, found using a Hilbert transform.
-
-\subsubsection*{Bootstrap estimations of the damping rate difference}
-Averaged traces for low and high-noise group displayed a damped sinusoidal rhythm.
-The first 4 days of rhythms ($T_s = 0.5,\ N = 192$) were fit using a damped sinusoid.
-To ensure the difference in damping rate between groups was statistically significant, a bootstrap analysis was performed.
-In each of $10,000$ bootstrap trials, cells were randomly assigned evenly to either the low-noise or the high-noise group (one cell was randomly omitted in each trial to ensure even group sizes).
-The absolute difference in damping rate between the two populations was recorded to yield a two-tailed test.
-The observed test statistic, $|d_h - d_l| = 6.65\times10^{-3}$, was found to be significant at the $\alpha = 0.05$ confidence threshold ($p = 0.0264$).
-
-\subsection*{Quantifying dose-dependent effects of small molecule modulators}
-Bioluminescence traces ($T_s = 1.67,\ N=71$) with increasing small molecule concentrations were fit with a damped sinusoid using the method described in a previous section.
-Because the small molecules were toxic at very high concentrations, experiments were removed from further analysis where $R^2 < 0.80$ (Figure~\ref{fig:small_molecule_r2}).
-
-\subsection*{{In silico} prediction of small molecule experiments}
-A previously published mathematical model of circadian rhythms \cite{Hirota2012} was used to predict the effects on population damping rate from the dose-dependent small molecule experiments.
-The parameters used to capture the effects of each small molecule were the same as described previously \cite{St.John2014}.
-The model was converted to a stochastic biochemical system and subsequently simulated using StochKit2 \cite{Sanft2011a} (via GillesPy, \url{https://github.com/JohnAbel/gillespy}).
-Population-level rhythms were found by taking the average of $1,000$ noninteracting oscillators, starting at identical initial conditions.
-The only parameter left unspecified by the deterministic model was the cell volume, $\Omega$, which controlled the amount of noise in the system.
-For each $\Omega$, a damped sinusoid was fit to the population-averaged state trajectory. 
-An $R^2$ value was calculated for each fit, taking into account all eight state variables.
-
-\subsubsection*{Fitting the volume parameter}
-We calculated an average experimental damping rate of $d = 0.0151$ from the $0 \mu M$ bioluminescence trajectories for both KL001 and longdaysin.
-{\itshape In silico} damping rates were calculated for logarithmically spaced values of $\Omega \in (100, 500)$.
-Ten independent groups of $1,000$ oscillators were simulated for each $\Omega$, from which the means and standard errors were found.
-Simulations in which $R^2 < 0.90$ were removed from further analysis.
-A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight (Figure~\ref{fig:vol_calibration}).
-The best fit was found to be $\Omega = 226.3 \pm 9.0$, and was used for subsequent model simulations.
-
-\subsubsection*{Parameter knockdown experiments}
-We replicated the effects of the small molecules KL001 and longdaysin mathematically through the reductions of the $vdcn$ and $vac1p$ parameters, respectively (Tables~\ref{tab:model}-\ref{tab:parset}).
-Knockdown simulations were performed with $20$ values of each parameter, linearly spaced between $100\%$ and $15\%$ of their nominal value.
-Similar to the volume calibration simulations, $10$ independent populations of $1,000$ oscillators were simulated from an initially entrained state.
-Means and standard errors in period and damping rate were calculated from each population-averaged trajectory.
-Simulations in which $R^2 < 0.90$ were removed from further analysis.
-
-\subsection*{Fitting the genome-wide siRNA screen}
-We analyzed the data and annotations for the $111,743$ wells ($T_s = 2,\ N = 72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
-Fits for which the $R^2 < 0.80$ were discarded.
-The natural logarithm of the amplitude was used, since it more closely resembled a normal distribution and was on a similar scale to the damping rate.
-Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
-Parameters were therefore normalized on a plate-by-plate basis using a robust z-score \cite{Birmingham2009}:
-\[
-  z_{R,i} = \frac{p_i - M(p_i)}{M\left(\left|p_i - M(p_i)\right|\right)}, \qquad i \in \{0, \dots, {\cal P} - 1 \}
-\]
-where $M(\cdot)$ denotes the median of a vector, and $p_i$ contains all the points in one plate, and $\cal P$ is the number of plates.
-We removed outlier points prior to calculating the moments of the distributions, Pearson correlation coefficients, and performing the multivariable linear regression.
-Outliers were defined as points which contained a z-metric (in either period, phase, amplitude, or damping rate) with an absolute value greater than eight.
-We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
-
-\subsubsection*{Detecting outlier perturbations}
-We found the average response to each gene perturbation by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
-A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
-A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
-Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the effect of the perturbed sample on the pooled covariance was neglected.
-
-\subsection*{Software}
-All computations were performed using Python.
-Code used to perform the analysis and produce the figures in this manuscript can be found online at \url{https://github.com/pstjohn/decay-methods}.
-
-\section*{Results and Discussion}
+\section*{Results}
 
@@
 The actions of small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
-By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence}).
+By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence},~\ref{fig:small_molecule_r2}).
 This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc} U2OS cells), indicating it is a fundamental property of the overall gene regulatory network.
@@
 
-\section*{Conclusion}
+\section*{Discussion}
 
@@
 Finally, the ability to extract an additional biologically relevant parameter from existing datasets will likely prove useful for many studies, as it allows further differentiation between perturbations which might otherwise have identical effects on bioluminescence rhythms.
+
+\section*{Materials and Methods}
+
+\subsection*{Fitting a damped sinusoid to experimental data}
+Raw experimental data $x(t_i), \  i \in \{0, \dots, N-1\}$ are first detrended using Hodrick-Prescott filter with a smoothing parameter $\gamma = 0.05\left(\nicefrac{24 \text{ hrs}}{T_s}\right)^4$, in which $T_s$ is the sampling rate (in hours) \cite{Ravn2002}.
+The detrended data are then filtered using a low-pass filter to remove high-frequency noise (forward-backward Butterworth filter with $n = 5,\ w_c = 0.1$).
+We denote the detrended and filtered experimental data by $y(t_i)$.
+A damped sinusoid, specified by:
+\[
+  \hat{y}(t) = A e^{-d t} \sin\left(\frac{2\pi t}{T} + \theta\right)
+\]
+is then fit to the filtered data.
+For numerical efficiency, the period, $T$, and damping rate, $d$, parameters are fit first using a matrix pencil method \cite{Hua1990}, reviewed in \cite{Zielinski2011}.
+Amplitude, $A$, and phase, $\theta$, parameters are subsequently fit using a linear least-squares regression.
+Overall $R^2$ values for the regression were calculated from the residual error between the detrended data and fitted sinusoid:
+\[
+  R^2 = 1 - \frac{\sum_{i = 0}^{N-1} (y(t_i) - \hat{y}(t_i))^2}{\sum_{i = 0}^{N-1} (y(t_i) - \bar{y}(t_i))^2}
+\]
+in which $\bar{y}(t_i)$ represents the mean of the detrended data.
+
+\subsection*{Processing single-cell bioluminescence data}
+
+Single-cell bioluminescence data for 79 cells was obtained from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
+As was done in the original study, a discrete wavelet transform (using PyWavelets, \url{http://www.pybytes.com/pywavelets}) was performed to detrend and remove noise.
+
+\subsubsection*{Sorting cells by noise level}
+As in the original study, various parameters describing the average noise level of each cell were collected.
+Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components.
+From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
+An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
+From these three noise variables, a combined noise metric was constructed by projecting the variables along their first principle component (using scikit-learn, \url{http://scikit-learn.org/}).
+Cells were ranked according to this combined noise metric, and a high-noise group and low-noise group were constructed by taking the 39 highest-noise and lowest-noise cells, respectively.
+The raw bioluminescence profiles were not initially synchronized, so the traces were offset to have the same starting phase in order to simulate the gradual desynchronization of a group of oscillators.
+This was accomplished by starting each trace at the first phase zero-crossing, found using a Hilbert transform.
+
+\subsubsection*{Bootstrap estimations of the damping rate difference}
+Averaged traces for low and high-noise group displayed a damped sinusoidal rhythm.
+The first 4 days of rhythms ($T_s = 0.5,\ N = 192$) were fit using a damped sinusoid.
+To ensure the difference in damping rate between groups was statistically significant, a bootstrap analysis was performed.
+In each of $10,000$ bootstrap trials, cells were randomly assigned evenly to either the low-noise or the high-noise group (one cell was randomly omitted in each trial to ensure even group sizes).
+The absolute difference in damping rate between the two populations was recorded to yield a two-tailed test.
+The observed test statistic, $|d_h - d_l| = 6.65\times10^{-3}$, was found to be significant at the $\alpha = 0.05$ confidence threshold ($p = 0.0264$).
+
+\subsection*{Quantifying dose-dependent effects of small molecule modulators}
+Bioluminescence traces ($T_s = 1.67,\ N=71$) with increasing small molecule concentrations were fit with a damped sinusoid using the method described in a previous section.
+Because the small molecules were toxic at very high concentrations, experiments were removed from further analysis where $R^2 < 0.80$ (Figure~\ref{fig:small_molecule_r2}).
+
+\subsection*{{In silico} prediction of small molecule experiments}
+A previously published mathematical model of circadian rhythms \cite{Hirota2012} was used to predict the effects on population damping rate from the dose-dependent small molecule experiments.
+The parameters used to capture the effects of each small molecule were the same as described previously \cite{St.John2014}.
+The model was converted to a stochastic biochemical system and subsequently simulated using StochKit2 \cite{Sanft2011a} (via GillesPy, \url{https://github.com/JohnAbel/gillespy}).
+Population-level rhythms were found by taking the average of $1,000$ noninteracting oscillators, starting at identical initial conditions.
+The only parameter left unspecified by the deterministic model was the cell volume, $\Omega$, which controlled the amount of noise in the system.
+For each $\Omega$, a damped sinusoid was fit to the population-averaged state trajectory. 
+An $R^2$ value was calculated for each fit, taking into account all eight state variables.
+
+\subsubsection*{Fitting the volume parameter}
+We calculated an average experimental damping rate of $d = 0.0151$ from the $0 \mu M$ bioluminescence trajectories for both KL001 and longdaysin.
+{\itshape In silico} damping rates were calculated for logarithmically spaced values of $\Omega \in (100, 500)$.
+Ten independent groups of $1,000$ oscillators were simulated for each $\Omega$, from which the means and standard errors were found.
+Simulations in which $R^2 < 0.90$ were removed from further analysis.
+A weighted least-squares regression (using statsmodels, \url{http://statsmodels.sourceforge.net/}) was performed for $\log d$ vs.\ $\log \Omega$, using the $\log \text{SEM}$ of each measurement as a regression weight (Figure~\ref{fig:vol_calibration}).
+The best fit was found to be $\Omega = 226.3 \pm 9.0$, and was used for subsequent model simulations.
+
+\subsubsection*{Parameter knockdown experiments}
+We replicated the effects of the small molecules KL001 and longdaysin mathematically through the reductions of the $vdcn$ and $vac1p$ parameters, respectively (Tables~\ref{tab:model}-\ref{tab:parset}).
+Knockdown simulations were performed with $20$ values of each parameter, linearly spaced between $100\%$ and $15\%$ of their nominal value.
+Similar to the volume calibration simulations, $10$ independent populations of $1,000$ oscillators were simulated from an initially entrained state.
+Means and standard errors in period and damping rate were calculated from each population-averaged trajectory.
+Simulations in which $R^2 < 0.90$ were removed from further analysis.
+
+\subsection*{Fitting the genome-wide siRNA screen}
+We analyzed the data and annotations for the $111,743$ wells ($T_s = 2,\ N = 72$) in the Zhang {\itshape et al.,} 2009 screen \cite{Zhang2009}.
+Fits for which the $R^2 < 0.80$ were discarded.
+The natural logarithm of the amplitude was used, since it more closely resembled a normal distribution and was on a similar scale to the damping rate.
+Plate to plate variation, as shown in Figure~\ref{fig:plate_variation}, was more severe than variation on a well-to-well basis, Figure~\ref{fig:well_variation}.
+Parameters were therefore normalized on a plate-by-plate basis using a robust z-score \cite{Birmingham2009}:
+\[
+  z_{R,i} = \frac{p_i - M(p_i)}{M\left(\left|p_i - M(p_i)\right|\right)}, \qquad i \in \{0, \dots, {\cal P} - 1 \}
+\]
+where $M(\cdot)$ denotes the median of a vector, and $p_i$ contains all the points in one plate, and $\cal P$ is the number of plates.
+We removed outlier points prior to calculating the moments of the distributions, Pearson correlation coefficients, and performing the multivariable linear regression.
+Outliers were defined as points which contained a z-metric (in either period, phase, amplitude, or damping rate) with an absolute value greater than eight.
+We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
+
+\subsubsection*{Detecting outlier perturbations}
+We found the average response to each gene perturbation by grouping the perturbed dataset by target gene ID (using pandas, \url{http://pandas.pydata.org/}).
+A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
+A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
+Because the control distribution ($n=11,253$) is much larger than that for any particular gene ID ($n \approx 4$), the effect of the perturbed sample on the pooled covariance was neglected.
+
+\subsection*{Software}
+All computations were performed using Python.
+Code used to perform the analysis and produce the figures in this manuscript can be found online at \url{https://github.com/pstjohn/decay-methods}.
 

### 2015-03-16 main_text.tex
updating acknowledgement

--- 2015-03-13 main_text.tex
@@
 \section*{Acknowledgments}
-We thank Tsuyoshi Hirota, Eric Zhang, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
+We thank Tsuyoshi Hirota, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
 This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.

### 2015-03-17 main_text.tex
adding italics?

@@
 
-\subsection*{{In silico} prediction of small molecule experiments}
+\subsection*{{\itshape In silico} prediction of small molecule experiments}
 A previously published mathematical model of circadian rhythms \cite{Hirota2012} was used to predict the effects on population damping rate from the dose-dependent small molecule experiments.

### 2015-05-01 response_letter.tex
starting response letter

\documentclass[11pt, letterpaper]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

% Math, graphics, and bibliography
\usepackage{amsmath,amssymb,amsfonts,mathrsfs,mathtools}

% My preferred fonts
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[sc]{mathpazo}
\usepackage{textcomp}

\usepackage{color}
\usepackage{tcolorbox}
\tcbuselibrary{breakable}
\tcbuselibrary{skins}
\usepackage{units}

\usepackage[parfill]{parskip} % For block paragraphs

\definecolor{gray}{gray}{0.4}
\newenvironment{reviewer}{\itshape\color{gray}}{}

\tcbset{skin=enhanced}

\newenvironment{manuscript}[1]{\begin{center}\begin{tcolorbox}[colback=green!5!white,colframe=green!75!black,width=0.8\textwidth,title={#1},breakable,fonttitle=\bfseries]}{\end{tcolorbox}\end{center}}

\begin{document}

\begin{flushright}
\today\\[2ex]
{\itshape Francis J. Doyle III}\\
Dept.\ of Chemical Engineering\\
Univ.\ of California, Santa Barbara\\
Santa Barbara, CA 93106-5080\\
\end{flushright}

{\itshape Christopher V. Rao}\\
Associate Editor\\
PLOS Computational Biology\\

{\itshape Douglas Lauffenburger}\\
Associate Editor\\
PLOS Computational Biology\\

Thank you for your email on April 20th, 2015, inviting us to submit a revised version of our manuscript, ``Quantifying stochastic noise in cultured circadian reporter cells'' (MS: PCOMPBIOL-D-15-00431) by Peter C. St. John and Francis J. Doyle~III.

We thank the reviewers for their detailed comments of the manuscript and suggestions for how to improve the work.
We believe addressing these concerns and suggestions have greatly improved the manuscript.

Please see our detailed responses to the reviewers below:

\section*{Reviewer \#1}
\begin{reviewer}
The premise of the manuscript is quite interesting and pertinent to the study of the circadian clock and of biological oscillators more generally. 
Under the assumptions that the individual oscillators are self-autonomous with non-damping amplitudes and do not exert any influence on each other (which can be reasonable in certain types of dissociated cell cultures), the population-level damping rate is related to the level of stochastic noise in the individual oscillators. 
Dose-dependent effects of small molecule modulators on the noise in the clock can then be studied via the damping rate of the population, which is much easier than trying to record individual oscillators to measure such effects. 
The authors' further analysis leads them to the intriguing conjecture that the circadian clock has achieved an optimal balance between amplitude and stochastic noise.

Overall the approach is appealing and seems promising, with plausible future applications. However, some issues require further attention to ensure the method is effective and sound.

1) What exactly is meant in the manuscript by ``stochastic noise'' is not entirely clear. It would be helpful if the authors explained what is being covered under this term: Extrinsic noise? Intrinsic noise? Only noise generated within each cell by the molecular mechanism, or larger scale effects as well? What specifically are the sources of the noise under consideration, and what are the consequences for the individual oscillators' amplitude, phase, and period (and variability thereof)?
\end{reviewer}

We thank the reviewer for their overall positive reception of the manuscript. 
In regards to their first critique, we mainly are considering intrinsic noise generated by the molecular mechanism.
While sources of extrinsic noise are certainly present, experimental studies (i.e., Herzog {\itshape et. al., JBR} 2004) have largely concluded that period lengths of circadian oscillations are more variable at the single-cell level (i.e., peak-to-peak time) than at the population level (differences in mean period between cells).
We agree that additional discussion of this point in manuscript is necessary, and have therefore expanded the introduction to include an explicit description of the different sources and effects of stochastic noise:

\begin{manuscript}{Page 3}
 Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells.
 {\bfseries
At the single-cell level, transcription is strongly affected by cellular noise, comprised of both {\itshape intrinsic} and {\itshape extrinsic} sources.
Intrinsic noise is generated from the low molecular counts of the mRNA and protein species involved.
As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length (14). 
Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude.
For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle has been shown to be larger than the variability in mean period length between cells (2).
}
In cell cultures which lack cell-to-cell coupling, it has been shown that  stochastic noise at the single-cell level is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase (14, 15).  
\end{manuscript}

\begin{reviewer}
2) Damping of the population level oscillation is typically caused by two main factors: heterogeneity in period among the oscillators and stochastic variation in cycle length from cycle to cycle in each oscillator due to noise. 
The authors should address how these two factors contribute to the damping rate and how to distinguish them. 
For instance, what if a clock modulator is causing increased variability in period across the population, rather than increased stochastic noise? That is, the period of certain oscillators could be changed more than that of other oscillators (which could occur given that the system is highly nonlinear); you could even have the average period remain stable but some oscillators have their periods shortened by the treatment while others are lengthened. 
This would increase the damping rate, but not be related to a change in stochastic noise. 
A set of noise-free oscillators with heterogeneous periods will also damp at the population level, and damp faster if the variance in the period is increased.
\end{reviewer}

This is true, and it is likely not possible to effectively differentiate between the two sources - added supplemental figure.
\begin{itemize}
  \item We have demonstrated single-cell noise is {\itshape sufficient} to explain the observed changes in population damping. (sorting results, mathematical model predictions)
  \item We (and others previously) demonstrate single-cell variability is greater than cell heterogeneity.
\end{itemize}
Regardless, it is possible damping rates are changed through differences in extrinsic noise. The insight we generate is still valid, however, since the dynamic consequences (faster desynchronization of cells) are similar.

\begin{reviewer}
The authors do have a paragraph addressing the effect of heterogeneity, but they seem to assume the heterogeneity will remain unchanged by the clock modulators. However, looking at Table 1, this assumption doesn't hold. The variance is enormously increased by the perturbation, and one would expect this increased variability is likely occurring at the single cell level as well.
\end{reviewer}

\begin{itemize}
  \item Not necessarily. Table 1 shows that the variance in population-level periods is increased as a result of the {\itshape entire} siRNA screen.
  \item At the single-cell level, each cell in a particular well would experience a perturbation from the same siRNA, which we would expect to have the same effect across cells.
\end{itemize}
I'll perhaps re-word the figure caption to make this point more clear.

\begin{reviewer}
3) The claim that the stochastic noise can be altered independently from amplitude doesn't seem sufficiently well substantiated. As the authors point out, a major source of noise in single cell rhythms is the low molecular counts of clock mRNAs and proteins. Increasing the amplitude generally implies larger molecular counts and therefore reduced noise. Also, as shown in Figure 5, siRNA perturbation led to greatly increased variability in period and amplitude, which could be the main cause of the increased damping rate (rather than increased stochastic noise). That is, the effect of the perturbation could be to amplify the heterogeneity, rather than the individual noise levels. For instance, a change in a key rate parameter governing the nonlinear molecular clock mechanism might lead not only to a change in the mean period across the population, but also to a change in the spread of the period distribution. 
\end{reviewer}

Again, I'll point out each siRNA has a distinct effect on the clock. Taken in aggregate, these distinct effects result in increased variance at the well-level.
\begin{itemize}
  \item {\itshape Increasing the amplitude generally implies larger molecular counts and therefore reduced noise} - This is not necessarily the case. Noise can come from many locations within the gene regulatory network which are not represented by the reporter. In general, noise and amplitude are not correlated (maybe there's data for this?)
\end{itemize}

\begin{reviewer}
4) Using the PER2::LUC fibroblast recordings to test the ideas provides a clever test of the method. However, there are still some concerns about whether the method would work as proposed. For instance, the assumption is that the cells will have synchronized phases at the beginning of the experiment. For the test, the authors aligned the individual cell traces, but in an experiment the cells would presumably be synchronized using a medium change or some other synchronizing pulse. However, such perturbations also tend to transiently increase the amplitude of the individual cells. How would that effect be distinguished from the damping due to stochastic noise and phases drifting apart?
\end{reviewer}

The windowing function would likely not resemble a damped sinusoid. I'll refer to the previous work. For the siRNA screen, the synchronizing pulse (medium change) is consistent across all the wells, and would therefore likely not affect our overall analysis.

\begin{reviewer}
5) In the methods, what is meant by the (8hr,258hr) wavelet component? What type of wavelet transform/filter is being applied, and what do the 8hr and 258hr values refer to? Similarly, please clarify what the (1hr,8hr) wavelet component is.
\end{reviewer}

I'll go into more detail in the manuscript regarding the discrete wavelet transform.

\section*{Reviewer \#2}
\begin{reviewer}
I have enjoyed reading the paper by St.\~John and Doyle - its an innovative analysis of single cell circadian data and I think it will be of interest to a wide audience. 
In particular the finding that dephasing occurs mostly due to noise rather than period heterogeneity is fascinating. 
My major concern is the statement that ``Higher noise results in faster damping in population-level rhythms'' - though I agree that this follows from their data analysis, I am not so convinced that this can generally "serve as an accurate measure of cell-autonomous stochastic noise", as mentioned in the abstract and other parts of the paper. 
For example see Fig. 9 of the simulation study presented in Thomas, Philipp, Hannes Matuschek, and Ramon Grima. 
``Intrinsic noise analyzer: a software package for the exploration of stochastic biochemical kinetics using the system size expansion.'' PloS one 7.6 (2012): e38518.
Therein it is shown, using a model of a rudimental circadian oscillator involving strong negative feedback, that the population-level rhythm predicted using ensemble averaged stochastic simulations (of uncoupled cells) shows damped oscillations which the rate equations (deterministic and hence no noise) miss!
That is in this case noise induces synchronous oscillations at the population level, rather than damping them! This is, i believe, an example which runs counter to the author's claims about noise leading to damped population level oscillations, generally. 
I would suggest the authors to discuss this possibility in the Discussion, for completeness sake, and also as a cautionary note on the use of damping as a measure of single cell noise.
\end{reviewer}

Here I'll mainly point out that {\itshape damping} does not imply reduced amplitude. Damping is the rate at which the amplitude decays.
Certainly for a noise-induced oscillator, the presence of noise will result in higher amplitudes. Our comments apply mainly to changes in noise and the resulting changes in exponential amplitude decay.
I'll add a sentence or two explaining the difference between amplitude and damping more clearly.

\section*{Reviewer \#3}

\begin{reviewer}
These authors expand on previous work (in Biophys J 2014) where they showed that random phase diffusion in oscillators cause population level signals to dampen exponentially. They show that this property can be exploited to infer changes in noise levels in the circadian oscillators of single cells from population level measurements. This is a potentially very useful method, and I would be looking forward to try it on our own data. I have concerns as outlined below, but I want to stress that I am generally favorable to this work and I tried to keep the suggested revisions and additional computations requested reasonable, none of them should be too time-consuming in relation to the added value.
\end{reviewer}

great!

\begin{reviewer}
Main concerns:

M1. The author's comments notwithstanding, there are some concerns regarding the possible contribution of period heterogeneity to the quantified population damping. Specifically, the decays in Fig 3A actually looks somewhat like the decay in theoretical Figure S6, right panel, which was computed assuming period heterogeneity. Of course, the authors convincingly show that population period heterogeneity is less than peak-to-peak variation in percent. I still request a control experiment, in which simulated time-series of observed peak-to-peak variation AND observed population period heterogeneity are analyzed with the authors' algorithm. The method should be shown to work as for instance noise levels are increased or decreased also for this case. This could be a complement to Figure 1C, where the same thing was done for oscillators with identical period.
\end{reviewer}

Similar to Reviewer \#1's concerns. This is largely an identifiability issue, as the particular sources of noise cannot be confidently identified from the population-level data (only overall noise).
We have added the control experiment demonstrating this.

\begin{reviewer}
M2. The method is of potential use for many labs including my own. What I was missing, however, are practical guidelines: How many samples are needed, which sampling rates, etc. are required for the method to work and yield ? How does it scale with increasing or decreasing sampling rates? Importantly, how do I decide whether a possible change in damping and thus noise level is significant, can one bootstrap this? Should one always use the fitting procedure with matrix pencil method, Hilbert transform to align phases, etc.? Here, a few trials on simulated time-series may suffice to provide such information.
\end{reviewer}

I'll add a paragraph on practical considerations for implementing the algorithm. Sampling rate is less important than overall recording duration. There are many ways to do the nonlinear regression and analysis, which likely depends on the data being analyzed. For looking specifically at the damping rate, the best method is likely detrending, followed by a semilog plot of window vs.\ time, from which more traditional statistical techniques on comparing two regressions (ancova, bayesian inference)

\begin{reviewer}
Minor concerns:

Relating to question M1, the authors refer to their previous work, ref. 23, where it was shown that phase drift leads to an exponentially decreasing population signal. Otoh, period heterogeneity would lead to a time-squared exponent; where is the latter formula derived? If this was done by Rougemont and Naef, they should be cited.
\end{reviewer}

An additional reference for Rougemont and Naef can be added, as they do discuss a variant of this. However, the idea that ballistic particles diffuse according to $t^2$ (and brownian particles according to $t$) is very old.

\begin{reviewer}
On p. 3, comments regarding reference 17: these authors rather submitted that either damped or self-sustained oscillators describe the data equally well, which actually more justifies the present method, which assumes self-sustained oscillations.
\end{reviewer}

This point will be clarified (our main idea is that noise is important), although our method does not assume self-sustained oscillations in the deterministic limit.

\begin{reviewer}
The method concerns primarily detection of noise levels, in the introduction it is emphasized that an amplitude boosting may ultimately be desirable in e.g. a clinical setting. Introduction would be strengthened by also including more use cases for the detection of noise strength.
\end{reviewer}

Will be added.

\begin{reviewer}
For the discussion: authors argue that the amplitude/noise ratio seem to have evolved to something close to an optimum. Yet, in the paper by Liu et al. in Cell 129, 2007, Cry2 knockouts seemed to have even stronger rhythms?
\end{reviewer}

We don't have access to the raw data of Liu et. al.,, but higher amplitude (as suggested by Liu) is not the same as stronger rhythms. I will check our database (many Cry2 siRNA replicates) to see if this trend is replicated in the siRNA screen.

\section*{Reviewer \#4}

\begin{reviewer}
Circadian rhythms play important roles in maintaining biological homeostasis. Recent research suggests that environmental conditions, like night work or an unusual feeding schedule, may have important implications for developing metabolic disease. Therefore it is of interest to identify therapeutic means to manipulate the circadian rhythms. Here, St. John and Doyle explore the relationship between stochastic noise and circadian oscillations, as typically observed at the cell population level and suggest a method to infer stochastic noise properties from population behavior. Overall, the work seems too tightly focused on a narrow scientific niche and the method for deducing noise properties is not convincing. The major criticisms are as follows:

1. The relationship between stochastic noise or cell-to-cell variability and how that impacts a population response is not particularly new. The observation that cell to cell variability gives rise to a dampened response at the cell population level has been observed before for other signaling systems, such as NF-kappaB (e.g., Nelson et al. Science (2004) 306:704-708). It is then unclear as to what really is new here.
\end{reviewer}

What is new is that this phenomenon can be quantified to give additional information on stochastic noise. From the datasets, this is an independent quantity (from period, amplitude), and therefore likely provides new insight into the underlying oscillator.

Also, we're attempting to downplaying the effect of cell-to-cell variability.

\begin{reviewer}
2. The authors use a leap of logic to justify their approach in extracting stochastic noise parameters from cell population data. First the authors show that by fitting a collection of models to single cell data and then averaging these mathematical models together, they can predict that the cell population response will become dampened when there is significant variability in the single cell responses (inductive reasoning?). Based on this demonstration, they then develop an approach that does the opposite: from population data, they try to deduce the values for the stochastic noise properties (deductive reasoning?). The problem seems to be that there could be other explanations for why miRNAs within a library affect a signaling network, such as miRNAs limit the efficiency of the biochemical signaling network that underpins the oscillatory behavior. Overall, the approach is not convincing in providing the claimed insight. 
\end{reviewer}

We are demonstrating that single-cell noise is {\itshape sufficient} to explain the observed damping behavior, and indeed quantitatively accurate predictions for population-level damping rate are obtained from a simple model not fit to the noise properties.

Indeed, it would be great if single-cell data was available for each siRNA perturbation. This is likely not possible, as single-cell data is difficult to collect.

Cells are likely not arrhythmic as a result of siRNA perturbation, as the siRNA method has been well-adopted by the circadian community for effects on amplitude and period.

\begin{reviewer}
Minor comments:
1. ``which'' is used incorrectly pretty much throughout the manuscript. should be ``that''
\end{reviewer}

This has been corrected.

We would like to once again thank all our reviewers, as we feel our manuscript has benefited greatly from their suggestions and critiques.

\end{document}

### 2015-05-01 main_text.tex
starting response letter

--- 2015-03-17 main_text.tex
@@
 In this paper, we show that it is possible to estimate how much the noise of a system has changed by looking only at the averaged protein production of an entire population of cells.
-Such an approach allows us to analyze prior data for high-throughput screens, and show that the unperturbed clock has been highly optimized to be both accurate and high amplitude.
+Such an approach allows us to analyze prior data from high-throughput screens, and show that the unperturbed clock has been highly optimized to be both accurate and high amplitude.
 
@@
 Cell-to-cell coupling in the SCN keeps individual cells in tight synchrony \cite{Herzog2004}. 
+It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
 Within each tissue, cellular-level rhythms in gene transcription are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations \cite{Ueda2005}.
@@
 Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells.
-Transcription at the single-cell level is strongly affected by the low molecular counts of the mRNA and protein species involved. 
+At the single-cell level, transcription is strongly affected by cellular noise, comprised of both {\itshape intrinsic} and {\itshape extrinsic} sources.
+Intrinsic noise is generated from the low molecular counts of the mRNA and protein species involved.
 As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length \cite{Welsh2004}. 
-It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
-Additionally, it has been shown that in cell cultures which lack cell-to-cell coupling, stochastic noise at the single-cell level is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.
+Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude.
+For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle has been shown to be larger than the variability in mean period length between cells \cite{Herzog2004}.
+In cell cultures which lack cell-to-cell coupling, it has been shown that  stochastic noise at the single-cell level is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.
 
@@
 
-We first sought to determine whether changes to single-cell stochastic noise affect population-level damping rate.
+We first sought to determine whether changes to intrinsic stochastic noise are sufficient to explain population-level changes in damping rate.
 To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
@@
 
-% \subsubsection*{A small molecule causes dose-dependent changes in the population-level damping rate of cultured cells}
+\subsubsection*{A small molecule causes dose-dependent changes in the population-level damping rate of cultured cells}
 
-We next demonstrate that perturbations to the transcriptional oscillator are capable of altering single-cell level stochastic noise.
+We next demonstrate that perturbations to the transcriptional oscillator are capable of altering population-level damping rate.
 The actions of small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
@@
 
-% \subsection*{Mathematical model predicts KL001 damping rate increase comes from increased single-cell stochastic noise}
+\subsection*{Mathematical model predicts KL001 damping rate increase comes from increased single-cell stochastic noise}
 
-To lend further support to the conclusion that the dose-dependent changes in damping rate from KL001 is due to changes in single-cell noise characteristics, we employed a model of circadian rhythms previously used to explain the effects of both small molecule perturbations \cite{St.John2014}, summarized in Tables \ref{tab:model}-\ref{tab:parset}.
+We next test the hypothesis that the dose-dependent changes in damping rate from KL001 is due to changes in intrinsic (cell-autonomous) noise characteristics.
+To do this, we employed a mathematical model of circadian rhythms previously used to explain the effects of both small molecule perturbations \cite{St.John2014}, summarized in Tables \ref{tab:model}-\ref{tab:parset}.
 In order to capture changes to noise characteristics, we first converted the model to a stochastic biochemical system.
@@
 Reductions in parameters previously attributed to the activities of each small molecule caused dose-dependent changes in period and damping rate at the population-level that closely matched experimental results (Figure~\ref{fig:simulation}B).
-As the model includes no cell-to-cell communication, this difference is manifested solely by changing the noise characteristics of individual cells.
+As the model includes no cell-to-cell communication or heterogeneity, this difference is manifested solely by changing the noise characteristics of individual cells.
+This computational result
 
@@
 
-Experimental work has shown that cell-autonomous fibroblast cells have a distribution of free-running periods \cite{Leise2012}.
+In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping.
+However, experimental work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods \cite{Leise2012}.
 Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
-While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, in the preceding analyses we neglect the effect of period heterogeneity.
-To motivate this assumption, we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis (Figure~\ref{fig:cell2cell-variability}).
-Additionally, it is important to note that the dephasing kinetics of these two alternative processes are distinguishable from population-level data.
+While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, experimental evidence suggests intrinsic stochastic noise plays a more significant role.
+We show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}).
+This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise \cite{Herzog2004}.
+
+However, it is possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
+Such a change, caused perhaps by an unequal uptake of drug across cells, would result in an altered distribution in mean period length, and different dephasing kinetics.
+
+% Additionally, it is important to note that the dephasing kinetics of these two alternative processes result in different dynamics at the population-level.
 In dephasing due solely to stochastic noise, the probability density of finding a cell with any given phase follows a standard convection-diffusion regime, resulting in exponentially damped rhythms \cite{St.John2014a}.
@@
 This type of anomalous diffusion results in a different damping profile, as highlighted in Figure~\ref{fig:dephasing-rates}.
-Since the data used in this study is well-described by an exponentially damped sinusoid, and since a mathematical model assuming homogeneous mean periods accurately predicts the experimental damping rate changes, this assumption is likely justified.
-Additionally, since this study is mainly concerned with changes in the damping rate (rather than the absolute damping rate), it is unlikely that period heterogeneity plays a major role.
+Since the data used in this study is well-described by an exponentially damped sinusoid, and since a mathematical model assuming homogeneous mean periods accurately predicts experimental results, we expect our assumption that changes to intrinsic noise are primarily responsible for the observed damping rate changes to be justified.
+
+However, it is possible that damping rate changes due to siRNA or small molecule perturbation are manifested by altering the system's extrinsic noise.
+Such a change, caused perhaps by an unequal uptake of drug across cells, would result in an altered distribution in mean period length, and different dephasing kinetics.
+A simple model of a stochastic oscillator was used to investigate the effects of altering intrinsic and extrinsic noise on the dephasing kinetics (Figure~\ref{fig:computational_dephasing}).
+While {\itshape in silico} dephasing due to solely intrinsic or extrinsic sources can be differentiated from population-level rhythms, it becomes more difficult to identify noise sources when multiple types of noise are present.
+The window function of population-level oscillations can be affected by many factors, including weak cell-to-cell coupling \cite{Rougemont2007}, temporary amplitude change \cite{St.John2014a}, and intrinsic and extrinsic noise.
+While altering intrinsic, or cell-autonomous, noise has been shown to be sufficient to explain the observed damping rate changes, it is possible that a portion of the changes to damping rate observed in regressed experimental data may be due to changes in extrinsic noise.
 
@@
   \label{fig:dephasing-rates}
+\end{figure}
+
+\begin{figure}[tbp]
+  \begin{center}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS7.pdf}
+  \end{center}
+  \caption{{\bfseries Intrinsic and extrinsic noise can be difficult to differentiate in .} ({\bfseries A}) }
+  \label{fig:computational_dephasing}
 \end{figure}

### 2015-05-06 response_letter.tex
First draft of edits

@@
 
-\newenvironment{manuscript}[1]{\begin{center}\begin{tcolorbox}[colback=green!5!white,colframe=green!75!black,width=0.8\textwidth,title={#1},breakable,fonttitle=\bfseries]}{\end{tcolorbox}\end{center}}
+\newenvironment{manuscript}[1]{\begin{center}\begin{tcolorbox}[colback=green!5!white,colframe=green!75!black,width=\textwidth,title={#1},breakable,fonttitle=\bfseries]}{\end{tcolorbox}\end{center}}
+% For hyperref links
+
+\usepackage[linktoc=all]{hyperref}
+\hypersetup{
+    colorlinks,
+    citecolor=black,
+    filecolor=black,
+    linkcolor=black,
+    urlcolor=black
+}
+
+\usepackage[margin=1cm]{caption}
+\renewcommand{\thefigure}{R\arabic{figure}}%
 
@@
 In regards to their first critique, we mainly are considering intrinsic noise generated by the molecular mechanism.
-While sources of extrinsic noise are certainly present, experimental studies (i.e., Herzog {\itshape et.
-al., JBR} 2004) have largely concluded that period lengths of circadian oscillations are more variable at the single-cell level (i.e., peak-to-peak time) than at the population level (differences in mean period between cells).
+In the study, we demonstrate that single-cell noise is {\itshape sufficient} to explain the observed changes in population damping.
+While sources of extrinsic noise are certainly present, experimental studies (i.e., Herzog {\itshape et al., JBR} 2004) have largely concluded that period lengths of circadian oscillations are more variable at the single-cell level (i.e., peak-to-peak time) than at the population level (differences in mean period between cells).
 We agree that additional discussion of this point in manuscript is necessary, and have therefore expanded the introduction to include an explicit description of the different sources and effects of stochastic noise:
@@
 {\bfseries
-At the single-cell level, transcription is strongly affected by cellular noise, comprised of both {\itshape intrinsic} and {\itshape extrinsic} sources.
-Intrinsic noise is generated from the low molecular counts of the mRNA and protein species involved.
-As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length (14). 
+  Transcription at the single-cell level is strongly affected by {\itshape intrinsic} cellular noise, caused by the low molecular counts of the mRNA and protein species involved.
+As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length from cycle to cycle (14).
+In addition to intrinsic noise, circadian oscillations are also affected by {\itshape extrinsic} noise sources.
 Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude.
-For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle has been shown to be larger than the variability in mean period length between cells (2).
+For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells (2).
+Both sources of noise have an effect on population-level rhythms: in cell cultures that lack cell-to-cell coupling, it has been shown that stochastic noise is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase (14, 15).
 }
-In cell cultures which lack cell-to-cell coupling, it has been shown that  stochastic noise at the single-cell level is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase (14, 15).
 \end{manuscript}
+
+Additionally, {\bfseries we have updated the language throughout the manuscript} to use the intrinsic and extrinsic vocabulary to refer to noise generated within the cell or outside the cell, respectively.
 
@@
 
-This is true, and it is likely not possible to effectively differentiate between the two sources - added supplemental figure.
-\begin{itemize}
-  \item We have demonstrated single-cell noise is {\itshape sufficient} to explain the observed changes in population damping. (sorting results, mathematical model predictions)
-  \item We (and others previously) demonstrate single-cell variability is greater than cell heterogeneity.
-\end{itemize}
-Regardless, it is possible damping rates are changed through differences in extrinsic noise.
-The insight we generate is still valid, however, since the dynamic consequences (faster desynchronization of cells) are similar.
+This is an valid concern, and one which was raised by multiple reviewers.
+In the original manuscript we mentioned that intrinsic (cycle-to-cycle variability) and extrinsic (period heterogeneity) sources could be differentiated based on their envelope functions.
+In response to the reviews, we have performed additional simulations to determine whether intrinsic or extrinsic noise sources are identifiable, presented in Figure S7.
+We conclude that while pure intrinsic vs.\ pure extrinsic dephasing can be distinguished, it becomes much more difficult when both sources of noise are present.
+
+While this revision does correct a statement in our previous version, we note that main parameter of biological interest is the overall stochastic noise from both sources, which is still being identified by the method.
+We have therefore updated the corresponding section of our manuscript to discuss the possibility and implications of changes to extrinsic noise:
+
+\begin{manuscript}{Page 13}
+  \section*{Effect of cell heterogeneity}
+  {\bfseries
+In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping. 
+However, experimental work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods (21). 
+Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping (30). 
+While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise likely plays a more significant role: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure S5).
+This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise (2).
+
+However, it is possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise. 
+Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics. 
+While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure S6), these differences are likely not identifiable from typical data (Figure S7).
+Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors. 
+Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.
+}
+\end{manuscript}
 
@@
 
-\begin{itemize}
-  \item Not necessarily. Table 1 shows that the variance in population-level periods is increased as a result of the {\itshape entire} siRNA screen.
-  \item At the single-cell level, each cell in a particular well would experience a perturbation from the same siRNA, which we would expect to have the same effect across cells.
-\end{itemize}
-I'll perhaps re-word the figure caption to make this point more clear.
+Indeed, the possibility that period heterogeneity may be changed by clock modulators has been added as discussed above.
+The increase in variance in Table 1, however, is most likely due to the variety of genes targeted by the siRNA library.
+At the single-cell level, each cell in the well would experience the same siRNA perturbation, which we suspect would lead to a similar change in period / amplitude.
+This is echoed by the consistency seen in the control wells.
+As this was perhaps a misunderstanding of the figure, we have added the following text to the caption of Figure 5:
+
+\begin{manuscript}{Page 5}
+  Figure 5: Distributions in fitted parameters for the genome-wide siRNA screen.
+{\bfseries For each well in the high-throughput screen, the period, amplitude and damping rate are calculated.
+After normalization,} distributions in robust z-scores closely resemble normal distributions.
+For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.
+\end{manuscript}
 
@@
 
-Again, I'll point out each siRNA has a distinct effect on the clock. Taken in aggregate, these distinct effects result in increased variance at the well-level.
-\begin{itemize}
-  \item {\itshape Increasing the amplitude generally implies larger molecular counts and therefore reduced noise} - This is not necessarily the case.
-Noise can come from many locations within the gene regulatory network which are not represented by the reporter.
-In general, noise and amplitude are not correlated (maybe there's data for this?)
-\end{itemize}
+In general, we agree with the reviewer on this point and have updated our language accordingly (see above).
+
+In regards to the reviewer's first comment, that amplitude and noise cannot be independently altered, we note that circadian oscillations are an emergent property from a very large transcriptional network.
+As such, the species present at low quantities which dominate the noise generation are not necessarily the same as those which are tied to the bioluminescence reporter.
+A reduction in effective concentration of a repressor, for instance, might lead to higher-amplitude transcription which is more susceptible to intrinsic noise.
+In addition to the linear independence we show between amplitude and damping rate, previous authors have also shown this independence: {\itshape ``The amplitude of rhythmicity also did not correlate with cycle-to-cycle variation of locomotor ($p = 0.17$), explant ($p = 0.24$), or cellular rhythms ($p = 0.15$)''} (Herzog {\itshape et al., JBR} 2004).
+
+Additionally, we caution that population-level distributions, as a result of perturbing each gene in the siRNA library, are likely not representative of the distributions we would expect to see at the single-cell level from a single siRNA.
 
@@
 
-The windowing function would likely not resemble a damped sinusoid.
-I'll refer to the previous work.
-For the siRNA screen, the synchronizing pulse (medium change) is consistent across all the wells, and would therefore likely not affect our overall analysis.
+This was actually the subject of a previous manuscript, St.\ John {\itshape et al., Biophys J}, 2014, which shows that the temporary amplitude change would largely be distinguished by a deviation from the exponential damping.
+This underscores the importance of collecting several days of data to accurately judge the damping rate.
+Additionally, for the siRNA screen, the synchronizing pulse (medium change) is consistent across all the wells.
+Since our method mainly analyses {\itshape differences} in damping rate, we suspect these effects would be minimal.
 
@@
 
-I'll go into more detail in the manuscript regarding the discrete wavelet transform.
+We used a discrete wavelet transform, the exact details of which are described in the original article (Leise et al., 2012 (18)).
+Since frequency bins are halved at each interval, the (8hr, 256hr) component represents the sum of the reconstructed 8-16, 16-32, 32-64, 64-128, 128-256 hr components.
+
+We have added additional text to the methods section detailing the wavelet transform we used
+
+\begin{manuscript}{Pages 4-5}
+  As was done in the original study, a discrete wavelet transform (using PyWavelets, \url{http://www.pybytes.com/pywavelets}) was performed to detrend and remove noise.
+{\bfseries A discrete wavelet transform decomposes the signal into multiple frequency bands (25).
+By only considering frequency bands close to the circadian frequency, high-frequency noise and low-frequency baseline oscillations can be removed.}
+\end{manuscript}
+
+\begin{manuscript}{Page 5}
+As in the original study, various parameters describing the average noise level of each cell were collected.
+{\bfseries
+Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components -- resulting in rhythms which only contained oscillations with periods between 8 and 258 hours.
+}
+From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
+An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
+{\bfseries
+This wavelet component was chosen as it contained only the highest-frequency noise in the signal's spectrum.
+}
+\end{manuscript}
 
@@
 \begin{reviewer}
-I have enjoyed reading the paper by St.\~John and Doyle - its an innovative analysis of single cell circadian data and I think it will be of interest to a wide audience. 
+I have enjoyed reading the paper by St.\ John and Doyle - its an innovative analysis of single cell circadian data and I think it will be of interest to a wide audience. 
 In particular the finding that dephasing occurs mostly due to noise rather than period heterogeneity is fascinating. 
@@
 
-Here I'll mainly point out that {\itshape damping} does not imply reduced amplitude. Damping is the rate at which the amplitude decays.
-Certainly for a noise-induced oscillator, the presence of noise will result in higher amplitudes.
-Our comments apply mainly to changes in noise and the resulting changes in exponential amplitude decay.
-I'll add a sentence or two explaining the difference between amplitude and damping more clearly.
+We thank the reviewer for their favorable impression of the manuscript.
+They raise the point that for some oscillators, the presence of noise is required for oscillations.
+This is very much in line with the theory of our manuscript, and is a point we raise in the introduction:
+{\itshape
+``Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, capture the noise characteristics seen in single-cell fibroblast data equally well as limit-cycle oscillators.''}
+
+We note that we do not intend {\itshape damping} to imply reduced amplitude.
+Instead, we use damping to indicate the rate at which the amplitude decays.
+Our analysis of the effect of noise on damping rate apply mainly to changes in noise and the resulting changes in exponential amplitude decay.
+We have added text to the manuscript to make our choice in vocabulary more clear:
+
+\begin{manuscript}{Page 4}
+ For numerical efficiency, the period, $T$, and damping rate, $d$, parameters are fit first using a matrix pencil method (24), reviewed in (25).
+Amplitude, $A$, and phase, $\theta$, parameters are subsequently fit using a linear least-squares regression.
+{\bfseries
+Note that in this manuscript we use amplitude to denote the initial rhythm strength, and damping rate to denote the rate at which this strength decays with time.}
+\end{manuscript}
 
@@
 
-great!
+We thank the reviewer for their positive critique of the study, and are excited that they would consider applying the method to their own work.
 
@@
 
-Similar to Reviewer \#1's concerns. This is largely an identifiability issue, as the particular sources of noise cannot be confidently identified from the population-level data (only overall noise).
-We have added the control experiment demonstrating this.
+These concerns were also shared by the first reviewer, and we have therefore substantially changed the language of the study to reflect the fact that extrinsic noise (period heterogeneity) could play a role in the damping rates.
+For Fig. 3A, these profiles were obtained from a mathematical model of homogeneous cells.
+The deviation from the exponential window is likely due to the limit cycle not being a perfect sinusoid.
+
+We thank the reviewer for suggesting the control experiment, which we performed almost exactly as described. The results of these computations have been added as Figure S7:
+
+\begin{manuscript}{Page 22}
+  \begin{center}
+  \includegraphics[width=0.51\textwidth]{figures/pdfs/FigS7.pdf}
+  \end{center}
+  {\bfseries Figure S7: Identifiability of intrinsic and extrinsic noise factors in population-level data.} A simple model with both intrinsic and extrinsic noise (Table~S3) was used to determine the identifiability of noise sources.
+Bioluminescence profiles were generated from a population of 1000 individual oscillators.
+The stochastic volume parameter, $\Omega$, controls the intrinsic noise, with larger volumes resulting in more deterministic profiles.
+Extrinsic noise was generated by using a normal distribution of free running periods, ${\cal N}(1, \sigma^2_T)$ days.
+Points represent window functions (amplitude over time) from 10 independent replications.
+Lines represent linear regressions, except in the right panel of part A, where a quadratic regression was used.
+  ({\bfseries A}) For models with pure intrinsic (left) or pure extrinsic (right) noise sources, amplitude damping proportional to $e^{dt}$ vs.\ $e^{dt^2}$, respectively, was readily distinguishable. 
+  ({\bfseries B}) For models with both intrinsic and extrinsic noise sources, it was not possible to differentiate between increasing intrinsic noise (left) and increasing extrinsic noise (right), as both showed nearly linear relationships between time and log amplitude.
+  The reduction in consistency between points seen for very small amplitude oscillations represents a fundamental limitation of the method, as noise dominates when oscillations approach steady state.
+
+\end{manuscript}
+
+Figure S7 indicated that although it was possible to differentiate between noise sources when only one type of noise was present, determining changes to one noise source when both are present is likely not identifiable from the data.
+Ultimately, however, we don't believe that the key insight of our method is determining the explicit source of the stochastic variation.
+Since both types of noise contribute to determining the population-level dynamics, quantifying the overall stochastic noise is the main goal.
+However, based on the evidence presented in this study and in previous studies, we offer the explanation that cell-autonomous noise is likely the principal source of rhythm damping.
 
@@
 
-I'll add a paragraph on practical considerations for implementing the algorithm.
+This is an excellent suggestion, and we have added additional sections to the manuscript describing practical considerations for experimental results.
+More important than the sampling rate is the overall duration of the experiment, as this allows a more confident regression of the amplitude window.
+
+The following paragraph was added to the discussion to address practical considerations in experimental design and method choice:
+
+\begin{manuscript}{Page 15}
+ In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
+ {\bfseries
+An overview of the computational steps involved in this method are outlined in Figure~S8.
+While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
+Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
+Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
+While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
+We also note that there are many available tools for detrending and regressing time-series data.
+While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.
+}
+\end{manuscript}
+
+Additionally, we have added a supplemental figure, Figure S8, to provide a more detailed overview of the method:
+
+\begin{manuscript}{Page 23}
+  \begin{center}
+  \includegraphics[width=0.8\textwidth]{figures/pdfs/flowchart_schematic.pdf}
+  \end{center}
+{\bfseries Figure S8: Overview of the computational method.} ({\bfseries A}) Raw bioluminescence traces are collected from cultured cells.
+Since the method is mainly dependent on amplitude over time, longer experiments yield more identifiable damping rates.
+({\bfseries B}) Raw traces are detrended and denoised.
+This can be accomplished using a variety of methods, including the discrete wavelet transform.
+({\bfseries C}) The window of the oscillations (amplitude vs.\ time) is plotted on a semi-log axis to make the exponential damping apparent.
+Window functions can be calculated using a Hilbert transform, or simply by using the absolute value of the local peaks (as was done here).
+A linear regression of the window function in semi-log space will yield both the initial amplitude (intercept) and damping rate (slope), with care being taken to avoid the over-influence of outlier points.
+Standard statistical techniques could then be used to verify that differences in slope are indeed significant.
+\end{manuscript}
+
+We elected not to include simulated time-series demonstrations of these points, mainly because of the difficulty in getting simulated time-series to match the variability seen in biological systems.
+Because of this, we believed recommended sampling rates and durations would not necessarily be reflective of biological data.
+
+In regards to using the Hilbert transform to align phases, this was done mainly to be able to analyze the dephasing kinetics from a single-cell perspective.
+Since there are relatively few sources of single-cell circadian data, this method is likely not applicable to many external studies.
+
+%
+
+% I'll add a paragraph on practical considerations for implementing the algorithm.
 Sampling rate is less important than overall recording duration.
@@
 
-An additional reference for Rougemont and Naef can be added, as they do discuss a variant of this.
-However, the idea that ballistic particles diffuse according to $t^2$ (and brownian particles according to $t$) is very old.
+We have largely re-worded this section of the text in response to the comments of other reviewers.
+Despite this, we have added a further citation to Rougement \& Naef, as they present a more detailed theoretical analysis of the windowing function:
+
+\begin{manuscript}{Page 21}
+   ({\itshape Right}) A deterministic, heterogeneous population dephases due to different free-running periods.
+In this case, the population displays ballistic phase diffusion, in which envelope of the population-averaged expression changes proportionally to $t^2$.
+{\bfseries (See Rougemont \& Naef, 2007 (22) for a more detailed discussion).}
+\end{manuscript}
+
+In general, however, the concept that Brownian particles diffuse with a variance that scales with $t$, and that ballistic particles diffuse with a variance that scales with $t^2$, is very old.
+The result presented here follows from the connection between amplitude and the variance of the phase probability.
 
@@
 
-This point will be clarified (our main idea is that noise is important), although our method does not assume self-sustained oscillations in the deterministic limit.
+We thank the reviewer for catching this inconsistency, have have updated the text accordingly:
+
+\begin{manuscript}{Page 3}
+  Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, {\bfseries capture the noise characteristics seen in single-cell fibroblast data equally well as limit-cycle oscillators (20).}
+\end{manuscript}
+
+However, we note that our assumptions and the concept of noise-driven oscillators are not necessarily exclusive.
+In the cited work, the authors demonstrate that given a constant noise input, the damped oscillator filters the noise to amplify noise in the circadian frequencies.
+This yields an oscillator with similarly sustained amplitudes to a noisy limit-cycle.
 
@@
 
-Will be added.
+In addition to expanding the discussion on cellular noise in response to a previous reviewer's comments, we have added some additional remarks on the importance of measuring stochastic noise:
+
+\begin{manuscript}{Page 3}
+  {\bfseries
+  The amount of noise in system is therefore linked to the ability of tissue-level clocks to maintain high-amplitude rhythms.
+
+Despite the averaging that occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
+Noise in circadian rhythms has long been considered an important factor in how circadian rhythms have evolved (18).} A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues (19).
+\end{manuscript}
+
+We also note that the importance of amplitude boosting is linked to the measurement of noise, as noise likely plays a key role in determining the rate at which rhythms damp in peripheral tissues.
 
@@
 
-We don't have access to the raw data of Liu et.
-al.,, but higher amplitude (as suggested by Liu) is not the same as stronger rhythms.
-I will check our database (many Cry2 siRNA replicates) to see if this trend is replicated in the siRNA screen.
+This is an interesting point.
+We are careful in saying that there may indeed be perturbations or knockouts which shift the clock to a region of higher stability, but that on average perturbations tend to lead to less robust rhythms.
+
+We sadly don't have access to the raw data of Liu et al.\ to look more closely at their observations.
+In the genome-wide siRNA screen, however, Cry2 knockdown was one of the several positive controls run in each plate.
+From this data, it seems that Cry2 knockdown seems to simultaneously reduce both the population-level amplitude and damping rate:
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=0.7\textwidth]{response_figs/cry2_ts.pdf}
+  \end{center}
+  \caption{{\bfseries Bioluminescence trajectories of control and Cry2 knockdown wells.} ({\itshape Left}) Control wells which did not contain an siRNA perturbation.
+({\itshape Right}) Wells with Cry2 siRNA.}
+\end{figure}
+
+Looking at the parameter fits of these data, we see that while Cry2 siRNA does indeed seem to make the oscillators less noisy (lower damping rate), it also slightly reduces their amplitude.
+These results are mirrored by the dose-dependent Cry2 knockdown from the same study (Zhang et al., Cell 2009), which shoes a reduction of amplitude at higher Cry2 si concentrations.
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=0.4\textwidth]{response_figs/cry2_dist.pdf}
+  \end{center}
+  \caption{{\bfseries Distribution of amplitude and damping rate fits for control and Cry2 siRNA wells.} The mean of both distributions is highlighted with a black dot.}
+\end{figure}
+
+However, since these points require comparing population-level amplitudes to single-cell level amplitudes (which may not be directly comparable), we decided to leave it out of the manuscript.
 
@@
 
-What is new is that this phenomenon can be quantified to give additional information on stochastic noise.
-From the datasets, this is an independent quantity (from period, amplitude), and therefore likely provides new insight into the underlying oscillator.
+We agree that the relationships between cell synchrony, stochastic noise, and damped oscillations have been previously explored in many fields, including circadian rhythms.
+In addition to a more through description of noise sources, we have included an explicit reference to the presence of these phenomena in other systems in the introduction:
 
-Also, we're attempting to downplaying the effect of cell-to-cell variability.
+\begin{manuscript}{Page 3}
+  {\bfseries This type of behavior has also been seen in
+  other experimental systems, such as NF-$\kappa$B signaling or yeast glycolytic oscillations (16, 17).}
+\end{manuscript}
+
+The novel contribution of our manuscript is that we can use changes in the damping rates of population-level data to gain new insight into the relative levels of noise present under different perturbations.
+From the datasets we analyze, damping appears to be independently altered (from period, amplitude) by genetic perturbations, and therefore likely provides new insight into the underlying oscillator.
+Our main message is that the damping rate is due to cellular noise, but that the damping rate contains information about the oscillator which is currently going unused.
 
@@
 
-We are demonstrating that single-cell noise is {\itshape sufficient} to explain the observed damping behavior, and indeed quantitatively accurate predictions for population-level damping rate are obtained from a simple model not fit to the noise properties.
+As in all modeling work, we must make simplifying assumptions that may be drawn from incomplete experimental data.
+In developing our approach for inferring noise properties from population-level data, we have invoked two main assumptions: that individual cells are both independent and sustained oscillators.
+We are careful to list these assumptions in the introduction, as we agree that the validity of our results rests upon whether or not these assumptions hold in the systems under consideration.
+However, to the best of our knowledge, there is relatively strong experimental support to both of these assumptions in cultured circadian reporter cells.
+In the earlier parts of our paper, we are demonstrating that single-cell noise is {\itshape sufficient} to explain the observed damping behavior, and indeed quantitatively accurate predictions for population-level damping rate are obtained from a simple model not fit to the noise properties.
 
-Indeed, it would be great if single-cell data was available for each siRNA perturbation. This is likely not possible, as single-cell data is difficult to collect.
-
-Cells are likely not arrhythmic as a result of siRNA perturbation, as the siRNA method has been well-adopted by the circadian community for effects on amplitude and period.
+In regards to other potential explanations for the effects of siRNA, we suspect intercellular signalling networks do not play a significant role in cultured cell oscillations (as we assume each cell is independent).
+These perturbations are thought to affect the cell-autonomous transcription-translation feedback loop, resulting in potentially altered noise properties.
+Indeed, the siRNA method has been well-accepted by the circadian community, and are widely used to study the effect of gene knockdowns on amplitude and period.
 
@@
 
-This has been corrected.
+We thank the reviewer for pointing out this error. We have corrected our usage throughout the manuscript.
 

### 2015-05-06 main_text.tex

@@
 Stochastic noise at the cellular level has been shown to play a fundamental role in circadian oscillations, influencing how groups of cells entrain to external cues and likely serving as the mechanism by which cell-autonomous rhythms are generated.
-Despite this importance, few studies have investigated how clock perturbations affect stochastic noise - even as increasing numbers of high-throughput screens categorize how gene knockdowns or small molecules can change clock period and amplitude.
+Despite this importance, few studies have investigated how clock perturbations affect stochastic noise -- even as increasing numbers of high-throughput screens categorize how gene knockdowns or small molecules can change clock period and amplitude.
 This absence is likely due to the difficulty associated with measuring cell-autonomous stochastic noise directly, which currently requires the careful collection and processing of single-cell data.
-In this study, we show that the damping rate of population-level bioluminescence recordings can serve as an accurate measure of cell-autonomous stochastic noise, and one which can be applied to future and existing high-throughput circadian screens.
+In this study, we show that the damping rate of population-level bioluminescence recordings can serve as an accurate measure of overall stochastic noise, and one that can be applied to future and existing high-throughput circadian screens.
 Using cell-autonomous fibroblast data, we first show directly that higher noise at the single-cell results in faster damping at the population level.
@@
 Finally, we hypothesize that the unperturbed clock is highly optimized for robust rhythms, as very few gene perturbations are capable of simultaneously increasing amplitude and lowering stochastic noise.
-Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on cell-autonomous noise, particularly with regard to the development of small-molecule circadian therapeutics.
+Ultimately, this study demonstrates the importance of considering the effect of circadian perturbations on stochastic noise, particularly with regard to the development of small-molecule circadian therapeutics.
 \end{abstract}
@@
 
-As most organisms exist in an environment which changes predictably with a 24-hour period, highly optimized genetic circuits turn on and off the production of key regulatory proteins to anticipate the day/night cycle.
+As most organisms exist in an environment that changes predictably with a 24-hour period, highly optimized genetic circuits turn on and off the production of key regulatory proteins to anticipate the day/night cycle.
 In humans, the demands of a modern society have required that we deviate from this evolutionarily prescribed sleep and feeding schedule, resulting in increased long-term risks of metabolic disease.
-There is therefore a desire to find pharmacological treatments which would restore the normal functioning of our circadian clock despite irregular behavioral schedules.
-One aspect of these treatments which is often overlooked in searching for candidate drugs is how these treatments might affect the accuracy of the circadian timing system.
+There is therefore a desire to find pharmacological treatments that would restore the normal functioning of our circadian clock despite irregular behavioral schedules.
+One aspect of these treatments that is often overlooked in searching for candidate drugs is how these treatments might affect the accuracy of the circadian timing system.
 % Since each cell in the body keeps its own internal time, increasing the sloppiness of the clock machinery can result in tissues not being able to agree on what time to turn on protein production.
@@
 In this paper, we show that it is possible to estimate how much the noise of a system has changed by looking only at the averaged protein production of an entire population of cells.
-Such an approach allows us to analyze prior data from high-throughput screens, and show that the unperturbed clock has been highly optimized to be both accurate and high amplitude.
+Such an approach allows us to analyze prior data from high-throughput screens, and show that the natural clock has been highly optimized to be both accurate and high amplitude.
 
@@
 In mammals, such rhythms are organized in a hierarchical fashion: at the tissue-level, the brain's suprachiasmatic nucleus (SCN) serves as the master pacemaker and keeps circadian oscillations in peripheral tissues in phase with the light-dark cycle.
-Cell-to-cell coupling in the SCN keeps individual cells in tight synchrony \cite{Herzog2004}. 
-It is currently thought that coupling between circadian oscillations outside the SCN, such as in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro}, is very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
-Within each tissue, cellular-level rhythms in gene transcription are generated by a large network of interacting regulatory elements, in which time-delayed negative feedback gives rise to sustained oscillations \cite{Ueda2005}.
+In the SCN, cell-to-cell coupling keeps individual cells in tight synchrony \cite{Herzog2004}, while coupling between circadian oscillations in peripheral tissues {\itshape in vivo} or cultured reporter cells {\itshape in vitro} is thought to be very weak or absent entirely \cite{Guenthner2014, Noguchi2013}.
+Within each tissue, cellular-level rhythms in gene transcription are generated by a large network of interacting gene regulatory elements, in which time-delayed transcription-translation negative feedback gives rise to sustained oscillations \cite{Ueda2005}.
 The robust oscillation of circadian factors has been linked to metabolic health \cite{Bass2012}, since rhythms compromised by gene knockout \cite{Marcheva2010} or irregular feeding schedules \cite{Hatori2012} result in an increased risk of metabolic disease.
-Additionally, as the amplitude of circadian transcription can be affected by lifestyle variables such as diet, age, or work schedules, there has been recent interest in developing pharmacological strategies for increasing the amplitude of circadian cycles in metabolic tissues \cite{Chen2013}.
+Additionally, as the amplitude of circadian transcription can be affected by lifestyle variables such as diet, age, or work schedule, there has been recent interest in developing pharmacological strategies for increasing the amplitude of circadian cycles in metabolic tissues \cite{Chen2013}.
 
@@
 The functional roles of different genes in circadian regulation have traditionally been studied using behavioral-level data and genetic knockout experiments \cite{Vitaterna1994}.
-Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors which affect circadian rhythmicity \cite{Zhang2009}.
+Bioluminescence-based cellular circadian reporters offer a more direct view of the gene regulatory network \cite{Balsalobre1998} and are amenable to high-throughput screens, allowing genome-wide exploration into factors that affect circadian rhythmicity \cite{Zhang2009}.
 Additionally, cultured circadian reporter cells allow the change in transcriptional amplitude following a perturbation to be quantified.
@@
 Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells.
-At the single-cell level, transcription is strongly affected by cellular noise, comprised of both {\itshape intrinsic} and {\itshape extrinsic} sources.
-Intrinsic noise is generated from the low molecular counts of the mRNA and protein species involved.
-As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length \cite{Welsh2004}. 
+Transcription at the single-cell level is strongly affected by {\itshape intrinsic} cellular noise, caused by the low molecular counts of the mRNA and protein species involved. 
+As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length from cycle to cycle \cite{Welsh2004}. 
+In addition to intrinsic noise, circadian oscillations are also affected by {\itshape extrinsic} noise sources.
 Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude.
-For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle has been shown to be larger than the variability in mean period length between cells \cite{Herzog2004}.
-In cell cultures which lack cell-to-cell coupling, it has been shown that  stochastic noise at the single-cell level is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.
+For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells \cite{Herzog2004}.
+Both sources of noise have an effect on population-level rhythms: in cell cultures that lack cell-to-cell coupling, it has been shown that  stochastic noise is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.
+This type of behavior has also been seen in other experimental systems, such as NF-$\kappa$B signaling or yeast glycolytic oscillations \cite{Nelson2004, Aon1992}.
+The amount of noise in system is therefore linked to the ability of tissue-level clocks to maintain high-amplitude rhythms.
 
-Despite the averaging which occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
+Despite the averaging that occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
+Noise in circadian rhythms has long been considered an important factor in how circadian rhythms have evolved \cite{Barkai2000}.
 A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues \cite{An2013}.
-Additional studies have argued that the basis of single-cell rhythmicity depends on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, better capture the noise characteristics seen in single-cell fibroblast data \cite{Westermark2009}.
+Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, capture the noise characteristics seen in single-cell fibroblast data equally well as limit-cycle oscillators \cite{Westermark2009}.
 Despite the importance of single-cell stochasticity in circadian rhythms, measuring stochastic noise currently requires careful preparation, recording, and image processing of individual cells \cite{Leise2012}.
-As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to single-cell noise as a result of a small molecule or genetic perturbation.
+As a result, while circadian perturbations have been postulated to affect single-cell stochasticity \cite{Rougemont2007}, no study has experimentally quantified changes to stochastic noise as a result of a small molecule or genetic perturbation.
 
-In this study, we demonstrate that changes to single-cell stochasticity can be reliably inferred from the changes to the damping rate of population-level bioluminescence recordings of cultured circadian reporters.
+In this study, we demonstrate that changes to stochastic noise can be reliably inferred from the changes to the damping rate of population-level bioluminescence recordings of cultured circadian reporters.
 Our method assumes that oscillations in individual cells are both
@@
 We demonstrate the validity and usefulness of such an approach on several types of circadian data.
-First, we show using single-cell fibroblast data that stochastic noise is directly related to population-level damping.
-Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to single-cell stochasticity is a likely mechanism.
-Finally, we calculate the genome-wide effects of siRNA knockdown on single-cell noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
+First, we show using single-cell fibroblast data that intrinsic stochastic noise is directly related to population-level damping.
+Next, we show that a small-molecule modulator is able to change damping rate in a dose-dependent fashion, and verify using a mathematical model that changes to intrinsic stochastic noise is a likely mechanism.
+Finally, we calculate the genome-wide effects of siRNA knockdown on overall stochastic noise, and demonstrate that population-level damping rate is independent of other circadian parameters, such as period or amplitude.
 Using this additional information, we show that circadian rhythms have likely evolved to an optimal point of high amplitude and low stochastic noise.
-Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on single-cell noise to be quantified in a high-throughput manner.
+Our results should prove especially important in the future search for small molecule circadian therapeutics, as it allows the effect of candidate drugs on stochastic noise to be quantified in a high-throughput manner.
 
@@
 Amplitude, $A$, and phase, $\theta$, parameters are subsequently fit using a linear least-squares regression.
+Note that in this manuscript we use amplitude to denote the initial rhythm strength, and damping rate to denote the rate at which this strength decays with time.
 Overall $R^2$ values for the regression were calculated from the residual error between the detrended data and fitted sinusoid:
@@
 As was done in the original study, a discrete wavelet transform (using PyWavelets, \url{http://www.pybytes.com/pywavelets}) was performed to detrend and remove noise.
+A discrete wavelet transform decomposes the signal into multiple frequency bands \cite{Leise2011}. 
+By only considering frequency bands close to the circadian frequency, high-frequency noise and low-frequency baseline oscillations can be removed.
 
@@
 As in the original study, various parameters describing the average noise level of each cell were collected.
-Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components.
+Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components -- resulting in rhythms which only contained oscillations with periods between 8 and 258 hours.
 From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
 An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
+This wavelet component was chosen as it contained only the highest-frequency noise in the signal's spectrum.
 From these three noise variables, a combined noise metric was constructed by projecting the variables along their first principle component (using scikit-learn, \url{http://scikit-learn.org/}).
@@
 
-We first sought to determine whether changes to intrinsic stochastic noise are sufficient to explain population-level changes in damping rate.
+While both intrinsic and extrinsic noise sources can contribute to population-level damping, intrinsic noise is thought to play a more significant role in circadian systems \cite{Herzog2004}.
+We therefore first sought to determine whether changes to intrinsic stochastic noise alone are sufficient to explain population-level changes in damping rate.
 To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
@@
 This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc} U2OS cells), indicating it is a fundamental property of the overall gene regulatory network.
+This result indicates that stochastic noise can be altered by perturbations known to affect the transcription-translation feedback loop.
 
@@
 
-\subsection*{Mathematical model predicts KL001 damping rate increase comes from increased single-cell stochastic noise}
+\subsubsection*{Mathematical model predicts KL001 damping rate increase comes from increased single-cell stochastic noise}
 
@@
 As the model includes no cell-to-cell communication or heterogeneity, this difference is manifested solely by changing the noise characteristics of individual cells.
-This computational result
+This quantitative prediction by the mathematical model lends further support to the assumptions that individual cells are independent and sustained oscillators, and intrinsic noise plays the dominant role in determining population-level damping.
 
@@
 Additionally, of the fits with a high $R^2$ value, only a small minority ($0.1\%$) had a negative damping rate.
-This trend lends further support to the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
+This trend supports the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
 
@@
   \end{center}
-  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} Distributions in robust z-scores closely resemble normal distributions.
+  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} For each well in the high-throughput screen, the period, amplitude and damping rate are calculated.
+After normalization, distributions in robust z-scores closely resemble normal distributions.
 For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.}
@@
 
-Since an siRNA's effect on period length does not effectively predict its effect on amplitude or phase (and vice-versa), we hypothesized that stochastic noise would similarly be independently affected by siRNA perturbations.
+Since an siRNA's effect on period length does not effectively predict its effect on amplitude or phase (and vice-versa), we hypothesized that damping rate would similarly be independently affected by siRNA perturbations.
 Low Pearson correlation coefficients between normalized parameter distributions defend this hypothesis, with the highest correlations among variables seen between amplitude and damping rate ($\rho = 0.285$, Table~\ref{tab:corr}).
@@
 The ability of a population of oscillators to maintain robust oscillations is a function both of its initial amplitude as well as its damping rate.
-A robust oscillator is therefore one with high amplitude and low stochastic noise.
+A robust oscillator is therefore one with high initial amplitude and low stochastic noise.
 Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate that few outlier points reside in the lower damping rate, higher amplitude quadrant.
@@
 Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
-While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, experimental evidence suggests intrinsic stochastic noise plays a more significant role.
-We show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}).
+While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise likely plays a more significant role: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}).
 This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise \cite{Herzog2004}.
@@
 However, it is possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
-Such a change, caused perhaps by an unequal uptake of drug across cells, would result in an altered distribution in mean period length, and different dephasing kinetics.
+Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics.
+While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure~\ref{fig:dephasing-rates}), these differences are likely not identifiable from typical data (Figure~\ref{fig:computational_dephasing}).
+Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
+Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.
 
-% Additionally, it is important to note that the dephasing kinetics of these two alternative processes result in different dynamics at the population-level.
-In dephasing due solely to stochastic noise, the probability density of finding a cell with any given phase follows a standard convection-diffusion regime, resulting in exponentially damped rhythms \cite{St.John2014a}.
-In dephasing due solely to period heterogeneity, the variance of the phase probability density broadens in time proportionally to $t^2$, rather than $t$ in standard diffusion.
-This type of anomalous diffusion results in a different damping profile, as highlighted in Figure~\ref{fig:dephasing-rates}.
-Since the data used in this study is well-described by an exponentially damped sinusoid, and since a mathematical model assuming homogeneous mean periods accurately predicts experimental results, we expect our assumption that changes to intrinsic noise are primarily responsible for the observed damping rate changes to be justified.
-
-However, it is possible that damping rate changes due to siRNA or small molecule perturbation are manifested by altering the system's extrinsic noise.
-Such a change, caused perhaps by an unequal uptake of drug across cells, would result in an altered distribution in mean period length, and different dephasing kinetics.
-A simple model of a stochastic oscillator was used to investigate the effects of altering intrinsic and extrinsic noise on the dephasing kinetics (Figure~\ref{fig:computational_dephasing}).
-While {\itshape in silico} dephasing due to solely intrinsic or extrinsic sources can be differentiated from population-level rhythms, it becomes more difficult to identify noise sources when multiple types of noise are present.
-The window function of population-level oscillations can be affected by many factors, including weak cell-to-cell coupling \cite{Rougemont2007}, temporary amplitude change \cite{St.John2014a}, and intrinsic and extrinsic noise.
-While altering intrinsic, or cell-autonomous, noise has been shown to be sufficient to explain the observed damping rate changes, it is possible that a portion of the changes to damping rate observed in regressed experimental data may be due to changes in extrinsic noise.
+% % Additionally, it is important to note that the dephasing kinetics of these two alternative processes result in different dynamics at the population-level.
+% In dephasing due solely to stochastic noise, the probability density of finding a cell with any given phase follows a standard convection-diffusion regime, resulting in exponentially damped rhythms \cite{St.John2014a}.
+% In dephasing due solely to period heterogeneity, the variance of the phase probability density broadens in time proportionally to $t^2$, rather than $t$ in standard diffusion.
+% This type of anomalous diffusion results in a different damping profile, as highlighted in Figure~\ref{fig:dephasing-rates}.
+% Since the data used in this study is well-described by an exponentially damped sinusoid, and since a mathematical model assuming homogeneous mean periods accurately predicts experimental results, we expect our assumption that changes to intrinsic noise are primarily responsible for the observed damping rate changes to be justified.
+% 
+% However, it is possible that damping rate changes due to siRNA or small molecule perturbation are manifested by altering the system's extrinsic noise.
+% Such a change, caused perhaps by an unequal uptake of drug across cells, would result in an altered distribution in mean period length, and different dephasing kinetics.
+% A simple model of a stochastic oscillator was used to investigate the effects of altering intrinsic and extrinsic noise on the dephasing kinetics (Figure~\ref{fig:computational_dephasing}).
+% While {\itshape in silico} dephasing due to solely intrinsic or extrinsic sources can be differentiated from population-level rhythms, it becomes more difficult to identify noise sources when multiple types of noise are present.
+% The window function of population-level oscillations can be affected by many factors, including weak cell-to-cell coupling \cite{Rougemont2007}, temporary amplitude change \cite{St.John2014a}, and intrinsic and extrinsic noise.
+% While altering intrinsic, or cell-autonomous, noise has been shown to be sufficient to explain the observed damping rate changes, it is possible that a portion of the changes to damping rate observed in regressed experimental data may be due to changes in extrinsic noise.
 
@@
 In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
+An overview of the computational steps involved in this method are outlined in Figure~\ref{fig:comp_overview}.
+While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
+Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
+Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
+While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
+We also note that there are many available tools for detrending and regressing time-series data.
+While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.
+
 As high amplitude circadian oscillations are important for maintaining metabolic health, many studies have sought to find small molecule candidates which increase oscillatory amplitude \cite{Chen2013}.
@@
   ({\itshape Right}) A deterministic, heterogeneous population dephases due to different free-running periods.
-In this case, the population displays ballistic phase diffusion, in which envelope of the population-averaged expression changes proportionally to $t^2$.}
+In this case, the population displays ballistic phase diffusion, in which envelope of the population-averaged expression changes proportionally to $t^2$.
+(See Rougemont \& Naef, 2007 \cite{Rougemont2007} for a more detailed discussion).}
   \label{fig:dephasing-rates}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS7.pdf}
+    \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS7.pdf}
   \end{center}
-  \caption{{\bfseries Intrinsic and extrinsic noise can be difficult to differentiate in .} ({\bfseries A}) }
+  \caption{{\bfseries Identifiability of intrinsic and extrinsic noise factors in population-level data.} A simple model with both intrinsic and extrinsic noise (Table~\ref{tab:novak}) was used to determine the identifiability of noise sources.
+Bioluminescence profiles were generated from a population of 1000 individual oscillators.
+The stochastic volume parameter, $\Omega$, controls the intrinsic noise, with larger volumes resulting in more deterministic profiles.
+Extrinsic noise was generated by using a normal distribution of free running periods, ${\cal N}(1, \sigma^2_T)$ days.
+Points represent window functions (amplitude over time) from 10 independent replications.
+Lines represent linear regressions, except in the right panel of part A, where a quadratic regression was used.
+  ({\bfseries A}) For models with pure intrinsic (left) or pure extrinsic (right) noise sources, amplitude damping proportional to $e^{dt}$ vs.\ $e^{dt^2}$, respectively, was readily distinguishable. 
+  ({\bfseries B}) For models with both intrinsic and extrinsic noise sources, it was not possible to differentiate between increasing intrinsic noise (left) and increasing extrinsic noise (right), as both showed nearly linear relationships between time and log amplitude.
+  The reduction in consistency between points seen for very small amplitude oscillations represents a fundamental limitation of the method, as noise dominates when oscillations approach steady state.
+}
   \label{fig:computational_dephasing}
+\end{figure}
+
+\begin{figure}[tbp]
+  \begin{center}
+    \includegraphics[width=\textwidth]{figures/pdfs/flowchart_schematic.pdf}
+  \end{center}
+  \caption{{\bfseries Overview of the computational method.} ({\bfseries A}) Raw bioluminescence traces are collected from cultured cells.
+Since the method is mainly dependent on amplitude over time, longer experiments yield more identifiable damping rates.
+({\bfseries B}) Raw traces are detrended and denoised.
+This can be accomplished using a variety of methods, including the discrete wavelet transform.
+({\bfseries C}) The window of the oscillations (amplitude vs.\ time) is plotted on a semi-log axis to make the exponential damping apparent.
+Window functions can be calculated using a Hilbert transform, or simply by using the absolute value of the local peaks (as was done here).
+A linear regression of the window function in semi-log space will yield both the initial amplitude (intercept) and damping rate (slope), with care being taken to avoid the over-influence of outlier points.
+Standard statistical techniques could then be used to verify that differences in slope are indeed significant.}
+  \label{fig:comp_overview}
 \end{figure}
@@
 \begin{table}
+  \caption{{\bfseries Simple model of transcription-translation oscillations with intrinsic and extrinsic noise.} A model with both intrinsic and extrinsic noise, developed to examine the effect of each noise source on population-level amplitudes.
+Intrinsic noise is generated by simulating the solution stochastically using GillesPy.
+Extrinsic noise is generated by varying the free-running period, $t_c$.
+Equations adapted from \cite{Novak2008}.}
+  \centering
+  \begin{align*}
+    \frac{dX}{dt} &= t_c \left(\frac{1}{1 + Y} - X\right) \\
+    \frac{dY}{dt} &= t_c \left(k_t \; X - k_d \; Y - \frac{Y}{\alpha_0 + \alpha_1 \; Y + \alpha_2 Y^2}\right)
+  \end{align*}
+
+  \begin{tabular}{rl}
+    \toprule Parameter & Value \\\midrule
+    $k_t$ & 20 \\
+    $k_d$ & 1  \\
+    $P$   & 4 (or 2)  \\
+    $\alpha_0$ & 0.005 \\
+    $\alpha_1$ & 0.05  \\
+    $\alpha_2$ & 0.1   \\
+    $t_c$ & 3.516 \\\bottomrule
+    \end{tabular}
+  \label{tab:novak}
+\end{table}
+
+\begin{table}
 \begin{center}

### 2015-05-06 main_text.tex
Renaming figure files for plos conventions

@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/noise_ts_and_boot.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig1.pdf}
   \end{center}
@@
   \begin{center}
-    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/mainfig_dose.pdf}}
+    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig2.pdf}}
   \end{center}
@@
   \begin{center}
-    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/main_fig_simulations.pdf}}
+    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig3.pdf}}
   \end{center}
@@
   \begin{center}
-    \includegraphics[]{figures/pdfs/r2_vs_damping.pdf}
+    \includegraphics[]{figures/pdfs/Fig4.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/fitted_parameters.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig5.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.65\textwidth]{figures/pdfs/outliers.pdf}
+    \includegraphics[width=0.65\textwidth]{figures/pdfs/Fig6.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/pdfs/small_molecule_r2.pdf}
+    \includegraphics[width=0.5\textwidth]{figures/pdfs/FigS1.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/pdfs/volume_calibration.pdf}
+    \includegraphics[width=0.5\textwidth]{figures/pdfs/FigS2.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/zhang_plates.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS3.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/zhang_wells.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS4.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{figures/pdfs/flowchart_schematic.pdf}
+    \includegraphics[width=\textwidth]{figures/pdfs/FigS8.pdf}
   \end{center}

### 2015-05-06 response_letter.tex
updating pages and changed text in response

@@
 Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells. 
-{\bfseries
-  Transcription at the single-cell level is strongly affected by {\itshape intrinsic} cellular noise, caused by the low molecular counts of the mRNA and protein species involved.
-As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length from cycle to cycle (14).
+Transcription at the single-cell level is strongly affected by {\bfseries {\itshape intrinsic} cellular noise, caused by} the low molecular counts of the mRNA and protein species involved.
+As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length from {\bfseries cycle to cycle (14).
 In addition to intrinsic noise, circadian oscillations are also affected by {\itshape extrinsic} noise sources.
@@
 For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells (2).
-Both sources of noise have an effect on population-level rhythms: in cell cultures that lack cell-to-cell coupling, it has been shown that stochastic noise is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase (14, 15).
-}
+Both sources of noise have an effect on population-level rhythms:} in cell cultures that lack cell-to-cell coupling, {\bfseries it has been shown that stochastic noise} is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase (14, 15).
+
 \end{manuscript}
@@
 
-\begin{manuscript}{Page 13}
+\begin{manuscript}{Page 8}
   \section*{Effect of cell heterogeneity}
@@
 In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping. 
-However, experimental work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods (21). 
+However, experimental} work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods (21). 
 Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping (30). 
-While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise likely plays a more significant role: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure S5).
+While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, {\bfseries intrinsic stochastic noise likely plays a more significant role}: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis {\bfseries in cultured fibroblast cells (Figure S5).
 This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise (2).
@@
 Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors. 
-Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.
-}
+Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.}
 \end{manuscript}
@@
 
-\begin{manuscript}{Page 5}
+\begin{manuscript}{Page 7}
   Figure 5: Distributions in fitted parameters for the genome-wide siRNA screen.
@@
 
-\begin{manuscript}{Pages 4-5}
+\begin{manuscript}{Pages 10}
   As was done in the original study, a discrete wavelet transform (using PyWavelets, \url{http://www.pybytes.com/pywavelets}) was performed to detrend and remove noise.
@@
 
-\begin{manuscript}{Page 5}
+\begin{manuscript}{Page 10}
 As in the original study, various parameters describing the average noise level of each cell were collected.
@@
 
-\begin{manuscript}{Page 4}
+\begin{manuscript}{Page 10}
  For numerical efficiency, the period, $T$, and damping rate, $d$, parameters are fit first using a matrix pencil method (24), reviewed in (25).
@@
 
-\begin{manuscript}{Page 22}
+\begin{manuscript}{Page 16}
   \begin{center}
@@
 
-\begin{manuscript}{Page 15}
+\begin{manuscript}{Page 9}
  In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
@@
 
-\begin{manuscript}{Page 23}
+\begin{manuscript}{Page 16}
   \begin{center}
-  \includegraphics[width=0.8\textwidth]{figures/pdfs/flowchart_schematic.pdf}
+  \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS8.pdf}
   \end{center}
@@
 
-\begin{manuscript}{Page 21}
+\begin{manuscript}{Page 16}
    ({\itshape Right}) A deterministic, heterogeneous population dephases due to different free-running periods.
@@
 \begin{manuscript}{Page 3}
-  Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, {\bfseries capture the noise characteristics seen in single-cell fibroblast data equally well as limit-cycle oscillators (20).}
+  Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, capture the noise characteristics seen in single-cell fibroblast data {\bfseries equally well as limit-cycle oscillators (20).}
 \end{manuscript}
@@
   {\bfseries
-  The amount of noise in system is therefore linked to the ability of tissue-level clocks to maintain high-amplitude rhythms.
+  The amount of noise in system is therefore linked to the ability of tissue-level clocks to maintain high-amplitude rhythms. }
 
 Despite the averaging that occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
-Noise in circadian rhythms has long been considered an important factor in how circadian rhythms have evolved (18).} A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues (19).
+{\bfseries Noise in circadian rhythms has long been considered an important factor in how circadian rhythms have evolved (18).} A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues (19).
 \end{manuscript}
@@
 
+\vspace{4ex}
+\begin{flushright}
+  Sincerely,\\[2ex]
+
+  \includegraphics[width=0.3\textwidth]{/home/peter/Documents/Research/figures/frank_signature.pdf}\\[1ex]
+Francis J. Doyle III
+\end{flushright}
+
 \end{document}

### 2015-05-06 response_letter.tex
final edits prior to sending to frank

@@
 We thank the reviewer for their overall positive reception of the manuscript. 
-In regards to their first critique, we mainly are considering intrinsic noise generated by the molecular mechanism.
-In the study, we demonstrate that single-cell noise is {\itshape sufficient} to explain the observed changes in population damping.
+In regards to their first critique, we are considering both intrinsic and extrinsic noise. 
+In the study, we demonstrate that single-cell (intrinsic) noise is {\itshape sufficient} to explain the observed changes in population damping.
 While sources of extrinsic noise are certainly present, experimental studies (i.e., Herzog {\itshape et al., JBR} 2004) have largely concluded that period lengths of circadian oscillations are more variable at the single-cell level (i.e., peak-to-peak time) than at the population level (differences in mean period between cells).
@@
 
-This is an valid concern, and one which was raised by multiple reviewers.
+This is an valid concern, and one that was raised by multiple reviewers.
 In the original manuscript we mentioned that intrinsic (cycle-to-cycle variability) and extrinsic (period heterogeneity) sources could be differentiated based on their envelope functions.
@@
 Indeed, the possibility that period heterogeneity may be changed by clock modulators has been added as discussed above.
-The increase in variance in Table 1, however, is most likely due to the variety of genes targeted by the siRNA library.
-At the single-cell level, each cell in the well would experience the same siRNA perturbation, which we suspect would lead to a similar change in period / amplitude.
-This is echoed by the consistency seen in the control wells.
-As this was perhaps a misunderstanding of the figure, we have added the following text to the caption of Figure 5:
+The increase in variance in Table 1, however, is largely due to the variety of genes targeted by the siRNA library.
+At the single-cell level, each cell in the well would experience the same siRNA perturbation, which we suspect would lead to a similar change in period, amplitude, and intrinsic noise.
+This suspicion is motivated by the consistency seen in the control wells.
+As this was perhaps a misunderstanding of the table, we have added the following text to the caption of Figure 5:
 
@@
 
-In regards to the reviewer's first comment, that amplitude and noise cannot be independently altered, we note that circadian oscillations are an emergent property from a very large transcriptional network.
-As such, the species present at low quantities which dominate the noise generation are not necessarily the same as those which are tied to the bioluminescence reporter.
-A reduction in effective concentration of a repressor, for instance, might lead to higher-amplitude transcription which is more susceptible to intrinsic noise.
+In regards to the reviewer's first comment -- that amplitude and noise cannot be independently altered -- we note that circadian oscillations are an emergent property from a very large transcriptional network.
+As such, the species present at low quantities that dominate the noise generation are not necessarily the same as those that are tied to the bioluminescence reporter.
+A reduction in effective concentration of a repressor, for instance, might lead to higher-amplitude transcription that is more susceptible to intrinsic noise.
 In addition to the linear independence we show between amplitude and damping rate, previous authors have also shown this independence: {\itshape ``The amplitude of rhythmicity also did not correlate with cycle-to-cycle variation of locomotor ($p = 0.17$), explant ($p = 0.24$), or cellular rhythms ($p = 0.15$)''} (Herzog {\itshape et al., JBR} 2004).
+This point has been added to the manuscript in the section describing the effect of period heterogeneity:
+
+\begin{manuscript}{Page 8}
+ we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure S5).
+{\bfseries This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise (2).}
+\end{manuscript}
 
@@
 This was actually the subject of a previous manuscript, St.\ John {\itshape et al., Biophys J}, 2014, which shows that the temporary amplitude change would largely be distinguished by a deviation from the exponential damping.
-This underscores the importance of collecting several days of data to accurately judge the damping rate.
-Additionally, for the siRNA screen, the synchronizing pulse (medium change) is consistent across all the wells.
-Since our method mainly analyses {\itshape differences} in damping rate, we suspect these effects would be minimal.
+Since the each of the data sets we show is well-described by an exponential amplitude window, we suspect initial transients do not play a substantial role in these results.
+Additionally, for the siRNA screen, the synchronizing pulse (medium change) is consistent across all the wells -- since our method mainly analyses {\itshape differences} in damping rate, this may also minimize its influence.
+
+However, this point does underscore the importance of collecting several days of data to accurately judge the damping rate.
+We have therefore added an additional comment to the discussion regarding the effect of initial transients.
+
+\begin{manuscript}{Page 9}
+  {\bfseries Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.}
+\end{manuscript}
 
@@
 
-We have added additional text to the methods section detailing the wavelet transform we used
+We have added additional text to the methods section detailing the wavelet transform we used:
 
@@
 {\bfseries
-Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components -- resulting in rhythms which only contained oscillations with periods between 8 and 258 hours.
+Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components -- resulting in rhythms that only contained oscillations with periods between 8 and 258 hours.
 }
@@
 
-We note that we do not intend {\itshape damping} to imply reduced amplitude.
-Instead, we use damping to indicate the rate at which the amplitude decays.
+We note that we did not intend {\itshape damping} to imply reduced amplitude.
+Instead, we use the term ``damping'' to indicate the rate at which the amplitude decays.
 Our analysis of the effect of noise on damping rate apply mainly to changes in noise and the resulting changes in exponential amplitude decay.
@@
 {\bfseries
-Note that in this manuscript we use amplitude to denote the initial rhythm strength, and damping rate to denote the rate at which this strength decays with time.}
+  Note that in this manuscript we use {\itshape amplitude} to denote the initial rhythm strength, and {\itshape damping rate} to denote the rate at which this strength decays with time.}
 \end{manuscript}
@@
 
-These concerns were also shared by the first reviewer, and we have therefore substantially changed the language of the study to reflect the fact that extrinsic noise (period heterogeneity) could play a role in the damping rates.
+These concerns were also shared by the first reviewer, and we have therefore {\bfseries substantially changed the language of the study to reflect the fact that extrinsic noise (period heterogeneity) could play a role in the damping rates.}
 For Fig. 3A, these profiles were obtained from a mathematical model of homogeneous cells.
@@
   \begin{center}
-  \includegraphics[width=0.51\textwidth]{figures/pdfs/FigS7.pdf}
+  \includegraphics[width=0.71\textwidth]{figures/pdfs/FigS7.pdf}
   \end{center}
@@
 More important than the sampling rate is the overall duration of the experiment, as this allows a more confident regression of the amplitude window.
-
 The following paragraph was added to the discussion to address practical considerations in experimental design and method choice:
@@
   \begin{center}
-  \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS8.pdf}
+  \includegraphics[width=\textwidth]{figures/pdfs/FigS8.pdf}
   \end{center}
@@
 
-We have largely re-worded this section of the text in response to the comments of other reviewers.
+{\bfseries We have largely re-worded this section of the text} in response to the comments of other reviewers.
 Despite this, we have added a further citation to Rougement \& Naef, as they present a more detailed theoretical analysis of the windowing function:
@@
 In the cited work, the authors demonstrate that given a constant noise input, the damped oscillator filters the noise to amplify noise in the circadian frequencies.
-This yields an oscillator with similarly sustained amplitudes to a noisy limit-cycle.
+This effect yields an oscillator with similarly sustained amplitudes to a noisy limit-cycle.
 
@@
 This is an interesting point.
-We are careful in saying that there may indeed be perturbations or knockouts which shift the clock to a region of higher stability, but that on average perturbations tend to lead to less robust rhythms.
+We are careful in saying that there may indeed be perturbations or knockouts that shift the clock to a region of higher stability, but that on average perturbations tend to lead to less robust rhythms.
 
@@
   \end{center}
-  \caption{{\bfseries Bioluminescence trajectories of control and Cry2 knockdown wells.} ({\itshape Left}) Control wells which did not contain an siRNA perturbation.
+  \caption{{\bfseries Bioluminescence trajectories of control and Cry2 knockdown wells.} ({\itshape Left}) Control wells that did not contain an siRNA perturbation.
 ({\itshape Right}) Wells with Cry2 siRNA.}
@@
 Looking at the parameter fits of these data, we see that while Cry2 siRNA does indeed seem to make the oscillators less noisy (lower damping rate), it also slightly reduces their amplitude.
-These results are mirrored by the dose-dependent Cry2 knockdown from the same study (Zhang et al., Cell 2009), which shoes a reduction of amplitude at higher Cry2 si concentrations.
+These results are mirrored by the dose-dependent Cry2 knockdown from the same study (Zhang et al., Cell 2009), which shows a reduction of amplitude at higher Cry2 si concentrations:
+\clearpage
 
@@
 
-However, since these points require comparing population-level amplitudes to single-cell level amplitudes (which may not be directly comparable), we decided to leave it out of the manuscript.
+However, since these points require comparing population-level amplitudes to single-cell level amplitudes (which may not be directly comparable), we decided not to include it in the manuscript.
 
@@
 From the datasets we analyze, damping appears to be independently altered (from period, amplitude) by genetic perturbations, and therefore likely provides new insight into the underlying oscillator.
-Our main message is that the damping rate is due to cellular noise, but that the damping rate contains information about the oscillator which is currently going unused.
+Our main message is not that the damping rate is due to cellular noise, but that the damping rate contains information about the oscillator that is currently going unexplored.
 
@@
 As in all modeling work, we must make simplifying assumptions that may be drawn from incomplete experimental data.
-In developing our approach for inferring noise properties from population-level data, we have invoked two main assumptions: that individual cells are both independent and sustained oscillators.
+In developing our approach for inferring noise properties from population-level data, we have invoked two main assumptions: that individual cells are independent, and that oscillations in each cell are sustained.
 We are careful to list these assumptions in the introduction, as we agree that the validity of our results rests upon whether or not these assumptions hold in the systems under consideration.
-However, to the best of our knowledge, there is relatively strong experimental support to both of these assumptions in cultured circadian reporter cells.
-In the earlier parts of our paper, we are demonstrating that single-cell noise is {\itshape sufficient} to explain the observed damping behavior, and indeed quantitatively accurate predictions for population-level damping rate are obtained from a simple model not fit to the noise properties.
+However, to the best of our knowledge, there is relatively strong experimental support for both of these assumptions in cultured circadian reporter cells.
+
+In the other parts of our paper, we are demonstrating that single-cell noise is {\itshape sufficient} to explain the observed damping behavior - through both single-cell data and mathematical modeling.
+{\bfseries We have updated the language substantially throughout the manuscript in endeavoring to make this point more clear.}
 
@@
 
-We thank the reviewer for pointing out this error. We have corrected our usage throughout the manuscript.
+We thank the reviewer for pointing out this error; {\bfseries we have corrected our usage throughout the manuscript.}\\[3ex]
 

### 2015-05-06 main_text.tex
moving methods after results
Removing graphics from the manuscript
Moving conclusion, renaming sections
final edits prior to sending to frank

@@
 
+\section*{Results}
+
+\subsection*{Higher noise results in faster damping in population-level rhythms}
+
+While both intrinsic and extrinsic noise sources can contribute to population-level damping, intrinsic noise is thought to play a more significant role in circadian systems \cite{Herzog2004}.
+We therefore first sought to determine whether changes to intrinsic stochastic noise alone are sufficient to explain population-level changes in damping rate.
+To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
+Cells were sorted into two groups, a low-noise group and a high-noise group, based on the relative high-frequency noise, period variability, and amplitude variability present in each trace.
+Example rhythms from cells in both groups are shown in Figure~\ref{fig:fibroblast_noise}A.
+Because the cells were not synchronized at the start of the recording, this effect is replicated {\itshape in silico} by shifting each series in time to align their start phases.
+Population-level bioluminescence traces were then found by averaging the cellular PER2::LUC signal in each group.
+Both populations displayed averaged rhythms that resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures.
+Fitting the averaged expression of each group with a damped sinusoid revealed that the low-noise group also had a lower damping rate (Figure~\ref{fig:fibroblast_noise}B).
+The significance of this difference was confirmed via a bootstrap analysis (Figure~\ref{fig:fibroblast_noise}C), where cells were randomly assigned in each bootstrap trial to either the low-noise or the high-noise group.
+
+\begin{figure}[tbp]
+  \begin{center}
+    % \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig1.pdf}
+  \end{center}
+  \caption{{\bfseries Single-cell bioluminescence recordings show that higher stochastic noise results in faster damping at the population-level.}
+  Data on the bioluminescence of single-cell fibroblasts was taken from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
+({\bfseries A}) Cells were sorted into two groups depending on their degree of stochastic noise.
+An example trace from each of the two groups is shown, demonstrating different levels of noise present in the dataset.
+({\bfseries B}) After artificially synchronizing each cell, we calculate averaged bioluminescence rhythms of each group (solid lines).
+A damped sinusoid fit to both groups reveals a difference in damping rate, demonstrated by fitted envelope functions ($\pm A\exp{-dt}$, dashed lines).
+({\bfseries C}) The observed absolute difference in damping rate was shown to be significant ($p = 0.0264$) using $10,000$ bootstrap trials.}
+\label{fig:fibroblast_noise}
+\end{figure}
+
+\subsection*{Clock perturbations can change single-cell stochastic noise}
+
+\subsubsection*{A small molecule causes dose-dependent changes in the population-level damping rate of cultured cells}
+
+We next demonstrate that perturbations to the transcriptional oscillator are capable of altering population-level damping rate.
+The actions of small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
+By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence},~\ref{fig:small_molecule_r2}).
+This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc} U2OS cells), indicating it is a fundamental property of the overall gene regulatory network.
+This result indicates that stochastic noise can be altered by perturbations known to affect the transcription-translation feedback loop.
+
+\begin{figure}[tbp]
+  \begin{center}
+    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig2.pdf}}
+  \end{center}
+  \caption{{\bfseries Small molecule modulator KL001 increases damping rate in a dose-dependent fashion.}
+  Experimental data on the dose-dependent effects of small molecules KL001 and longdaysin on cultured circadian reporter cells was taken from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.
+({\bfseries A}) Detrended bioluminescence signals from the two reporter systems and two small molecules are shown normalized by the fitted amplitude, period, and phase.
+The normalized bioluminescence highlights the dose-dependent change in damping rate seen with the KL001 application (top), but not with longdaysin (bottom).
+({\bfseries B}) Quantification of the dose-dependent change in damping rate caused by small molecule modulators.
+While both molecules lead to a dose-dependent increase in period, only KL001 shows a reliable change in damping rate.}
+\label{fig:dose_dependence}
+\end{figure}
+
+\subsubsection*{Mathematical model predicts KL001 damping rate increase comes from increased single-cell stochastic noise}
+
+We next test the hypothesis that the dose-dependent changes in damping rate from KL001 is due to changes in intrinsic (cell-autonomous) noise characteristics.
+To do this, we employed a mathematical model of circadian rhythms previously used to explain the effects of both small molecule perturbations \cite{St.John2014}, summarized in Tables \ref{tab:model}-\ref{tab:parset}.
+In order to capture changes to noise characteristics, we first converted the model to a stochastic biochemical system.
+Population-level rhythms were generated by averaging the trajectories of $1,000$ individual, noninteracting oscillators.
+The only free parameter in converting the existing deterministic model to a stochastic one is the cell volume, which was determined by fitting the observed population-level damping rate to that of the experimental control traces (Figure~\ref{fig:vol_calibration}).
+
+The model was then used to predict the effects of KL001 and longdaysin administration on single-cell noise and population damping rates (Figure~\ref{fig:simulation}A).
+Reductions in parameters previously attributed to the activities of each small molecule caused dose-dependent changes in period and damping rate at the population-level that closely matched experimental results (Figure~\ref{fig:simulation}B).
+As the model includes no cell-to-cell communication or heterogeneity, this difference is manifested solely by changing the noise characteristics of individual cells.
+This quantitative prediction by the mathematical model lends further support to the assumptions that individual cells are independent and sustained oscillators, and intrinsic noise plays the dominant role in determining population-level damping.
+
+\begin{figure}[tbp]
+  \begin{center}
+    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig3.pdf}}
+  \end{center}
+  \caption{{\bfseries Mathematical model accurately predicts dose-dependent changes in damping rate.}
+({\bfseries A}) Example single-cell trajectories (top) and population-averaged trajectories (bottom, mean of $1,000$ cells) of cells under various treatments.
+Cells with the nominal parameter set (left, black) closely match the experimental damping rate for unperturbed cells.
+Cells with simulated KL001 action (red, center) are noisier at the single-cell level, and show faster damping at the population-level.
+Cells with simulated longdaysin action show slightly more accurate single-cell rhythms, with a corresponding reduction in the population-level damping rate.
+({\bfseries B}) The model accurately predicts the general trend of period vs.\ damping rate for both KL001 and longdaysin perturbations.
+Experimental data points represent the mean of two replications at each concentration.
+Computational data points represent the mean of ten independent population simulations.}
+\label{fig:simulation}
+\end{figure}
+
+\subsection*{Genome-wide effects of siRNA knockdown on single-cell stochastic noise}
+
+Unlike from using single-cell imaging, inferring stochastic noise from the desynchronization rate of population-level recordings can be applied to existing and future high-throughput circadian screens.
+We demonstrate the insights that can be gained from such an approach by analyzing the publicly available genome-wide siRNA screen from Zhang {\itshape et al.,} 2009 \cite{Zhang2009}.
+The results of fitting a damped sinusoid to each of the $111,743$ bioluminescence trajectories is shown in Figure~\ref{fig:fit_quality}, in which $86\%$ of fits had an $R^2 > 0.8$.
+Since sinusoidal parameters can only be confidently inferred for fits in which the $R^2$ was sufficiently high, wells were removed from further analysis if $R^2 < 0.8$.
+Additionally, of the fits with a high $R^2$ value, only a small minority ($0.1\%$) had a negative damping rate.
+This trend supports the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
+
+\begin{figure}[tbp]
+  \begin{center}
+    % \includegraphics[]{figures/pdfs/Fig4.pdf}
+  \end{center}
+  \caption{{\bfseries Fit quality vs.\ damping rate for the genome-wide siRNA screen.}
+({\bfseries A}) A plot of the $111,743$ individual fits shows that the majority of fits have a high $R^2$ value and positive damping rate. Only fits with $R^2 > 0.8$ were kept for further analysis.
+({\bfseries B}) Examples chosen randomly from each of the four quadrant regions in (A). Sinusoidal parameters for fits 1-2 can be more confidently inferred than those for fits 3-4.}
+\label{fig:fit_quality}
+\end{figure}
+
+We next checked how parameters varied on a plate-to-plate and well-to-well basis.
+Well-to-well variation was relatively absent, aside from expected variation due to long- and short-period controls (Figure~\ref{fig:well_variation}).
+Fits were normalized to remove plate-to-plate variation (Figure~\ref{fig:plate_variation}) using a robust z-score \cite{Birmingham2009}.
+Additionally, we separated wells into a ``perturbed'' category and ``control'' category, depending on whether or not the well contained an siRNA perturbation.
+As we show in Figure~\ref{fig:fit_distributions}, all fitted parameters displayed normal-like distributions, in which the control distributions showed much tighter clustering around the most likely values.
+Quantifications of the mean, variance, skewness, and kurtosis for each distribution are shown in Table~\ref{tab:fit_distributions}.
+
+\begin{figure}[tbp]
+  \begin{center}
+    % \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig5.pdf}
+  \end{center}
+  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} For each well in the high-throughput screen, the period, amplitude and damping rate are calculated.
+After normalization, distributions in robust z-scores closely resemble normal distributions.
+For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.}
+\label{fig:fit_distributions}
+\end{figure}
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{lrrrrrr}\toprule
+      {} & \multicolumn{2}{c}{$T$} & \multicolumn{2}{c}{$\ln A$} & \multicolumn{2}{c}{$d$} \\
+      {}         & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}               & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           \\\midrule
+      Mean     & $-0.234$ & $0.187$ & $0.443$  & $-0.343$ & $0.043$  & $0.090$    \\
+      Variance & $ 0.599$ & $3.313$ & $0.605$  & $ 3.074$ & $0.771$  & $2.850$    \\
+      Skewness & $ 0.153$ & $0.367$ & $-1.823$ & $-0.580$ & $-0.107$ & $0.371$    \\
+      Kurtosis & $ 3.772$ & $0.591$ & $8.329$  & $ 0.476$ & $2.423$  & $0.373$    \\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+  \caption{{\bfseries Moments of the fitted parameter distributions (Figure~\ref{fig:fit_distributions}) after normalization and outlier removal.} Parameters were normalized by subtracting the median and dividing by the median absolute deviation on a plate-by-plate basis.
+Control distributions had less variance, were less skewed, and were more peaked then their perturbed counterparts.}
+  \label{tab:fit_distributions}
+\end{table}
+
+\subsubsection*{Damping rate is independent of other sinusoidal parameters}
+
+Since an siRNA's effect on period length does not effectively predict its effect on amplitude or phase (and vice-versa), we hypothesized that damping rate would similarly be independently affected by siRNA perturbations.
+Low Pearson correlation coefficients between normalized parameter distributions defend this hypothesis, with the highest correlations among variables seen between amplitude and damping rate ($\rho = 0.285$, Table~\ref{tab:corr}).
+Additionally, a multivariate linear regression of damping rate as a function of period, amplitude, phase, and perturbation type (control or perturbed, a categorical variable) produced an $R^2$ value of only $0.169$ (Table~\ref{tab:ols_reg}).
+These results reinforce the notion that the population-level damping rate describes an independent feature of the underlying oscillator.
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{rrrrr}
+      \toprule
+      {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
+      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\
+      $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
+      $T$      & $-0.142$ & $-0.022$  & $1.000 $ & $-0.113$\\
+      $\theta$ & $-0.269$ & $-0.112$  & $-0.113$ & $1.000 $\\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+  \caption{{\bfseries Correlation among normalized parameters of the high-throughput siRNA screen.} Pearson correlation coefficients are relatively low between fitted parameters, indicating that changes to damping rate (and thereby stochastic noise) are not explained by changes to period, amplitude, or phase.}
+  \label{tab:corr}
+\end{table}
+
+\subsubsection*{The unperturbed clock lies at the Pareto frontier of high amplitude and low noise}
+
+The ability of a population of oscillators to maintain robust oscillations is a function both of its initial amplitude as well as its damping rate.
+A robust oscillator is therefore one with high initial amplitude and low stochastic noise.
+Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate that few outlier points reside in the lower damping rate, higher amplitude quadrant.
+In order to find perturbations that confidently change robustness, we grouped the siRNA perturbations by target gene and performed a two-sample Hotelling's $T^2$ test against the control population.
+The resulting significant gene perturbations ($75\%$ of all genes) are shown in Figure~\ref{fig:outlier_dist}C.
+Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations ($3.3\%$) simultaneously increase amplitude and decrease stochastic noise.
+It therefore appears that the unperturbed clock optimizes some trade-off between high amplitude and low stochastic noise, such that it is unlikely the knockdown of a single gene will make oscillations more robust.
+
+\begin{figure}[tbp]
+  \begin{center}
+    % \includegraphics[width=0.65\textwidth]{figures/pdfs/Fig6.pdf}
+  \end{center}
+  \caption{{\bfseries Effects of siRNA knockdowns on amplitude and damping rate.}
+  Clock robustness is a function of both amplitude and damping rate.
+  Distributions in amplitude and damping rate for control wells ({\bfseries A}) or perturbed wells ({\bfseries B}) indicate that perturbations tend to shift the clock towards regions of higher damping rate or lower amplitude.
+Green dots in each figure indicates the mean of the control population.
+({\bfseries C}) Averaged effect of siRNA knockdown after grouping the perturbed population by Gene ID.
+Only those genes that were significantly different from the control distribution are shown (Hotelling's $T^2$ test, $\alpha = 0.01$).
+({\bfseries D}) Radial histogram of the significant gene perturbations shown in (C).
+The area of each slice is proportional to the frequency of perturbations away from the mean in that direction.
+Very few gene knockdowns result in both higher amplitude and lower damping rate (red slices, lower right quadrant).
+}
+\label{fig:outlier_dist}
+\end{figure}
+
+\subsection*{Effect of cell heterogeneity}
+
+In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping.
+However, experimental work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods \cite{Leise2012}.
+Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
+While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise likely plays a more significant role: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}).
+This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise \cite{Herzog2004}.
+
+However, it is possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
+Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics.
+While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure~\ref{fig:dephasing-rates}), these differences are likely not identifiable from typical data (Figure~\ref{fig:computational_dephasing}).
+Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
+Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.
+
+\section*{Discussion}
+
+In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
+An overview of the computational steps involved in this method are outlined in Figure~\ref{fig:comp_overview}.
+While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
+Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
+Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
+While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
+We also note that there are many available tools for detrending and regressing time-series data.
+While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.
+
+As high amplitude circadian oscillations are important for maintaining metabolic health, many studies have sought to find small molecule candidates that increase oscillatory amplitude \cite{Chen2013}.
+In the search for circadian clock therapeutics, high-throughput methods are frequently used to screen for such drugs, often neglecting potential effects on cell-autonomous noise.
+While we have demonstrated a method by which the effects of small molecules on noise can be inferred from high-throughput methods, we have also shown that the potential improvement of clock robustness may be limited.
+Amplitudes of circadian rhythms may therefore be best increased by small molecule therapies that act transiently to synchronize peripheral oscillators.
+While such a method would require accurate alignment of drug administration to the correct circadian phase, a recent {\itshape in silico} study has demonstrated the potential effectiveness of such an approach in improving amplitudes in peripheral tissues \cite{St.John2014a}.
+Finally, the ability to extract an additional biologically relevant parameter from existing datasets will likely prove useful for many studies, as it allows further differentiation between perturbations that might otherwise have identical effects on bioluminescence rhythms.
+
 \section*{Materials and Methods}
@@
 Amplitude, $A$, and phase, $\theta$, parameters are subsequently fit using a linear least-squares regression.
-Note that in this manuscript we use amplitude to denote the initial rhythm strength, and damping rate to denote the rate at which this strength decays with time.
+Note that in this manuscript we use {\itshape amplitude} to denote the initial rhythm strength, and {\itshape damping rate} to denote the rate at which this strength decays with time.
 Overall $R^2$ values for the regression were calculated from the residual error between the detrended data and fitted sinusoid:
@@
 As in the original study, various parameters describing the average noise level of each cell were collected.
-Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components -- resulting in rhythms which only contained oscillations with periods between 8 and 258 hours.
+Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components -- resulting in rhythms that only contained oscillations with periods between 8 and 258 hours.
 From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
@@
 We removed outlier points prior to calculating the moments of the distributions, Pearson correlation coefficients, and performing the multivariable linear regression.
-Outliers were defined as points which contained a z-metric (in either period, phase, amplitude, or damping rate) with an absolute value greater than eight.
-We chose the ``control'' wells to be those which contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
+Outliers were defined as points that contained a z-metric (in either period, phase, amplitude, or damping rate) with an absolute value greater than eight.
+We chose the ``control'' wells to be those that contained no siRNA, as these proved to be more numerous than those containing reference siRNA perturbations and were clustered similarly to the highest-density regions of the perturbed fits.
 
@@
 Code used to perform the analysis and produce the figures in this manuscript can be found online at \url{https://github.com/pstjohn/decay-methods}.
-
-\section*{Results and Discussion}
-
-\subsection*{Higher noise results in faster damping in population-level rhythms}
-
-While both intrinsic and extrinsic noise sources can contribute to population-level damping, intrinsic noise is thought to play a more significant role in circadian systems \cite{Herzog2004}.
-We therefore first sought to determine whether changes to intrinsic stochastic noise alone are sufficient to explain population-level changes in damping rate.
-To do this, we calculated noise characteristics from experimental data on individual PER2::LUC fibroblast cells \cite{Leise2012}.
-Cells were sorted into two groups, a low-noise group and a high-noise group, based on the relative high-frequency noise, period variability, and amplitude variability present in each trace.
-Example rhythms from cells in both groups are shown in Figure~\ref{fig:fibroblast_noise}A.
-Because the cells were not synchronized at the start of the recording, this effect is replicated {\itshape in silico} by shifting each series in time to align their start phases.
-Population-level bioluminescence traces were then found by averaging the cellular PER2::LUC signal in each group.
-Both populations displayed averaged rhythms which resembled a damped sinusoid, similar to those seen in bioluminescence recordings of entire cell cultures.
-Fitting the averaged expression of each group with a damped sinusoid revealed that the low-noise group also had a lower damping rate (Figure~\ref{fig:fibroblast_noise}B).
-The significance of this difference was confirmed via a bootstrap analysis (Figure~\ref{fig:fibroblast_noise}C), where cells were randomly assigned in each bootstrap trial to either the low-noise or the high-noise group.
-
-\begin{figure}[tbp]
-  \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig1.pdf}
-  \end{center}
-  \caption{{\bfseries Single-cell bioluminescence recordings show that higher stochastic noise results in faster damping at the population-level.}
-  Data on the bioluminescence of single-cell fibroblasts was taken from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
-({\bfseries A}) Cells were sorted into two groups depending on their degree of stochastic noise.
-An example trace from each of the two groups is shown, demonstrating different levels of noise present in the dataset.
-({\bfseries B}) After artificially synchronizing each cell, we calculate averaged bioluminescence rhythms of each group (solid lines).
-A damped sinusoid fit to both groups reveals a difference in damping rate, demonstrated by fitted envelope functions ($\pm A\exp{-dt}$, dashed lines).
-({\bfseries C}) The observed absolute difference in damping rate was shown to be significant ($p = 0.0264$) using $10,000$ bootstrap trials.}
-\label{fig:fibroblast_noise}
-\end{figure}
-
-\subsection*{Clock perturbations can change single-cell stochastic noise}
-
-\subsubsection*{A small molecule causes dose-dependent changes in the population-level damping rate of cultured cells}
-
-We next demonstrate that perturbations to the transcriptional oscillator are capable of altering population-level damping rate.
-The actions of small-molecule circadian modulators KL001 and longdaysin are well-characterized, and are known to affect circadian period and amplitude in a dose-dependent fashion \cite{St.John2014}.
-By fitting experimental data on the population-level responses to increasing dosages of each molecule with a damped sinusoid, we show that KL001, but not longdaysin, increases damping rate in a dose-dependent fashion (Figure~\ref{fig:dose_dependence},~\ref{fig:small_molecule_r2}).
-This change in damping rate is consistent across both reporter systems ({\itshape Bmal1-dLuc} and {\itshape Per2-dLuc} U2OS cells), indicating it is a fundamental property of the overall gene regulatory network.
-This result indicates that stochastic noise can be altered by perturbations known to affect the transcription-translation feedback loop.
-
-\begin{figure}[tbp]
-  \begin{center}
-    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig2.pdf}}
-  \end{center}
-  \caption{{\bfseries Small molecule modulator KL001 increases damping rate in a dose-dependent fashion.}
-  Experimental data on the dose-dependent effects of small molecules KL001 and longdaysin on cultured circadian reporter cells was taken from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.
-({\bfseries A}) Detrended bioluminescence signals from the two reporter systems and two small molecules are shown normalized by the fitted amplitude, period, and phase.
-The normalized bioluminescence highlights the dose-dependent change in damping rate seen with the KL001 application (top), but not with longdaysin (bottom).
-({\bfseries B}) Quantification of the dose-dependent change in damping rate caused by small molecule modulators.
-While both molecules lead to a dose-dependent increase in period, only KL001 shows a reliable change in damping rate.}
-\label{fig:dose_dependence}
-\end{figure}
-
-\subsubsection*{Mathematical model predicts KL001 damping rate increase comes from increased single-cell stochastic noise}
-
-We next test the hypothesis that the dose-dependent changes in damping rate from KL001 is due to changes in intrinsic (cell-autonomous) noise characteristics.
-To do this, we employed a mathematical model of circadian rhythms previously used to explain the effects of both small molecule perturbations \cite{St.John2014}, summarized in Tables \ref{tab:model}-\ref{tab:parset}.
-In order to capture changes to noise characteristics, we first converted the model to a stochastic biochemical system.
-Population-level rhythms were generated by averaging the trajectories of $1,000$ individual, noninteracting oscillators.
-The only free parameter in converting the existing deterministic model to a stochastic one is the cell volume, which was determined by fitting the observed population-level damping rate to that of the experimental control traces (Figure~\ref{fig:vol_calibration}).
-
-The model was then used to predict the effects of KL001 and longdaysin administration on single-cell noise and population damping rates (Figure~\ref{fig:simulation}A).
-Reductions in parameters previously attributed to the activities of each small molecule caused dose-dependent changes in period and damping rate at the population-level that closely matched experimental results (Figure~\ref{fig:simulation}B).
-As the model includes no cell-to-cell communication or heterogeneity, this difference is manifested solely by changing the noise characteristics of individual cells.
-This quantitative prediction by the mathematical model lends further support to the assumptions that individual cells are independent and sustained oscillators, and intrinsic noise plays the dominant role in determining population-level damping.
-
-\begin{figure}[tbp]
-  \begin{center}
-    \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig3.pdf}}
-  \end{center}
-  \caption{{\bfseries Mathematical model accurately predicts dose-dependent changes in damping rate.}
-({\bfseries A}) Example single-cell trajectories (top) and population-averaged trajectories (bottom, mean of $1,000$ cells) of cells under various treatments.
-Cells with the nominal parameter set (left, black) closely match the experimental damping rate for unperturbed cells.
-Cells with simulated KL001 action (red, center) are noisier at the single-cell level, and show faster damping at the population-level.
-Cells with simulated longdaysin action show slightly more accurate single-cell rhythms, with a corresponding reduction in the population-level damping rate.
-({\bfseries B}) The model accurately predicts the general trend of period vs.\ damping rate for both KL001 and longdaysin perturbations.
-Experimental data points represent the mean of two replications at each concentration.
-Computational data points represent the mean of ten independent population simulations.}
-\label{fig:simulation}
-\end{figure}
-
-\subsection*{Genome-wide effects of siRNA knockdown on single-cell stochastic noise}
-
-Unlike from using single-cell imaging, inferring stochastic noise from the desynchronization rate of population-level recordings can be applied to existing and future high-throughput circadian screens.
-We demonstrate the insights that can be gained from such an approach by analyzing the publicly available genome-wide siRNA screen from Zhang {\itshape et al.,} 2009 \cite{Zhang2009}.
-The results of fitting a damped sinusoid to each of the $111,743$ bioluminescence trajectories is shown in Figure~\ref{fig:fit_quality}, in which $86\%$ of fits had an $R^2 > 0.8$.
-Since sinusoidal parameters can only be confidently inferred for fits in which the $R^2$ was sufficiently high, wells were removed from further analysis if $R^2 < 0.8$.
-Additionally, of the fits with a high $R^2$ value, only a small minority ($0.1\%$) had a negative damping rate.
-This trend supports the assumption that intercellular synchronization is unlikely in cultured U2OS cells.
-
-\begin{figure}[tbp]
-  \begin{center}
-    \includegraphics[]{figures/pdfs/Fig4.pdf}
-  \end{center}
-  \caption{{\bfseries Fit quality vs.\ damping rate for the genome-wide siRNA screen.}
-({\bfseries A}) A plot of the $111,743$ individual fits shows that the majority of fits have a high $R^2$ value and positive damping rate. Only fits with $R^2 > 0.8$ were kept for further analysis.
-({\bfseries B}) Examples chosen randomly from each of the four quadrant regions in (A). Sinusoidal parameters for fits 1-2 can be more confidently inferred than those for fits 3-4.}
-\label{fig:fit_quality}
-\end{figure}
-
-We next checked how parameters varied on a plate-to-plate and well-to-well basis.
-Well-to-well variation was relatively absent, aside from expected variation due to long- and short-period controls (Figure~\ref{fig:well_variation}).
-Fits were normalized to remove plate-to-plate variation (Figure~\ref{fig:plate_variation}) using a robust z-score \cite{Birmingham2009}.
-Additionally, we separated wells into a ``perturbed'' category and ``control'' category, depending on whether or not the well contained an siRNA perturbation.
-As we show in Figure~\ref{fig:fit_distributions}, all fitted parameters displayed normal-like distributions, in which the control distributions showed much tighter clustering around the most likely values.
-Quantifications of the mean, variance, skewness, and kurtosis for each distribution are shown in Table~\ref{tab:fit_distributions}.
-
-\begin{figure}[tbp]
-  \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig5.pdf}
-  \end{center}
-  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} For each well in the high-throughput screen, the period, amplitude and damping rate are calculated.
-After normalization, distributions in robust z-scores closely resemble normal distributions.
-For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.}
-\label{fig:fit_distributions}
-\end{figure}
-
-\begin{table}
-  \begin{center}
-    \begin{tabular}{lrrrrrr}\toprule
-      {} & \multicolumn{2}{c}{$T$} & \multicolumn{2}{c}{$\ln A$} & \multicolumn{2}{c}{$d$} \\
-      {}         & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}               & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           \\\midrule
-      Mean     & $-0.234$ & $0.187$ & $0.443$  & $-0.343$ & $0.043$  & $0.090$    \\
-      Variance & $ 0.599$ & $3.313$ & $0.605$  & $ 3.074$ & $0.771$  & $2.850$    \\
-      Skewness & $ 0.153$ & $0.367$ & $-1.823$ & $-0.580$ & $-0.107$ & $0.371$    \\
-      Kurtosis & $ 3.772$ & $0.591$ & $8.329$  & $ 0.476$ & $2.423$  & $0.373$    \\
-      \bottomrule
-    \end{tabular}
-  \end{center}
-  \caption{{\bfseries Moments of the fitted parameter distributions (Figure~\ref{fig:fit_distributions}) after normalization and outlier removal.} Parameters were normalized by subtracting the median and dividing by the median absolute deviation on a plate-by-plate basis.
-Control distributions had less variance, were less skewed, and were more peaked then their perturbed counterparts.}
-  \label{tab:fit_distributions}
-\end{table}
-
-\subsubsection*{Damping rate is independent of other sinusoidal parameters}
-
-Since an siRNA's effect on period length does not effectively predict its effect on amplitude or phase (and vice-versa), we hypothesized that damping rate would similarly be independently affected by siRNA perturbations.
-Low Pearson correlation coefficients between normalized parameter distributions defend this hypothesis, with the highest correlations among variables seen between amplitude and damping rate ($\rho = 0.285$, Table~\ref{tab:corr}).
-Additionally, a multivariate linear regression of damping rate as a function of period, amplitude, phase, and perturbation type (control or perturbed, a categorical variable) produced an $R^2$ value of only $0.169$ (Table~\ref{tab:ols_reg}).
-These results reinforce the notion that the population-level damping rate describes an independent feature of the underlying oscillator.
-
-\begin{table}
-  \begin{center}
-    \begin{tabular}{rrrrr}
-      \toprule
-      {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
-      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\
-      $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
-      $T$      & $-0.142$ & $-0.022$  & $1.000 $ & $-0.113$\\
-      $\theta$ & $-0.269$ & $-0.112$  & $-0.113$ & $1.000 $\\
-      \bottomrule
-    \end{tabular}
-  \end{center}
-  \caption{{\bfseries Correlation among normalized parameters of the high-throughput siRNA screen.} Pearson correlation coefficients are relatively low between fitted parameters, indicating that changes to damping rate (and thereby stochastic noise) are not explained by changes to period, amplitude, or phase.}
-  \label{tab:corr}
-\end{table}
-
-\subsubsection*{The unperturbed clock lies at the Pareto frontier of high amplitude and low noise}
-
-The ability of a population of oscillators to maintain robust oscillations is a function both of its initial amplitude as well as its damping rate.
-A robust oscillator is therefore one with high initial amplitude and low stochastic noise.
-Scatter plots of the control and perturbed distributions are shown in Figure~\ref{fig:outlier_dist}A-B, which indicate that few outlier points reside in the lower damping rate, higher amplitude quadrant.
-In order to find perturbations which confidently change robustness, we grouped the siRNA perturbations by target gene and performed a two-sample Hotelling's $T^2$ test against the control population.
-The resulting significant gene perturbations ($75\%$ of all genes) are shown in Figure~\ref{fig:outlier_dist}C.
-Quantifying the distribution in outliers by quadrant, it is clear that only a small minority of perturbations ($3.3\%$) simultaneously increase amplitude and decrease stochastic noise.
-It therefore appears that the unperturbed clock optimizes some trade-off between high amplitude and low stochastic noise, such that it is unlikely the knockdown of a single gene will make oscillations more robust.
-
-\begin{figure}[tbp]
-  \begin{center}
-    \includegraphics[width=0.65\textwidth]{figures/pdfs/Fig6.pdf}
-  \end{center}
-  \caption{{\bfseries Effects of siRNA knockdowns on amplitude and damping rate.}
-  Clock robustness is a function of both amplitude and damping rate.
-  Distributions in amplitude and damping rate for control wells ({\bfseries A}) or perturbed wells ({\bfseries B}) indicate that perturbations tend to shift the clock towards regions of higher damping rate or lower amplitude.
-Green dots in each figure indicates the mean of the control population.
-({\bfseries C}) Averaged effect of siRNA knockdown after grouping the perturbed population by Gene ID.
-Only those genes which were significantly different from the control distribution are shown (Hotelling's $T^2$ test, $\alpha = 0.01$).
-({\bfseries D}) Radial histogram of the significant gene perturbations shown in (C).
-The area of each slice is proportional to the frequency of perturbations away from the mean in that direction.
-Very few gene knockdowns result in both higher amplitude and lower damping rate (red slices, lower right quadrant).
-}
-\label{fig:outlier_dist}
-\end{figure}
-
-\subsection*{Effect of cell heterogeneity}
-
-In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping.
-However, experimental work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods \cite{Leise2012}.
-Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
-While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise likely plays a more significant role: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}).
-This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise \cite{Herzog2004}.
-
-However, it is possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
-Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics.
-While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure~\ref{fig:dephasing-rates}), these differences are likely not identifiable from typical data (Figure~\ref{fig:computational_dephasing}).
-Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
-Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.
-
-% % Additionally, it is important to note that the dephasing kinetics of these two alternative processes result in different dynamics at the population-level.
-% In dephasing due solely to stochastic noise, the probability density of finding a cell with any given phase follows a standard convection-diffusion regime, resulting in exponentially damped rhythms \cite{St.John2014a}.
-% In dephasing due solely to period heterogeneity, the variance of the phase probability density broadens in time proportionally to $t^2$, rather than $t$ in standard diffusion.
-% This type of anomalous diffusion results in a different damping profile, as highlighted in Figure~\ref{fig:dephasing-rates}.
-% Since the data used in this study is well-described by an exponentially damped sinusoid, and since a mathematical model assuming homogeneous mean periods accurately predicts experimental results, we expect our assumption that changes to intrinsic noise are primarily responsible for the observed damping rate changes to be justified.
-% 
-% However, it is possible that damping rate changes due to siRNA or small molecule perturbation are manifested by altering the system's extrinsic noise.
-% Such a change, caused perhaps by an unequal uptake of drug across cells, would result in an altered distribution in mean period length, and different dephasing kinetics.
-% A simple model of a stochastic oscillator was used to investigate the effects of altering intrinsic and extrinsic noise on the dephasing kinetics (Figure~\ref{fig:computational_dephasing}).
-% While {\itshape in silico} dephasing due to solely intrinsic or extrinsic sources can be differentiated from population-level rhythms, it becomes more difficult to identify noise sources when multiple types of noise are present.
-% The window function of population-level oscillations can be affected by many factors, including weak cell-to-cell coupling \cite{Rougemont2007}, temporary amplitude change \cite{St.John2014a}, and intrinsic and extrinsic noise.
-% While altering intrinsic, or cell-autonomous, noise has been shown to be sufficient to explain the observed damping rate changes, it is possible that a portion of the changes to damping rate observed in regressed experimental data may be due to changes in extrinsic noise.
-
-\section*{Conclusion}
-
-% Our method assumes independent and sustained oscillators at the single-cell level.
-% While these assumptions have not always been upheld in the circadian literature \cite{Izumo2003}, more recent studies have demonstrated the validity of these assumptions in cultured fibroblast cells.
-% Earlier mathematical studies have analysed on the damping rate of oscillators without assuming independent or homogeneous populations \cite{Rougemont2007}.
-% However, such approaches introduce many additional fitted parameters, limiting their applicability to high-throughput screens.
-
-In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
-An overview of the computational steps involved in this method are outlined in Figure~\ref{fig:comp_overview}.
-While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
-Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
-Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
-While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
-We also note that there are many available tools for detrending and regressing time-series data.
-While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.
-
-As high amplitude circadian oscillations are important for maintaining metabolic health, many studies have sought to find small molecule candidates which increase oscillatory amplitude \cite{Chen2013}.
-In the search for circadian clock therapeutics, high-throughput methods are frequently used to screen for such drugs, often neglecting potential effects on cell-autonomous noise.
-While we have demonstrated a method by which the effects of small molecules on noise can be inferred from high-throughput methods, we have also shown that the potential improvement of clock robustness may be limited.
-Amplitudes of circadian rhythms may therefore be best increased by small molecule therapies which act transiently to synchronize peripheral oscillators.
-While such a method would require accurate alignment of drug administration to the correct circadian phase, a recent {\itshape in silico} study has demonstrated the potential effectiveness of such an approach in improving amplitudes in peripheral tissues \cite{St.John2014a}.
-Finally, the ability to extract an additional biologically relevant parameter from existing datasets will likely prove useful for many studies, as it allows further differentiation between perturbations which might otherwise have identical effects on bioluminescence rhythms.
 
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/pdfs/FigS1.pdf}
+    % \includegraphics[width=0.5\textwidth]{figures/pdfs/FigS1.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.5\textwidth]{figures/pdfs/FigS2.pdf}
+    % \includegraphics[width=0.5\textwidth]{figures/pdfs/FigS2.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS3.pdf}
+    % \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS3.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS4.pdf}
+    % \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS4.pdf}
   \end{center}
   \caption{{\bfseries Well-to-well variation in the Zhang {\itshape et al.,} 2009 genome-wide siRNA screen.} Similar to Figure~\ref{fig:plate_variation}, dots indicate the median value of each parameter in each well, with lines extending from the $5^\text{th}$ to $95^\text{th}$ percentile.
-Well position did not seem to affect the fitted values, particularly in the middle regions which contained the siRNA knockdown library.
+Well position did not seem to affect the fitted values, particularly in the middle regions that contained the siRNA knockdown library.
 Wells on either end showed significant variation, but these are likely due to the spotting of long- and short-period controls in the same position on each plate.}
@@
   \begin{center}
-    \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS7.pdf}
+    % \includegraphics[width=0.7\textwidth]{figures/pdfs/FigS7.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=\textwidth]{figures/pdfs/FigS8.pdf}
+    % \includegraphics[width=\textwidth]{figures/pdfs/FigS8.pdf}
   \end{center}
@@
 \end{center} 
-\caption{{\bfseries Multivariable linear regression results.} Fit statistics which demonstrate the effect of each other fitted parameter on damping rate.
+\caption{{\bfseries Multivariable linear regression results.} Fit statistics that demonstrate the effect of each other fitted parameter on damping rate.
 Of particular note is the perturbation type categorical variable, which demonstrates that the presence of siRNA perturbation increases damping rate on average, controlling for changes in other variables.

### 2015-05-20 response_letter.tex
Frank's comments

@@
 {\itshape Douglas Lauffenburger}\\
-Associate Editor\\
+Deputy Editor\\
 PLOS Computational Biology\\
@@
 
-These concerns were also shared by the first reviewer, and we have therefore {\bfseries substantially changed the language of the study to reflect the fact that extrinsic noise (period heterogeneity) could play a role in the damping rates.}
+These concerns were also shared by the first reviewer, and we have therefore {\bfseries substantially changed the language of the study to reflect the fact that extrinsic noise (period heterogeneity) could play a role in the damping rates} (see page 3 of this response).
 For Fig. 3A, these profiles were obtained from a mathematical model of homogeneous cells.
@@
 
-{\bfseries We have largely re-worded this section of the text} in response to the comments of other reviewers.
+{\bfseries We have largely re-worded this section of the text} in response to the comments of other reviewers (see page 3 of this response).
 Despite this, we have added a further citation to Rougement \& Naef, as they present a more detailed theoretical analysis of the windowing function:
@@
 
-We sadly don't have access to the raw data of Liu et al.\ to look more closely at their observations.
+We unfortunately do not have access to the raw data of Liu et al.\ to look more closely at their observations.
 In the genome-wide siRNA screen, however, Cry2 knockdown was one of the several positive controls run in each plate.

### 2015-06-30 response_letter.tex
adding review comments to response

--- 2015-05-06 response_letter.tex
@@
 \usepackage{tcolorbox}
-\tcbuselibrary{breakable}
-\tcbuselibrary{skins}
+% \tcbuselibrary{breakable}
+% \tcbuselibrary{skins}
 \usepackage{units}
@@
 
-\tcbset{skin=enhanced}
+% \tcbset{skin=enhanced}
 
@@
 {\itshape Douglas Lauffenburger}\\
-Associate Editor\\
+Deputy Editor\\
 PLOS Computational Biology\\
 
-Thank you for your email on April 20th, 2015, inviting us to submit a revised version of our manuscript, ``Quantifying stochastic noise in cultured circadian reporter cells'' (MS: PCOMPBIOL-D-15-00431) by Peter C.
+Thank you for your email on June 23th, 2015, inviting us to submit a further revised version of our manuscript, ``Quantifying stochastic noise in cultured circadian reporter cells'' (MS: PCOMPBIOL-D-15-00431) by Peter C.
 St.
@@
 
-We thank the reviewers for their detailed comments of the manuscript and suggestions for how to improve the work.
-We believe addressing these concerns and suggestions have greatly improved the manuscript.
+We are pleased to hear that two of the reviewers recommend our work for publication.
+We thank the reviewers for their second round of thoughtful critiques, which we feel further improves the quality of the manuscript.
 
@@
 \begin{reviewer}
-The premise of the manuscript is quite interesting and pertinent to the study of the circadian clock and of biological oscillators more generally. 
-Under the assumptions that the individual oscillators are self-autonomous with non-damping amplitudes and do not exert any influence on each other (which can be reasonable in certain types of dissociated cell cultures), the population-level damping rate is related to the level of stochastic noise in the individual oscillators.
-Dose-dependent effects of small molecule modulators on the noise in the clock can then be studied via the damping rate of the population, which is much easier than trying to record individual oscillators to measure such effects.
-The authors' further analysis leads them to the intriguing conjecture that the circadian clock has achieved an optimal balance between amplitude and stochastic noise.
+The revision improved the manuscript, but a couple of minor issues remain to be addressed.
 
-Overall the approach is appealing and seems promising, with plausible future applications. However, some issues require further attention to ensure the method is effective and sound.
-
-1) What exactly is meant in the manuscript by ``stochastic noise'' is not entirely clear.
-It would be helpful if the authors explained what is being covered under this term: Extrinsic noise?
-Intrinsic noise?
-Only noise generated within each cell by the molecular mechanism, or larger scale effects as well?
-What specifically are the sources of the noise under consideration, and what are the consequences for the individual oscillators' amplitude, phase, and period (and variability thereof)?
+The added discussion on types of noise should be revised to include the work of Elowitz et al (Science 2002) on extrinsic vs intrinsic noise.
+The Elowitz study does quantitatively distinguish extrinsic vs intrinsic noise in their experiments, while the revised manuscript states this is only possible in theory (so the wording should be altered at the bottom of page 6 and the Elowitz study cited).
 \end{reviewer}
 
-We thank the reviewer for their overall positive reception of the manuscript. 
-In regards to their first critique, we mainly are considering intrinsic noise generated by the molecular mechanism.
-In the study, we demonstrate that single-cell noise is {\itshape sufficient} to explain the observed changes in population damping.
-While sources of extrinsic noise are certainly present, experimental studies (i.e., Herzog {\itshape et al., JBR} 2004) have largely concluded that period lengths of circadian oscillations are more variable at the single-cell level (i.e., peak-to-peak time) than at the population level (differences in mean period between cells).
-We agree that additional discussion of this point in manuscript is necessary, and have therefore expanded the introduction to include an explicit description of the different sources and effects of stochastic noise:
-
-\begin{manuscript}{Page 3}
-Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells. 
-Transcription at the single-cell level is strongly affected by {\bfseries {\itshape intrinsic} cellular noise, caused by} the low molecular counts of the mRNA and protein species involved.
-As a result, bioluminescence traces of individual cells are stochastic, with significant variability in both amplitude and period length from {\bfseries cycle to cycle (14).
-In addition to intrinsic noise, circadian oscillations are also affected by {\itshape extrinsic} noise sources.
-Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude.
-For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells (2).
-Both sources of noise have an effect on population-level rhythms:} in cell cultures that lack cell-to-cell coupling, {\bfseries it has been shown that stochastic noise} is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase (14, 15).
-
-\end{manuscript}
-
-Additionally, {\bfseries we have updated the language throughout the manuscript} to use the intrinsic and extrinsic vocabulary to refer to noise generated within the cell or outside the cell, respectively.
+\begin{reviewer}
+Also, in the revised part of the introduction addressing the role of noise in characterizing oscillations, the work by Bratsun et al (PNAS 2005) should be mentioned as another caution.
+For instance, they point out that ``the 'noise signature' of a system dominated by delay can be similar to a system dominated by intrinsic noise.
+This result implies that care must taken in attributing variability to purely stochastic sources, because delay-induced variability can appear empirically similar to fluctuations arising from intrinsic noise.''
+\end{reviewer}
 
 \begin{reviewer}
-2) Damping of the population level oscillation is typically caused by two main factors: heterogeneity in period among the oscillators and stochastic variation in cycle length from cycle to cycle in each oscillator due to noise.
-The authors should address how these two factors contribute to the damping rate and how to distinguish them. 
-For instance, what if a clock modulator is causing increased variability in period across the population, rather than increased stochastic noise?
-That is, the period of certain oscillators could be changed more than that of other oscillators (which could occur given that the system is highly nonlinear); you could even have the average period remain stable but some oscillators have their periods shortened by the treatment while others are lengthened.
-This would increase the damping rate, but not be related to a change in stochastic noise. 
-A set of noise-free oscillators with heterogeneous periods will also damp at the population level, and damp faster if the variance in the period is increased.
+In the new expanded paragraph discussing the effect of cell period heterogeneity, the authors dismiss the significance of intrinsic differences in period across the population compared to cycle-to-cycle fluctuations and don’t directly quantify the relative contributions.
+However, Cohen et al (Journal Theoretical Biology 2013) quantified it for this same set of fibroblasts using a Bayesian hierarchical model and found the standard deviation of periods across the population to be 0.89h compared to the within-cell variability 1.43h.
+So while cycle-to-cycle fluctuations are indeed larger, the variability across the population is still quite substantial, and these values should be mentioned in this section to quantify the relative contributions of these sources of variability to the overall decay rate.
 \end{reviewer}
-
-This is an valid concern, and one which was raised by multiple reviewers.
-In the original manuscript we mentioned that intrinsic (cycle-to-cycle variability) and extrinsic (period heterogeneity) sources could be differentiated based on their envelope functions.
-In response to the reviews, we have performed additional simulations to determine whether intrinsic or extrinsic noise sources are identifiable, presented in Figure S7.
-We conclude that while pure intrinsic vs.\ pure extrinsic dephasing can be distinguished, it becomes much more difficult when both sources of noise are present.
-
-While this revision does correct a statement in our previous version, we note that main parameter of biological interest is the overall stochastic noise from both sources, which is still being identified by the method.
-We have therefore updated the corresponding section of our manuscript to discuss the possibility and implications of changes to extrinsic noise:
-
-\begin{manuscript}{Page 8}
-  \section*{Effect of cell heterogeneity}
-  {\bfseries
-In the preceding sections, we have demonstrated that changes to single-cell intrinsic noise are sufficient to explain the observed changes in population-level damping. 
-However, experimental} work has shown that cell-autonomous fibroblast cells have a distribution of mean free-running periods (21). 
-Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping (30). 
-While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, {\bfseries intrinsic stochastic noise likely plays a more significant role}: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis {\bfseries in cultured fibroblast cells (Figure S5).
-This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise (2).
-
-However, it is possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise. 
-Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics. 
-While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure S6), these differences are likely not identifiable from typical data (Figure S7).
-Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors. 
-Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.}
-\end{manuscript}
-
-\begin{reviewer}
-The authors do have a paragraph addressing the effect of heterogeneity, but they seem to assume the heterogeneity will remain unchanged by the clock modulators.
-However, looking at Table 1, this assumption doesn't hold.
-The variance is enormously increased by the perturbation, and one would expect this increased variability is likely occurring at the single cell level as well.
-\end{reviewer}
-
-Indeed, the possibility that period heterogeneity may be changed by clock modulators has been added as discussed above.
-The increase in variance in Table 1, however, is most likely due to the variety of genes targeted by the siRNA library.
-At the single-cell level, each cell in the well would experience the same siRNA perturbation, which we suspect would lead to a similar change in period / amplitude.
-This is echoed by the consistency seen in the control wells.
-As this was perhaps a misunderstanding of the figure, we have added the following text to the caption of Figure 5:
-
-\begin{manuscript}{Page 7}
-  Figure 5: Distributions in fitted parameters for the genome-wide siRNA screen.
-{\bfseries For each well in the high-throughput screen, the period, amplitude and damping rate are calculated.
-After normalization,} distributions in robust z-scores closely resemble normal distributions.
-For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.
-\end{manuscript}
-
-\begin{reviewer}
-3) The claim that the stochastic noise can be altered independently from amplitude doesn't seem sufficiently well substantiated.
-As the authors point out, a major source of noise in single cell rhythms is the low molecular counts of clock mRNAs and proteins.
-Increasing the amplitude generally implies larger molecular counts and therefore reduced noise.
-Also, as shown in Figure 5, siRNA perturbation led to greatly increased variability in period and amplitude, which could be the main cause of the increased damping rate (rather than increased stochastic noise).
-That is, the effect of the perturbation could be to amplify the heterogeneity, rather than the individual noise levels.
-For instance, a change in a key rate parameter governing the nonlinear molecular clock mechanism might lead not only to a change in the mean period across the population, but also to a change in the spread of the period distribution.
-\end{reviewer}
-
-In general, we agree with the reviewer on this point and have updated our language accordingly (see above).
-
-In regards to the reviewer's first comment, that amplitude and noise cannot be independently altered, we note that circadian oscillations are an emergent property from a very large transcriptional network.
-As such, the species present at low quantities which dominate the noise generation are not necessarily the same as those which are tied to the bioluminescence reporter.
-A reduction in effective concentration of a repressor, for instance, might lead to higher-amplitude transcription which is more susceptible to intrinsic noise.
-In addition to the linear independence we show between amplitude and damping rate, previous authors have also shown this independence: {\itshape ``The amplitude of rhythmicity also did not correlate with cycle-to-cycle variation of locomotor ($p = 0.17$), explant ($p = 0.24$), or cellular rhythms ($p = 0.15$)''} (Herzog {\itshape et al., JBR} 2004).
-
-Additionally, we caution that population-level distributions, as a result of perturbing each gene in the siRNA library, are likely not representative of the distributions we would expect to see at the single-cell level from a single siRNA.
-
-\begin{reviewer}
-4) Using the PER2::LUC fibroblast recordings to test the ideas provides a clever test of the method.
-However, there are still some concerns about whether the method would work as proposed.
-For instance, the assumption is that the cells will have synchronized phases at the beginning of the experiment.
-For the test, the authors aligned the individual cell traces, but in an experiment the cells would presumably be synchronized using a medium change or some other synchronizing pulse.
-However, such perturbations also tend to transiently increase the amplitude of the individual cells.
-How would that effect be distinguished from the damping due to stochastic noise and phases drifting apart?
-\end{reviewer}
-
-This was actually the subject of a previous manuscript, St.\ John {\itshape et al., Biophys J}, 2014, which shows that the temporary amplitude change would largely be distinguished by a deviation from the exponential damping.
-This underscores the importance of collecting several days of data to accurately judge the damping rate.
-Additionally, for the siRNA screen, the synchronizing pulse (medium change) is consistent across all the wells.
-Since our method mainly analyses {\itshape differences} in damping rate, we suspect these effects would be minimal.
-
-\begin{reviewer}
-5) In the methods, what is meant by the (8hr,258hr) wavelet component?
-What type of wavelet transform/filter is being applied, and what do the 8hr and 258hr values refer to?
-Similarly, please clarify what the (1hr,8hr) wavelet component is.
-\end{reviewer}
-
-We used a discrete wavelet transform, the exact details of which are described in the original article (Leise et al., 2012 (18)).
-Since frequency bins are halved at each interval, the (8hr, 256hr) component represents the sum of the reconstructed 8-16, 16-32, 32-64, 64-128, 128-256 hr components.
-
-We have added additional text to the methods section detailing the wavelet transform we used
-
-\begin{manuscript}{Pages 10}
-  As was done in the original study, a discrete wavelet transform (using PyWavelets, \url{http://www.pybytes.com/pywavelets}) was performed to detrend and remove noise.
-{\bfseries A discrete wavelet transform decomposes the signal into multiple frequency bands (25).
-By only considering frequency bands close to the circadian frequency, high-frequency noise and low-frequency baseline oscillations can be removed.}
-\end{manuscript}
-
-\begin{manuscript}{Page 10}
-As in the original study, various parameters describing the average noise level of each cell were collected.
-{\bfseries
-Traces were denoised and detrended by keeping only the $(8 \text{hr}, 258 \text{hr})$ wavelet components -- resulting in rhythms which only contained oscillations with periods between 8 and 258 hours.
-}
-From these smoothed trajectories, a Hilbert transform was used estimate points at which the phase crossed $0$ to find period and amplitude coefficients of variation.
-An additional noise parameter, the standard deviation in the $(1 \text{hr}, 8 \text{hr})$ wavelet components divided by the overall rhythm amplitude, was used to quantify the high-frequency noise of the system.
-{\bfseries
-This wavelet component was chosen as it contained only the highest-frequency noise in the signal's spectrum.
-}
-\end{manuscript}
-
-\section*{Reviewer \#2}
-\begin{reviewer}
-I have enjoyed reading the paper by St.\ John and Doyle - its an innovative analysis of single cell circadian data and I think it will be of interest to a wide audience. 
-In particular the finding that dephasing occurs mostly due to noise rather than period heterogeneity is fascinating. 
-My major concern is the statement that ``Higher noise results in faster damping in population-level rhythms'' - though I agree that this follows from their data analysis, I am not so convinced that this can generally "serve as an accurate measure of cell-autonomous stochastic noise", as mentioned in the abstract and other parts of the paper.
-For example see Fig. 9 of the simulation study presented in Thomas, Philipp, Hannes Matuschek, and Ramon Grima. 
-``Intrinsic noise analyzer: a software package for the exploration of stochastic biochemical kinetics using the system size expansion.'' PloS one 7.6 (2012): e38518.
-Therein it is shown, using a model of a rudimental circadian oscillator involving strong negative feedback, that the population-level rhythm predicted using ensemble averaged stochastic simulations (of uncoupled cells) shows damped oscillations which the rate equations (deterministic and hence no noise) miss!
-That is in this case noise induces synchronous oscillations at the population level, rather than damping them!
-This is, i believe, an example which runs counter to the author's claims about noise leading to damped population level oscillations, generally.
-I would suggest the authors to discuss this possibility in the Discussion, for completeness sake, and also as a cautionary note on the use of damping as a measure of single cell noise.
-\end{reviewer}
-
-We thank the reviewer for their favorable impression of the manuscript.
-They raise the point that for some oscillators, the presence of noise is required for oscillations.
-This is very much in line with the theory of our manuscript, and is a point we raise in the introduction:
-{\itshape
-``Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, capture the noise characteristics seen in single-cell fibroblast data equally well as limit-cycle oscillators.''}
-
-We note that we do not intend {\itshape damping} to imply reduced amplitude.
-Instead, we use damping to indicate the rate at which the amplitude decays.
-Our analysis of the effect of noise on damping rate apply mainly to changes in noise and the resulting changes in exponential amplitude decay.
-We have added text to the manuscript to make our choice in vocabulary more clear:
-
-\begin{manuscript}{Page 10}
- For numerical efficiency, the period, $T$, and damping rate, $d$, parameters are fit first using a matrix pencil method (24), reviewed in (25).
-Amplitude, $A$, and phase, $\theta$, parameters are subsequently fit using a linear least-squares regression.
-{\bfseries
-Note that in this manuscript we use amplitude to denote the initial rhythm strength, and damping rate to denote the rate at which this strength decays with time.}
-\end{manuscript}
-
-\section*{Reviewer \#3}
-
-\begin{reviewer}
-These authors expand on previous work (in Biophys J 2014) where they showed that random phase diffusion in oscillators cause population level signals to dampen exponentially.
-They show that this property can be exploited to infer changes in noise levels in the circadian oscillators of single cells from population level measurements.
-This is a potentially very useful method, and I would be looking forward to try it on our own data.
-I have concerns as outlined below, but I want to stress that I am generally favorable to this work and I tried to keep the suggested revisions and additional computations requested reasonable, none of them should be too time-consuming in relation to the added value.
-\end{reviewer}
-
-We thank the reviewer for their positive critique of the study, and are excited that they would consider applying the method to their own work.
-
-\begin{reviewer}
-Main concerns:
-
-M1.
-The author's comments notwithstanding, there are some concerns regarding the possible contribution of period heterogeneity to the quantified population damping.
-Specifically, the decays in Fig 3A actually looks somewhat like the decay in theoretical Figure S6, right panel, which was computed assuming period heterogeneity.
-Of course, the authors convincingly show that population period heterogeneity is less than peak-to-peak variation in percent.
-I still request a control experiment, in which simulated time-series of observed peak-to-peak variation AND observed population period heterogeneity are analyzed with the authors' algorithm.
-The method should be shown to work as for instance noise levels are increased or decreased also for this case.
-This could be a complement to Figure 1C, where the same thing was done for oscillators with identical period.
-\end{reviewer}
-
-These concerns were also shared by the first reviewer, and we have therefore substantially changed the language of the study to reflect the fact that extrinsic noise (period heterogeneity) could play a role in the damping rates.
-For Fig. 3A, these profiles were obtained from a mathematical model of homogeneous cells.
-The deviation from the exponential window is likely due to the limit cycle not being a perfect sinusoid.
-
-We thank the reviewer for suggesting the control experiment, which we performed almost exactly as described. The results of these computations have been added as Figure S7:
-
-\begin{manuscript}{Page 16}
-  \begin{center}
-  \includegraphics[width=0.51\textwidth]{figures/pdfs/FigS7.pdf}
-  \end{center}
-  {\bfseries Figure S7: Identifiability of intrinsic and extrinsic noise factors in population-level data.} A simple model with both intrinsic and extrinsic noise (Table~S3) was used to determine the identifiability of noise sources.
-Bioluminescence profiles were generated from a population of 1000 individual oscillators.
-The stochastic volume parameter, $\Omega$, controls the intrinsic noise, with larger volumes resulting in more deterministic profiles.
-Extrinsic noise was generated by using a normal distribution of free running periods, ${\cal N}(1, \sigma^2_T)$ days.
-Points represent window functions (amplitude over time) from 10 independent replications.
-Lines represent linear regressions, except in the right panel of part A, where a quadratic regression was used.
-  ({\bfseries A}) For models with pure intrinsic (left) or pure extrinsic (right) noise sources, amplitude damping proportional to $e^{dt}$ vs.\ $e^{dt^2}$, respectively, was readily distinguishable. 
-  ({\bfseries B}) For models with both intrinsic and extrinsic noise sources, it was not possible to differentiate between increasing intrinsic noise (left) and increasing extrinsic noise (right), as both showed nearly linear relationships between time and log amplitude.
-  The reduction in consistency between points seen for very small amplitude oscillations represents a fundamental limitation of the method, as noise dominates when oscillations approach steady state.
-
-\end{manuscript}
-
-Figure S7 indicated that although it was possible to differentiate between noise sources when only one type of noise was present, determining changes to one noise source when both are present is likely not identifiable from the data.
-Ultimately, however, we don't believe that the key insight of our method is determining the explicit source of the stochastic variation.
-Since both types of noise contribute to determining the population-level dynamics, quantifying the overall stochastic noise is the main goal.
-However, based on the evidence presented in this study and in previous studies, we offer the explanation that cell-autonomous noise is likely the principal source of rhythm damping.
-
-\begin{reviewer}
-M2.
-The method is of potential use for many labs including my own.
-What I was missing, however, are practical guidelines: How many samples are needed, which sampling rates, etc.
-are required for the method to work and yield ?
-How does it scale with increasing or decreasing sampling rates?
-Importantly, how do I decide whether a possible change in damping and thus noise level is significant, can one bootstrap this?
-Should one always use the fitting procedure with matrix pencil method, Hilbert transform to align phases, etc.?
-Here, a few trials on simulated time-series may suffice to provide such information.
-\end{reviewer}
-
-This is an excellent suggestion, and we have added additional sections to the manuscript describing practical considerations for experimental results.
-More important than the sampling rate is the overall duration of the experiment, as this allows a more confident regression of the amplitude window.
-
-The following paragraph was added to the discussion to address practical considerations in experimental design and method choice:
-
-\begin{manuscript}{Page 9}
- In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
- {\bfseries
-An overview of the computational steps involved in this method are outlined in Figure~S8.
-While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
-Because the damping rate must be inferred from the time-varying amplitude, collecting bioluminescence data for longer time periods yields more accurate results and reduces the potential impact of initial transient regions.
-Additionally, a sampling rate that is high enough to confidently capture the peaks and troughs of gene expression is required -- in this study, the 2 hr sampling window of the siRNA screen proved sufficient.
-While achieving such a rate is typically not difficult for bioluminescence experiments, it may limit the method's applicability in experiments where samples need to be analyzed at each time point.
-We also note that there are many available tools for detrending and regressing time-series data.
-While we prioritized computationally efficient methods (which could be scaled to genome-wide screens), the best methods for any particular application may vary depending on the data.
-}
-\end{manuscript}
-
-Additionally, we have added a supplemental figure, Figure S8, to provide a more detailed overview of the method:
-
-\begin{manuscript}{Page 16}
-  \begin{center}
-  \includegraphics[width=0.8\textwidth]{figures/pdfs/FigS8.pdf}
-  \end{center}
-{\bfseries Figure S8: Overview of the computational method.} ({\bfseries A}) Raw bioluminescence traces are collected from cultured cells.
-Since the method is mainly dependent on amplitude over time, longer experiments yield more identifiable damping rates.
-({\bfseries B}) Raw traces are detrended and denoised.
-This can be accomplished using a variety of methods, including the discrete wavelet transform.
-({\bfseries C}) The window of the oscillations (amplitude vs.\ time) is plotted on a semi-log axis to make the exponential damping apparent.
-Window functions can be calculated using a Hilbert transform, or simply by using the absolute value of the local peaks (as was done here).
-A linear regression of the window function in semi-log space will yield both the initial amplitude (intercept) and damping rate (slope), with care being taken to avoid the over-influence of outlier points.
-Standard statistical techniques could then be used to verify that differences in slope are indeed significant.
-\end{manuscript}
-
-We elected not to include simulated time-series demonstrations of these points, mainly because of the difficulty in getting simulated time-series to match the variability seen in biological systems.
-Because of this, we believed recommended sampling rates and durations would not necessarily be reflective of biological data.
-
-In regards to using the Hilbert transform to align phases, this was done mainly to be able to analyze the dephasing kinetics from a single-cell perspective.
-Since there are relatively few sources of single-cell circadian data, this method is likely not applicable to many external studies.
-
-%
-
-% I'll add a paragraph on practical considerations for implementing the algorithm.
-Sampling rate is less important than overall recording duration.
-There are many ways to do the nonlinear regression and analysis, which likely depends on the data being analyzed.
-For looking specifically at the damping rate, the best method is likely detrending, followed by a semilog plot of window vs.\ time, from which more traditional statistical techniques on comparing two regressions (ancova, bayesian inference)
-
-\begin{reviewer}
-Minor concerns:
-
-Relating to question M1, the authors refer to their previous work, ref.
-23, where it was shown that phase drift leads to an exponentially decreasing population signal.
-Otoh, period heterogeneity would lead to a time-squared exponent; where is the latter formula derived?
-If this was done by Rougemont and Naef, they should be cited.
-\end{reviewer}
-
-We have largely re-worded this section of the text in response to the comments of other reviewers.
-Despite this, we have added a further citation to Rougement \& Naef, as they present a more detailed theoretical analysis of the windowing function:
-
-\begin{manuscript}{Page 16}
-   ({\itshape Right}) A deterministic, heterogeneous population dephases due to different free-running periods.
-In this case, the population displays ballistic phase diffusion, in which envelope of the population-averaged expression changes proportionally to $t^2$.
-{\bfseries (See Rougemont \& Naef, 2007 (22) for a more detailed discussion).}
-\end{manuscript}
-
-In general, however, the concept that Brownian particles diffuse with a variance that scales with $t$, and that ballistic particles diffuse with a variance that scales with $t^2$, is very old.
-The result presented here follows from the connection between amplitude and the variance of the phase probability.
-
-\begin{reviewer}
-On p.
-3, comments regarding reference 17: these authors rather submitted that either damped or self-sustained oscillators describe the data equally well, which actually more justifies the present method, which assumes self-sustained oscillations.
-\end{reviewer}
-
-We thank the reviewer for catching this inconsistency, have have updated the text accordingly:
-
-\begin{manuscript}{Page 3}
-  Additional studies have suggested that the basis of single-cell rhythmicity may depend on stochastic noise, as models of deterministically damped oscillators, when simulated stochastically, capture the noise characteristics seen in single-cell fibroblast data {\bfseries equally well as limit-cycle oscillators (20).}
-\end{manuscript}
-
-However, we note that our assumptions and the concept of noise-driven oscillators are not necessarily exclusive.
-In the cited work, the authors demonstrate that given a constant noise input, the damped oscillator filters the noise to amplify noise in the circadian frequencies.
-This yields an oscillator with similarly sustained amplitudes to a noisy limit-cycle.
-
-\begin{reviewer}
-The method concerns primarily detection of noise levels, in the introduction it is emphasized that an amplitude boosting may ultimately be desirable in e.g.
-a clinical setting.
-Introduction would be strengthened by also including more use cases for the detection of noise strength.
-\end{reviewer}
-
-In addition to expanding the discussion on cellular noise in response to a previous reviewer's comments, we have added some additional remarks on the importance of measuring stochastic noise:
-
-\begin{manuscript}{Page 3}
-  {\bfseries
-  The amount of noise in system is therefore linked to the ability of tissue-level clocks to maintain high-amplitude rhythms. }
-
-Despite the averaging that occurs at the population-level, cell-autonomous stochastic noise plays an important role in determining the function of the circadian oscillator.
-{\bfseries Noise in circadian rhythms has long been considered an important factor in how circadian rhythms have evolved (18).} A recent study showed that stochasticity is critical to the population-level response to a neuropeptide and forms the basis for how the SCN entrains to light-mediated cues (19).
-\end{manuscript}
-
-We also note that the importance of amplitude boosting is linked to the measurement of noise, as noise likely plays a key role in determining the rate at which rhythms damp in peripheral tissues.
-
-\begin{reviewer}
-For the discussion: authors argue that the amplitude/noise ratio seem to have evolved to something close to an optimum.
-Yet, in the paper by Liu et al.
-in Cell 129, 2007, Cry2 knockouts seemed to have even stronger rhythms?
-\end{reviewer}
-
-This is an interesting point.
-We are careful in saying that there may indeed be perturbations or knockouts which shift the clock to a region of higher stability, but that on average perturbations tend to lead to less robust rhythms.
-
-We sadly don't have access to the raw data of Liu et al.\ to look more closely at their observations.
-In the genome-wide siRNA screen, however, Cry2 knockdown was one of the several positive controls run in each plate.
-From this data, it seems that Cry2 knockdown seems to simultaneously reduce both the population-level amplitude and damping rate:
-
-\begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=0.7\textwidth]{response_figs/cry2_ts.pdf}
-  \end{center}
-  \caption{{\bfseries Bioluminescence trajectories of control and Cry2 knockdown wells.} ({\itshape Left}) Control wells which did not contain an siRNA perturbation.
-({\itshape Right}) Wells with Cry2 siRNA.}
-\end{figure}
-
-Looking at the parameter fits of these data, we see that while Cry2 siRNA does indeed seem to make the oscillators less noisy (lower damping rate), it also slightly reduces their amplitude.
-These results are mirrored by the dose-dependent Cry2 knockdown from the same study (Zhang et al., Cell 2009), which shoes a reduction of amplitude at higher Cry2 si concentrations.
-
-\begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=0.4\textwidth]{response_figs/cry2_dist.pdf}
-  \end{center}
-  \caption{{\bfseries Distribution of amplitude and damping rate fits for control and Cry2 siRNA wells.} The mean of both distributions is highlighted with a black dot.}
-\end{figure}
-
-However, since these points require comparing population-level amplitudes to single-cell level amplitudes (which may not be directly comparable), we decided to leave it out of the manuscript.
 
@@
 \begin{reviewer}
-Circadian rhythms play important roles in maintaining biological homeostasis.
-Recent research suggests that environmental conditions, like night work or an unusual feeding schedule, may have important implications for developing metabolic disease.
-Therefore it is of interest to identify therapeutic means to manipulate the circadian rhythms.
-Here, St.
-John and Doyle explore the relationship between stochastic noise and circadian oscillations, as typically observed at the cell population level and suggest a method to infer stochastic noise properties from population behavior.
-Overall, the work seems too tightly focused on a narrow scientific niche and the method for deducing noise properties is not convincing.
-The major criticisms are as follows:
-
-1.
-The relationship between stochastic noise or cell-to-cell variability and how that impacts a population response is not particularly new.
-The observation that cell to cell variability gives rise to a dampened response at the cell population level has been observed before for other signaling systems, such as NF-kappaB (e.g., Nelson et al.
-Science (2004) 306:704-708).
-It is then unclear as to what really is new here.
+In short, the authors didn't really address the concern about applying their simple model for deductive reasoning.
+They claim that as the siRNA method is well-accepted by the circadian community that their interpretation should be legit.
+Biological response is proportional to dose and target.
+It seems implicitly that the authors are assuming that the DOSE of siRNA transfected into every individual cell is exactly the same.
+The biological response is then interpreted solely as being attributed to the target.
+One look at a flow cytometry result of a siRNA experiment will tell you that the dose is not the same between cells.
+Different siRNA also don't work equally well either.
+So again the criticism that the approach is not convincing in providing the claimed insight still is unaddressed.
 \end{reviewer}
 
-We agree that the relationships between cell synchrony, stochastic noise, and damped oscillations have been previously explored in many fields, including circadian rhythms.
-In addition to a more through description of noise sources, we have included an explicit reference to the presence of these phenomena in other systems in the introduction:
-
-\begin{manuscript}{Page 3}
-  {\bfseries This type of behavior has also been seen in
-  other experimental systems, such as NF-$\kappa$B signaling or yeast glycolytic oscillations (16, 17).}
-\end{manuscript}
-
-The novel contribution of our manuscript is that we can use changes in the damping rates of population-level data to gain new insight into the relative levels of noise present under different perturbations.
-From the datasets we analyze, damping appears to be independently altered (from period, amplitude) by genetic perturbations, and therefore likely provides new insight into the underlying oscillator.
-Our main message is that the damping rate is due to cellular noise, but that the damping rate contains information about the oscillator which is currently going unused.
-
-\begin{reviewer}
-2.
-The authors use a leap of logic to justify their approach in extracting stochastic noise parameters from cell population data.
-First the authors show that by fitting a collection of models to single cell data and then averaging these mathematical models together, they can predict that the cell population response will become dampened when there is significant variability in the single cell responses (inductive reasoning?).
-Based on this demonstration, they then develop an approach that does the opposite: from population data, they try to deduce the values for the stochastic noise properties (deductive reasoning?).
-The problem seems to be that there could be other explanations for why miRNAs within a library affect a signaling network, such as miRNAs limit the efficiency of the biochemical signaling network that underpins the oscillatory behavior.
-Overall, the approach is not convincing in providing the claimed insight.
-\end{reviewer}
-
-As in all modeling work, we must make simplifying assumptions that may be drawn from incomplete experimental data.
-In developing our approach for inferring noise properties from population-level data, we have invoked two main assumptions: that individual cells are both independent and sustained oscillators.
-We are careful to list these assumptions in the introduction, as we agree that the validity of our results rests upon whether or not these assumptions hold in the systems under consideration.
-However, to the best of our knowledge, there is relatively strong experimental support to both of these assumptions in cultured circadian reporter cells.
-In the earlier parts of our paper, we are demonstrating that single-cell noise is {\itshape sufficient} to explain the observed damping behavior, and indeed quantitatively accurate predictions for population-level damping rate are obtained from a simple model not fit to the noise properties.
-
-In regards to other potential explanations for the effects of siRNA, we suspect intercellular signalling networks do not play a significant role in cultured cell oscillations (as we assume each cell is independent).
-These perturbations are thought to affect the cell-autonomous transcription-translation feedback loop, resulting in potentially altered noise properties.
-Indeed, the siRNA method has been well-accepted by the circadian community, and are widely used to study the effect of gene knockdowns on amplitude and period.
-
-\begin{reviewer}
-Minor comments:
-1. ``which'' is used incorrectly pretty much throughout the manuscript. should be ``that''
-\end{reviewer}
-
-We thank the reviewer for pointing out this error. We have corrected our usage throughout the manuscript.
-
-We would like to once again thank all our reviewers, as we feel our manuscript has benefited greatly from their suggestions and critiques.
-
-\vspace{4ex}
-\begin{flushright}
-  Sincerely,\\[2ex]
-
-  \includegraphics[width=0.3\textwidth]{/home/peter/Documents/Research/figures/frank_signature.pdf}\\[1ex]
-Francis J. Doyle III
-\end{flushright}
-
 \end{document}

### 2015-07-02 response_letter.tex
addressing point on including qualitative measures

@@
 \usepackage{tcolorbox}
-% \tcbuselibrary{breakable}
-% \tcbuselibrary{skins}
+\tcbuselibrary{breakable}
+\tcbuselibrary{skins}
 \usepackage{units}
@@
 
-% \tcbset{skin=enhanced}
+\tcbset{skin=enhanced}
 
@@
 
-We are pleased to hear that two of the reviewers recommend our work for publication.
+We are pleased to hear that two of the four reviewers recommend our work for publication in {\itshape PLoS Computational Biology}.
 We thank the reviewers for their second round of thoughtful critiques, which we feel further improves the quality of the manuscript.
@@
 \begin{reviewer}
-In the new expanded paragraph discussing the effect of cell period heterogeneity, the authors dismiss the significance of intrinsic differences in period across the population compared to cycle-to-cycle fluctuations and don’t directly quantify the relative contributions.
+In the new expanded paragraph discussing the effect of cell period heterogeneity, the authors dismiss the significance of intrinsic differences in period across the population compared to cycle-to-cycle fluctuations and don't directly quantify the relative contributions.
 However, Cohen et al (Journal Theoretical Biology 2013) quantified it for this same set of fibroblasts using a Bayesian hierarchical model and found the standard deviation of periods across the population to be 0.89h compared to the within-cell variability 1.43h.
@@
 \end{reviewer}
+
+Indeed, we agree with the reviewer that adding quantitative measures adds to the discussion of the relative importance of intrinsic vs.\ extrinsic period variability.
+We have therefore added these values from both studies.
+In addition, we have softened some of the language to avoid dismissing the effect of extrinsic variability, which likely indeed plays a non-negligible role in determining the damping rate.
+
+\begin{manuscript}{Page 6}
+  While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise {\bf may} play a more significant role.
+{\bf We} show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells {\bf (Figure S5): individual cell period lengths have an average inner quartile range (IQR) of 3.18 h, while cell-to-cell average period has an IQR of 1.55 h.
+These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population (25). 
+A similar} result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise (2).
+
+{\bf It is also} possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
+
+\end{manuscript}
 
@@
 \end{reviewer}
-
 \end{document}

### 2015-07-02 main_text.tex
second_revison
addressing point on including qualitative measures

@@
 
-\begin{figure}[tbp]
-  \begin{center}
-    % \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig1.pdf}
-  \end{center}
-  \caption{{\bfseries Single-cell bioluminescence recordings show that higher stochastic noise results in faster damping at the population-level.}
-  Data on the bioluminescence of single-cell fibroblasts was taken from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
-({\bfseries A}) Cells were sorted into two groups depending on their degree of stochastic noise.
-An example trace from each of the two groups is shown, demonstrating different levels of noise present in the dataset.
-({\bfseries B}) After artificially synchronizing each cell, we calculate averaged bioluminescence rhythms of each group (solid lines).
-A damped sinusoid fit to both groups reveals a difference in damping rate, demonstrated by fitted envelope functions ($\pm A\exp{-dt}$, dashed lines).
-({\bfseries C}) The observed absolute difference in damping rate was shown to be significant ($p = 0.0264$) using $10,000$ bootstrap trials.}
-\label{fig:fibroblast_noise}
-\end{figure}
-
 \subsection*{Clock perturbations can change single-cell stochastic noise}
@@
 This result indicates that stochastic noise can be altered by perturbations known to affect the transcription-translation feedback loop.
-
-\begin{figure}[tbp]
-  \begin{center}
-    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig2.pdf}}
-  \end{center}
-  \caption{{\bfseries Small molecule modulator KL001 increases damping rate in a dose-dependent fashion.}
-  Experimental data on the dose-dependent effects of small molecules KL001 and longdaysin on cultured circadian reporter cells was taken from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.
-({\bfseries A}) Detrended bioluminescence signals from the two reporter systems and two small molecules are shown normalized by the fitted amplitude, period, and phase.
-The normalized bioluminescence highlights the dose-dependent change in damping rate seen with the KL001 application (top), but not with longdaysin (bottom).
-({\bfseries B}) Quantification of the dose-dependent change in damping rate caused by small molecule modulators.
-While both molecules lead to a dose-dependent increase in period, only KL001 shows a reliable change in damping rate.}
-\label{fig:dose_dependence}
-\end{figure}
 
@@
 
-\begin{figure}[tbp]
-  \begin{center}
-    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig3.pdf}}
-  \end{center}
-  \caption{{\bfseries Mathematical model accurately predicts dose-dependent changes in damping rate.}
-({\bfseries A}) Example single-cell trajectories (top) and population-averaged trajectories (bottom, mean of $1,000$ cells) of cells under various treatments.
-Cells with the nominal parameter set (left, black) closely match the experimental damping rate for unperturbed cells.
-Cells with simulated KL001 action (red, center) are noisier at the single-cell level, and show faster damping at the population-level.
-Cells with simulated longdaysin action show slightly more accurate single-cell rhythms, with a corresponding reduction in the population-level damping rate.
-({\bfseries B}) The model accurately predicts the general trend of period vs.\ damping rate for both KL001 and longdaysin perturbations.
-Experimental data points represent the mean of two replications at each concentration.
-Computational data points represent the mean of ten independent population simulations.}
-\label{fig:simulation}
-\end{figure}
-
 \subsection*{Genome-wide effects of siRNA knockdown on single-cell stochastic noise}
@@
 
-\begin{figure}[tbp]
-  \begin{center}
-    % \includegraphics[]{figures/pdfs/Fig4.pdf}
-  \end{center}
-  \caption{{\bfseries Fit quality vs.\ damping rate for the genome-wide siRNA screen.}
-({\bfseries A}) A plot of the $111,743$ individual fits shows that the majority of fits have a high $R^2$ value and positive damping rate. Only fits with $R^2 > 0.8$ were kept for further analysis.
-({\bfseries B}) Examples chosen randomly from each of the four quadrant regions in (A). Sinusoidal parameters for fits 1-2 can be more confidently inferred than those for fits 3-4.}
-\label{fig:fit_quality}
-\end{figure}
-
 We next checked how parameters varied on a plate-to-plate and well-to-well basis.
@@
 
-\begin{figure}[tbp]
-  \begin{center}
-    % \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig5.pdf}
-  \end{center}
-  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} For each well in the high-throughput screen, the period, amplitude and damping rate are calculated.
-After normalization, distributions in robust z-scores closely resemble normal distributions.
-For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.}
-\label{fig:fit_distributions}
-\end{figure}
-
-\begin{table}
-  \begin{center}
-    \begin{tabular}{lrrrrrr}\toprule
-      {} & \multicolumn{2}{c}{$T$} & \multicolumn{2}{c}{$\ln A$} & \multicolumn{2}{c}{$d$} \\
-      {}         & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}               & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           \\\midrule
-      Mean     & $-0.234$ & $0.187$ & $0.443$  & $-0.343$ & $0.043$  & $0.090$    \\
-      Variance & $ 0.599$ & $3.313$ & $0.605$  & $ 3.074$ & $0.771$  & $2.850$    \\
-      Skewness & $ 0.153$ & $0.367$ & $-1.823$ & $-0.580$ & $-0.107$ & $0.371$    \\
-      Kurtosis & $ 3.772$ & $0.591$ & $8.329$  & $ 0.476$ & $2.423$  & $0.373$    \\
-      \bottomrule
-    \end{tabular}
-  \end{center}
-  \caption{{\bfseries Moments of the fitted parameter distributions (Figure~\ref{fig:fit_distributions}) after normalization and outlier removal.} Parameters were normalized by subtracting the median and dividing by the median absolute deviation on a plate-by-plate basis.
-Control distributions had less variance, were less skewed, and were more peaked then their perturbed counterparts.}
-  \label{tab:fit_distributions}
-\end{table}
-
 \subsubsection*{Damping rate is independent of other sinusoidal parameters}
@@
 These results reinforce the notion that the population-level damping rate describes an independent feature of the underlying oscillator.
-
-\begin{table}
-  \begin{center}
-    \begin{tabular}{rrrrr}
-      \toprule
-      {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
-      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\
-      $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
-      $T$      & $-0.142$ & $-0.022$  & $1.000 $ & $-0.113$\\
-      $\theta$ & $-0.269$ & $-0.112$  & $-0.113$ & $1.000 $\\
-      \bottomrule
-    \end{tabular}
-  \end{center}
-  \caption{{\bfseries Correlation among normalized parameters of the high-throughput siRNA screen.} Pearson correlation coefficients are relatively low between fitted parameters, indicating that changes to damping rate (and thereby stochastic noise) are not explained by changes to period, amplitude, or phase.}
-  \label{tab:corr}
-\end{table}
 
@@
 
-\begin{figure}[tbp]
-  \begin{center}
-    % \includegraphics[width=0.65\textwidth]{figures/pdfs/Fig6.pdf}
-  \end{center}
-  \caption{{\bfseries Effects of siRNA knockdowns on amplitude and damping rate.}
-  Clock robustness is a function of both amplitude and damping rate.
-  Distributions in amplitude and damping rate for control wells ({\bfseries A}) or perturbed wells ({\bfseries B}) indicate that perturbations tend to shift the clock towards regions of higher damping rate or lower amplitude.
-Green dots in each figure indicates the mean of the control population.
-({\bfseries C}) Averaged effect of siRNA knockdown after grouping the perturbed population by Gene ID.
-Only those genes that were significantly different from the control distribution are shown (Hotelling's $T^2$ test, $\alpha = 0.01$).
-({\bfseries D}) Radial histogram of the significant gene perturbations shown in (C).
-The area of each slice is proportional to the frequency of perturbations away from the mean in that direction.
-Very few gene knockdowns result in both higher amplitude and lower damping rate (red slices, lower right quadrant).
-}
-\label{fig:outlier_dist}
-\end{figure}
-
 \subsection*{Effect of cell heterogeneity}
@@
 Indeed, prior to the availability of single-cell data, studies explored the possibility that differences in mean periods served as the mechanism behind population-level damping \cite{Izumo2003}.
-While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise likely plays a more significant role: we show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}).
-This result has also been observed in dispersed SCN neurons, indicating cell heterogeneity is less severe than cell-autonomous noise \cite{Herzog2004}.
+While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise may play a more significant role.
+We show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}): individual cell period lengths have an average inner quartile range (IQR) of 3.18 h, while cell-to-cell average period has an IQR of 1.55 h.
+These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population \cite{Cohen2012}.
+A similar result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise \cite{Herzog2004}.
 
-However, it is possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
+It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
 Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics.
@@
 We thank Tsuyoshi Hirota, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
-This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.
+% This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.
 
@@
 \bibliography{condensed_library.bib}
+
+\clearpage
+
+\begin{figure}[h!]
+  \begin{center}
+    % \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig1.pdf}
+  \end{center}
+  \caption{{\bfseries Single-cell bioluminescence recordings show that higher stochastic noise results in faster damping at the population-level.}
+  Data on the bioluminescence of single-cell fibroblasts was taken from Leise {\itshape et al.,} 2012 \cite{Leise2012}.
+({\bfseries A}) Cells were sorted into two groups depending on their degree of stochastic noise.
+An example trace from each of the two groups is shown, demonstrating different levels of noise present in the dataset.
+({\bfseries B}) After artificially synchronizing each cell, we calculate averaged bioluminescence rhythms of each group (solid lines).
+A damped sinusoid fit to both groups reveals a difference in damping rate, demonstrated by fitted envelope functions ($\pm A\exp{-dt}$, dashed lines).
+({\bfseries C}) The observed absolute difference in damping rate was shown to be significant ($p = 0.0264$) using $10,000$ bootstrap trials.}
+\label{fig:fibroblast_noise}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig2.pdf}}
+  \end{center}
+  \caption{{\bfseries Small molecule modulator KL001 increases damping rate in a dose-dependent fashion.}
+  Experimental data on the dose-dependent effects of small molecules KL001 and longdaysin on cultured circadian reporter cells was taken from Hirota {\itshape et al.,} 2012 \cite{Hirota2012}.
+({\bfseries A}) Detrended bioluminescence signals from the two reporter systems and two small molecules are shown normalized by the fitted amplitude, period, and phase.
+The normalized bioluminescence highlights the dose-dependent change in damping rate seen with the KL001 application (top), but not with longdaysin (bottom).
+({\bfseries B}) Quantification of the dose-dependent change in damping rate caused by small molecule modulators.
+While both molecules lead to a dose-dependent increase in period, only KL001 shows a reliable change in damping rate.}
+\label{fig:dose_dependence}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    % \makebox[\textwidth][c]{\includegraphics[]{figures/pdfs/Fig3.pdf}}
+  \end{center}
+  \caption{{\bfseries Mathematical model accurately predicts dose-dependent changes in damping rate.}
+({\bfseries A}) Example single-cell trajectories (top) and population-averaged trajectories (bottom, mean of $1,000$ cells) of cells under various treatments.
+Cells with the nominal parameter set (left, black) closely match the experimental damping rate for unperturbed cells.
+Cells with simulated KL001 action (red, center) are noisier at the single-cell level, and show faster damping at the population-level.
+Cells with simulated longdaysin action show slightly more accurate single-cell rhythms, with a corresponding reduction in the population-level damping rate.
+({\bfseries B}) The model accurately predicts the general trend of period vs.\ damping rate for both KL001 and longdaysin perturbations.
+Experimental data points represent the mean of two replications at each concentration.
+Computational data points represent the mean of ten independent population simulations.}
+\label{fig:simulation}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    % \includegraphics[]{figures/pdfs/Fig4.pdf}
+  \end{center}
+  \caption{{\bfseries Fit quality vs.\ damping rate for the genome-wide siRNA screen.}
+({\bfseries A}) A plot of the $111,743$ individual fits shows that the majority of fits have a high $R^2$ value and positive damping rate. Only fits with $R^2 > 0.8$ were kept for further analysis.
+({\bfseries B}) Examples chosen randomly from each of the four quadrant regions in (A). Sinusoidal parameters for fits 1-2 can be more confidently inferred than those for fits 3-4.}
+\label{fig:fit_quality}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    % \includegraphics[width=0.8\textwidth]{figures/pdfs/Fig5.pdf}
+  \end{center}
+  \caption{{\bfseries Distributions in fitted parameters for the genome-wide siRNA screen.} For each well in the high-throughput screen, the period, amplitude and damping rate are calculated.
+After normalization, distributions in robust z-scores closely resemble normal distributions.
+For all parameters, the region of highest density is consistent between the control and perturbed populations, indicating many perturbations do not appreciably change clock dynamics.}
+\label{fig:fit_distributions}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    % \includegraphics[width=0.65\textwidth]{figures/pdfs/Fig6.pdf}
+  \end{center}
+  \caption{{\bfseries Effects of siRNA knockdowns on amplitude and damping rate.}
+  Clock robustness is a function of both amplitude and damping rate.
+  Distributions in amplitude and damping rate for control wells ({\bfseries A}) or perturbed wells ({\bfseries B}) indicate that perturbations tend to shift the clock towards regions of higher damping rate or lower amplitude.
+Green dots in each figure indicates the mean of the control population.
+({\bfseries C}) Averaged effect of siRNA knockdown after grouping the perturbed population by Gene ID.
+Only those genes that were significantly different from the control distribution are shown (Hotelling's $T^2$ test, $\alpha = 0.01$).
+({\bfseries D}) Radial histogram of the significant gene perturbations shown in (C).
+The area of each slice is proportional to the frequency of perturbations away from the mean in that direction.
+Very few gene knockdowns result in both higher amplitude and lower damping rate (red slices, lower right quadrant).
+}
+\label{fig:outlier_dist}
+\end{figure}
+
+\clearpage
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lrrrrrr}\toprule
+      {} & \multicolumn{2}{c}{$T$} & \multicolumn{2}{c}{$\ln A$} & \multicolumn{2}{c}{$d$} \\
+      {}         & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}               & \multicolumn{1}{c}{C}         & \multicolumn{1}{c}{P}           \\\midrule
+      Mean     & $-0.234$ & $0.187$ & $0.443$  & $-0.343$ & $0.043$  & $0.090$    \\
+      Variance & $ 0.599$ & $3.313$ & $0.605$  & $ 3.074$ & $0.771$  & $2.850$    \\
+      Skewness & $ 0.153$ & $0.367$ & $-1.823$ & $-0.580$ & $-0.107$ & $0.371$    \\
+      Kurtosis & $ 3.772$ & $0.591$ & $8.329$  & $ 0.476$ & $2.423$  & $0.373$    \\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+  \caption{{\bfseries Moments of the fitted parameter distributions (Figure~\ref{fig:fit_distributions}) after normalization and outlier removal.} Parameters were normalized by subtracting the median and dividing by the median absolute deviation on a plate-by-plate basis.
+Control distributions had less variance, were less skewed, and were more peaked then their perturbed counterparts.}
+  \label{tab:fit_distributions}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{rrrrr}
+      \toprule
+      {}       & $d$    & $\ln A$ & $T$    & $\theta$ \\\midrule
+      $d$      & $1.000 $ & $0.285 $  & $-0.142$ & $-0.269$\\
+      $\ln A$  & $0.285 $ & $1.000 $  & $-0.022$ & $-0.112$\\
+      $T$      & $-0.142$ & $-0.022$  & $1.000 $ & $-0.113$\\
+      $\theta$ & $-0.269$ & $-0.112$  & $-0.113$ & $1.000 $\\
+      \bottomrule
+    \end{tabular}
+  \end{center}
+  \caption{{\bfseries Correlation among normalized parameters of the high-throughput siRNA screen.} Pearson correlation coefficients are relatively low between fitted parameters, indicating that changes to damping rate (and thereby stochastic noise) are not explained by changes to period, amplitude, or phase.}
+  \label{tab:corr}
+\end{table}
 

### 2015-07-02 main_text.tex
addressing point on including qualitative measures

@@
 
-Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of cells.
+Bioluminescence rhythms at the cell culture or tissue-level are the result of the collective behavior of thousands of individual cells.
 Transcription at the single-cell level is strongly affected by {\itshape intrinsic} cellular noise, caused by the low molecular counts of the mRNA and protein species involved. 
@@
 Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude.
-For circadian systems, intrinsic noise has been shown to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells \cite{Herzog2004}.
-Both sources of noise have an effect on population-level rhythms: in cell cultures that lack cell-to-cell coupling, it has been shown that  stochastic noise is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.
+The effects of noise in biological systems has been well-studied, and relative amounts of intrinsic and extrinsic noise can be identified through carefully designed single-cell experiments \cite{Elowitz2002}.
+For circadian systems, intrinsic noise has been suspected to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells \cite{Herzog2004}.
+However, both sources of noise have an effect on population-level rhythms: in cell cultures that lack cell-to-cell coupling, it has been shown that  stochastic noise is manifested in damped oscillations at the population-level as individual oscillators gradually drift out of phase \cite{Nagoshi2004, Welsh2004}.
 This type of behavior has also been seen in other experimental systems, such as NF-$\kappa$B signaling or yeast glycolytic oscillations \cite{Nelson2004, Aon1992}.

### 2015-07-04 response_letter.tex
adding discussion of Elowitz et al

@@
 
+We thank the reviewer for recommending this important study, and agree that further discussion of existing work on biological noise is useful in the introduction.
+We have therefore added an additional point to the paragraph on intrinsic and extrinsic noise sources:
+
+\begin{manuscript}{Page 3}
+Extrinsic noise results from heterogeneity between cells, such as differences in size or physical environment, leading to differences on a cell-to-cell basis in expected period and amplitude. 
+{\bfseries The effects of noise in biological systems has been well-studied, and relative amounts of intrinsic and extrinsic noise can be identified through carefully designed single-cell experiments (15).
+}
+For circadian systems, intrinsic noise has been {\bfseries suspected} to play a larger role: a single cell's variability in period from cycle-to-cycle is larger than the variability in mean period length between cells (2).
+\end{manuscript}
+
+For the discussion on page 6 where we describe what can be inferred from experimental data, we had restricted our claims to population-level recordings - whereas the Elowitz study used data from single-cell imaging to differentiate between noise sources.
+We nevertheless acknowledge the potential for misunderstanding and have added further text describing experimental techniques which might differentiate between noise sources (with a further citation to the work done by Elowitz, {\itshape et al}):
+
+\begin{manuscript}{Pages 6-7}
+It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise. 
+Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics. 
+While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure S6), these differences are likely not identifiable from typical data (Figure S7).
+{\bfseries This distinction would be possible with single-cell level data, as has been done in other experimental systems (15). 
+However, such an experiment would likely not be amenable to high-throughput methods.}
+Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
+\end{manuscript}
+
 \begin{reviewer}
@@
 \end{reviewer}
+
+In Bratsun {\itshape et al, PNAS} 2005, the authors demonstrate that variability in gene transcription can occur both through intrinsic noise as well as through fast, deterministic oscillations. While
 
@@
 {\bf We} show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells {\bf (Figure S5): individual cell period lengths have an average inner quartile range (IQR) of 3.18 h, while cell-to-cell average period has an IQR of 1.55 h.
-These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population (25). 
+These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population (27). 
 A similar} result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise (2).

### 2015-07-05 response_letter.tex
finishing reviewer updates

@@
 
-In Bratsun {\itshape et al, PNAS} 2005, the authors demonstrate that variability in gene transcription can occur both through intrinsic noise as well as through fast, deterministic oscillations. While
+In Bratsun {\itshape et al, PNAS} 2005, the authors demonstrate that variability in the amplitude of gene transcription can occur both through intrinsic noise as well as through fast, deterministic oscillations.
+While it might be possible that fast oscillations in gene transcription (in addition to the slower, 24-hour circadian rhythms) underly most of the variation seen in ultradian frequencies, it would then be unlikely that we would observe the random drift in phases.
+Since our study deals mainly with the phase precision rather than the variability in transcription, we have decided to omit this reference to avoid confusing the reader.
 
@@
 \end{reviewer}
+
+We agree with the reviewer that the effect of an siRNA on each individual cell will indeed be varied.
+This effect would be manifested in changing the extrinsic noise of the system, as cell heterogeneity would be amplified by the distribution of siRNA effect.
+While we previously mentioned variability in drug uptake as a possible source of cell heterogeneity, we have expanded this section to make the reference to siRNA more explicit:
+
+\begin{manuscript}{Pages 6-7}
+It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system’s extrinsic noise. 
+{\bfseries Such a change could be caused by an unequal uptake of siRNA or drug on a cell-to-cell basis, as has been demonstrated by a distribution of single-cell knockdown efficiency through flow cytometry (27).
+This effect would increase cell heterogeneity and lead to faster dephasing kinetics.}
+\end{manuscript}
+
+While it is also true that different siRNA constructs will have different knockdown effectivenesses, this will be a systematic effect that will not effect either the intrinsic or extrinsic noise.
+In developing our approach, we do not assume that each siRNA will be equally effective - just as the knockdown of different genes will have varied effects on circadian rhythms.
+While we do group siRNA perturbations by target gene, this test mainly serves to guard against off-target activities.
+We have added a relevant explanation to the methods section clarifying this point:
+
+\begin{manuscript}{Page 10}
+A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
+While different siRNA constructs will have different knockdown efficiencies, grouping based on gene target helps to eliminate the effect of off-target activities. 
+\end{manuscript}
+
+Once more, we would like to thank each of our reviewers for their time and attention to detail.
+We believe their follow-up critiques have further improved the rigor of this work, and have included our appreciation for their efforts in the acknowledgements.
+
+\vspace{4ex}
+\begin{flushright}
+  Sincerely,\\[2ex]
+
+  \includegraphics[width=0.3\textwidth]{/home/peter/Documents/Research/figures/frank_signature.pdf}\\[1ex]
+Francis J. Doyle III
+\end{flushright}
+
 \end{document}
+
+\end{document}

### 2015-07-05 main_text.tex
adding discussion of Elowitz et al
finishing reviewer updates

@@
 While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise may play a more significant role.
-We show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}): individual cell period lengths have an average inner quartile range (IQR) of 3.18 h, while cell-to-cell average period has an IQR of 1.55 h.
-These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population \cite{Cohen2012}.
+We show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells (Figure~\ref{fig:cell2cell-variability}): individual cell period lengths have an average inner quartile range (IQR) of 3.18 hrs, while cell-to-cell average period has an IQR of 1.55 hrs.
+These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 hrs, and 0.89 hrs across the population \cite{Cohen2012}.
 A similar result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise \cite{Herzog2004}.
@@
 It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
-Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics.
-While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure~\ref{fig:dephasing-rates}), these differences are likely not identifiable from typical data (Figure~\ref{fig:computational_dephasing}).
+Such a change could be caused by an unequal uptake of siRNA or drug on a cell-to-cell basis, as has been demonstrated by a distribution of single-cell knockdown efficiency through flow cytometry \cite{Mousses2003}.
+This effect would increase cell heterogeneity and lead to faster dephasing kinetics.
+While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure~\ref{fig:dephasing-rates}), these differences are likely not identifiable from typical population-level data (Figure~\ref{fig:computational_dephasing}).
+This distinction would likely be possible with single-cell level data, as has been done in other experimental systems \cite{Elowitz2002}.
+However, such an experiment would likely not be amenable to high-throughput methods.
 Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors.
@@
 
-In this study, we have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
-An overview of the computational steps involved in this method are outlined in Figure~\ref{fig:comp_overview}.
+% Overview of experiment - make it clear the insights from each portion. Use as changes for last reviewer.
+
+In this study we have shown that the damping rate of population-level circadian oscillations can be changed by genetic or pharmacological perturbations.
+As populations-level rhythms are determined by the coherence of many individual cells, desynchronization due to stochastic noise is a likely explanation for population-level damping.
+Using single-cell data, we showed that population-level damping rate is proportional to single-cell noise.
+Furthermore, we used a computational model to predict the changes in damping rate from two small molecules, demonstrating that changes to intrinsic stochastic noise are sufficient to explain the observed damping rate changes.
+
+We have described a method by which changes to stochastic noise can be estimated from population-level circadian bioluminescence recordings.
+An overview of the computational steps involved in our method are outlined in Figure~\ref{fig:comp_overview}.
 While the method can be applied to existing experimental data, there are also practical considerations for the design of future experiments.
@@
 A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
+While different siRNA constructs will have different knockdown efficiencies, grouping based on gene target helps to eliminate the effect of off-target activities. 
 A robust covariance estimator was used to find the location and covariance of the control distribution (using scikit-learn).
@@
 We thank Tsuyoshi Hirota, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
+We also thank the efforts of our four anonymous reviewers, who's suggestions have improved the quality of the manuscript.
 % This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.

### 2015-07-05 response_letter.tex
sending changes to frank

@@
 
-\begin{manuscript}{Pages 6-7}
-It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise. 
-Such a change, caused perhaps by an unequal uptake of drug across cells, would result increased cell heterogeneity, and different dephasing kinetics. 
-While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure S6), these differences are likely not identifiable from typical data (Figure S7).
-{\bfseries This distinction would be possible with single-cell level data, as has been done in other experimental systems (15). 
-However, such an experiment would likely not be amenable to high-throughput methods.}
+\begin{manuscript}{Page 7}
+While differentiating between intrinsic and extrinsic noise sources from population-level data is possible in theory (Figure S6), these differences are likely not identifiable from typical {\bfseries population-level} data (Figure S7).
+{\bfseries This distinction would likely be possible with single-cell level data, as has been done in other experimental systems (15). 
+However, such an experiment would likely not be amenable to high- throughput methods.}
 Differences in damping rates are therefore best viewed as representative of changes to overall stochastic noise from both intrinsic and extrinsic factors. 
+Since both types of noise are important to determining the overall function of population-level rhythms, damping rates are still a valuable method of quantifying stochastic noise.
 \end{manuscript}
@@
   While it is true that the dephasing of a group of oscillators can be caused by both variance in the mean period as well as cycle-to-cycle variability, intrinsic stochastic noise {\bf may} play a more significant role.
-{\bf We} show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells {\bf (Figure S5): individual cell period lengths have an average inner quartile range (IQR) of 3.18 h, while cell-to-cell average period has an IQR of 1.55 h.
+{\bf We} show that there is greater variance in period on a cycle-to-cycle basis than on a cell-to-cell basis in cultured fibroblast cells {\bf (Figure S5): individual cell period lengths have an average inner quartile range (IQR) of 3.18 hrs, while cell-to-cell average period has an IQR of 1.55 h.
 These results are independently confirmed by a previous study using Bayesian modeling, which found a standard deviation of periods within cells of 1.43 h, and 0.89 h across the population (27). 
-A similar} result has also been observed in dispersed SCN neurons, suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are less severe than cell-autonomous noise (2).
+A similar} result has also been observed in dispersed SCN neurons, {\bfseries suggesting that while both intrinsic and extrinsic period heterogeneity likely contribute to the dephasing kinetics, cell-to-cell differences are} less severe than cell-autonomous noise (2).
 
 {\bf It is also} possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise.
-
 \end{manuscript}
@@
 
-We agree with the reviewer that the effect of an siRNA on each individual cell will indeed be varied.
+We regret that our previous revision did not address the reviewer's concern about the reasoning generated from each of our {\itshape in silico} experiments.
+We have therefore added an expanded section of the Discussion in which we clearly lay out the insights we claim from each portion of the study:
+
+\begin{manuscript}{Page 7}
+  {\bfseries In this study we have shown that the damping rate of population-level circadian oscillations can be changed by genetic or pharmacological perturbations.
+As populations-level rhythms are determined by the coherence of many individual cells, desynchronization due to stochastic noise is a likely explanation for population-level damping.
+Using single-cell data, we showed that population-level damping rate is proportional to single-cell noise.
+Furthermore, we used a computational model to predict the changes in damping rate from two small molecules, demonstrating that changes to intrinsic stochastic noise are sufficient to explain the observed damping rate changes.}
+\end{manuscript}
+
+In regards to the reviewer's specific comments on siRNA, we agree that the effect of an siRNA on each individual cell will indeed be varied.
 This effect would be manifested in changing the extrinsic noise of the system, as cell heterogeneity would be amplified by the distribution of siRNA effect.
@@
 {\bfseries Such a change could be caused by an unequal uptake of siRNA or drug on a cell-to-cell basis, as has been demonstrated by a distribution of single-cell knockdown efficiency through flow cytometry (27).
-This effect would increase cell heterogeneity and lead to faster dephasing kinetics.}
+This effect would increase cell heterogeneity and lead to faster} dephasing kinetics.
 \end{manuscript}
@@
 A Hotelling's $T^2$ test was used to determine whether the means of each gene knockdown was significantly different from the control population.
-While different siRNA constructs will have different knockdown efficiencies, grouping based on gene target helps to eliminate the effect of off-target activities. 
+{\bfseries While different siRNA constructs will have different knockdown efficiencies, grouping based on gene target helps to eliminate the effect of off-target activities.}
 \end{manuscript}
+
+\vspace{3ex}
 

### 2015-07-14 response_letter.tex
Frank's changes, uploading second revisions

@@
 \begin{manuscript}{Pages 6-7}
-It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system’s extrinsic noise. 
+It is also possible that damping rate changes due to siRNA or small molecule perturbation could be manifested through altering the system's extrinsic noise. 
 {\bfseries Such a change could be caused by an unequal uptake of siRNA or drug on a cell-to-cell basis, as has been demonstrated by a distribution of single-cell knockdown efficiency through flow cytometry (27).

### 2015-07-24 main_text.tex
Frank's changes, uploading second revisions
fixing typo

@@
 \section*{Acknowledgments}
-We thank Tsuyoshi Hirota, David Welsh, and John Abel for the helpful discussions, as well as the each of the authors of the original experiments for making their raw data available online.
-We also thank the efforts of our four anonymous reviewers, who's suggestions have improved the quality of the manuscript.
+We thank Tsuyoshi Hirota, David Welsh, and John Abel for the helpful discussions, as well as each of the authors of the original experiments for making their raw data available online.
+We also thank the efforts of our four anonymous reviewers, whose suggestions have improved the quality of the manuscript.
 % This work was supported by the National Institutes of Health/National Institute of General Medical Sciences under award number 1R01GM096873-01 and by the Institute for Collaborative Biotechnologies through grant W911NF-09-0001 from the U.S.\ Army Research Office.
